### 2015-04-24 manuscript.Rmd
Added markdown manuscript

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [CO$_2$] but tracks water availability
author: "Remko Duursma"
date: "April 24, 2015"
output:
  html_document:
    number_sections: yes
  pdf_document: default
  word_document: default
csl: global-change-biology.csl
bibliography: references.bib
---

```{r echo=FALSE, message=FALSE}
library(knitr)
load("cache/lai_workspace.RData")
source("load.R")
```

# Abstract {.unnumbered}

Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [CO$_2$] and water availability of unmanaged forests have been poorly quantified. We studied canopy leaf area dynamics for 27 months in a native evergreen Eucalyptus woodland exposed to Free Air CO$_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [CO$_2$]. We estimated *L* from diffuse canopy transmittance and canopy photos, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [CO$_2$] in this timeframe. We found that *L* varied between 1.0 and 2.0, but did not show a clear phenological pattern. Instead, increased *L* following leaf flushing occurred a few weeks after heavy rainfall, and *L* decreased steadily during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.

*Keywords*: leaf area index, FACE, phenology, drought, litter production

# Introduction

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, (Reich, 2012), and a key variable in land surface models (Kala et al., 2013; Walker et al., 2014). *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments (Herbert and Fownes, 1995; Hubbard et al., 2004; White et al., 2010) and observational studies across rainfall gradients (Ellis and Hatton, 2008; Iio et al., 2013) [@Jin2013]. Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO$_2$] (e*C*$_a$) due to increased net carbon uptake (De Kauwe et al., 2014; Hatton et al., 1992; Kergoat et al., 2002; Zhu et al., 2012) Also expected because increased quantum yield of photosynthesis in e*C*$_a$ would allow leaf retention into deeper shade (Norby et al., 1999), but little empirical evidence is available on the response of *L* to e*C*$_a$ in mature forests (Norby and Zak 2011). Quantifying the magnitude and direction of the e*C*$_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C$_a$ may provide an important feedback to climate change, and modify direct physiological responses to C$_a$ (Tor-ngern et al., 2015)(Murray et al., 2012).

Direct effects of e*C*$_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012). Although numerous studies have been conducted on the response of physiology and plant growth to e*C*$_a$, very few of these inform the response of steady-state *L* to e*C*$_a$, because it requires a sufficiently mature forest to be exposed to free air CO$_2$ enrichment (FACE) (Norby and Zak, 2011). Increased carbon availability through enhanced photosynthesis under e*C*$_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to e*C*$_a$ *L* (Norby and Zak, 2011), although positive responses to e*C*$_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to e*C*$_a$, leading to the prediction that stands with low *L* will show a larger response to e*C*$_a$. Because diminished returns in increasing LAI in response to e*C*$_a$ when LAI is larger (Norby et al., 2005)

In ecosystems where water availability limits productivity, we could expect e*C*$_a$ to further affect *L* (Donohue et al., 2009), since e*C*$_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.
In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C$_a$. Based on the same water-savings principle, we predict that *L* should also increase under e*C*$_a$ in water-limited forests.

Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g. (Richardson et al., 2012; Savoy and Mackay, 2015) whereas *L* responses to fluctuations in water availability are qualitatively well-known (more cites), but poorly quantified. The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability (Pook, 1985; Specht and Specht, 1989; Whitehead and Beadle, 2004)(Pook et al., 1997)(Limousin et al., 2009).The strong relationship between *L* and rainfall observed across Australia (Ellis and Hatton, 2008; Specht and Specht, 1989), and other continents (Donohue et al., 2013; Grier and Running, 1977; [@Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales (Pook, 1985; Pook et al., 1997), driven by leaf turnover (Misson et al., 2011) and leaf flushing and expansion (Muller et al., 2011), which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology (Misson et al., 2011).

We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to e*C*$_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that e*C*$_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO$_2$. In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to e*C*$_a$ (Norby and Zak, 2011). Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

# Materials and Methods

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial ﬂoodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l. elevation (33°37'S, 150°44'E). It is located within a remnant patch of native Cumberland Plain woodland (Tozer, 2003). The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca. 70 years. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 °C and mean annual precipitation of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).

The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated (Crous et al., 2015). Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees ha-1. The vegetation consists of an open woodland (basal area before treatment, BA = 27.6 ± 2.7 m$^2$ ha‑1, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by Microlaena stipoides (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted CO$_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009). The vegetation within three randomly selected rings was exposed to a CO$_2$ mole fraction (C$_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO$_2$ treatment rings in infrastructure and instrumentation but received ambient C$_a$, and are used as control plots. The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol‑1 above ambient C$_a$ in full operation mode. However, C$_a$ was increased gradually for six months at a rate of ca. 30 μmol CO$_2$ mol-1 month-1 until full operation mode was reached (Figure 1a).

Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP ® sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers. Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50 depth depending on the ring).

## Terminology
In the following, we use the term ‘leaf’ area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e. woody stems and infrastructure. This avoids the need to switch between different terminologies when comparing temporal changes in *L* to litter production, and absolute values between different methods. The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 m$^2$ were installed at 0.5m above the forest floor. Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14). We did not find a significant effect of C$_a$ treatment on SLA (Ellsworth, unpublished data).

## Estimation of leaf area index from digital photography
Leaf area index (*L*) was estimated from zenith canopy photographs following (Macfarlane et al., 2007). From October 2012 to October 2013, we took ca. monthly canopy photographs in each FACE ring. Photographs were taken in the morning on cloudy days. We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring. Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded. The camera was aligned with the north-south axis and set at 1.25 m height.  Camera lens was pointed directly upwards and levelled using a three axis bubble level. We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan). To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens. Following recommendations by Jonckheere et al. (2005), we used the automated Ridler algorithm to threshold the canopy photos. We then calculated canopy transmittance (τ_p) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold. *L* from canopy photos (Lp) was calculated with a simple Beer-Lambert law (Eq. 1).

"*L*" _p=-loga(τ_p)/k									(1)

where τ_p canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following (Macfarlane et al., 2007) recommendation for eucalypt forests.

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al. 2003. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (Fdiff) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where Fdiff > 0.98. We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

τ_d=Q ̅_below/Q_above								(2)

where τ_d is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}$ s-1), and Q ̅_below is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of τ_d, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of τ_d).

We then estimated leaf area index (*L*) from τ_d (*L*τ), following (Campbell and Norman, 2000). We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

τ_d=2∫_0^(π/2)▒〖e^(-LAI∙G(θ)/cosa(θ))  sina(θ)  cosa(θ)dθ〗						(3)

where θ is the solar zenith angle, G(θ) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000). Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 (Stenberg, 2006). We solved numerically for *L* in Eq. 3, using a smooth integration over θ, yielding *L*τ.

To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from τ_d (*L*τ) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus τ_d increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate τ_d  were available during this period. In theory, cumulative *L* leaf litter production (PL, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated equation (4),

c〖(LAI〗_τ-〖LAI〗_(t=0))=∫_(t=0)^t▒〖P_L dt〗								(4)

where c is a calibration parameter, Lt=0 is *L*τ at the start of the drought period. We used linear regression without intercept to estimate c in Eq. 4 as 0.68 (SE = 0.035). It should be stressed that even after calibration with litter production, *L*τ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 4.

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014). For all time-series, we used a base of 18 when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of C$_a$ treatment on *L*τ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.

To assess effects of e*C*$_a$ on *L*τ, we accounted for pre-treatment differences in canopy structure, as follows. The average *L*τ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C$_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual *L*τ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

# Results

Canopy transmittance (τ_d) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

An outbreak of psyllids (Cardiaspina sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in Eucalyptus moluccana, see (Hall et al., 2015)). This rapid decrease in *L* was evident in a quick increase in τ_d (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.

We compared calibrated *L*τ to an independent estimate of *L* from canopy photos (Lp), which were collected nine times between Oct. 2012 and Oct. 2013. Without further calibration, the two estimates were reasonably well correlated, though the canopy photos gave slightly higher estimates of *L* (Fig. 2, R2=0.71, RMSE=0.12). This is encouraging because the photos measure *L* near the zenith (+/- 15°), while *L*τ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

We found no differences in the *L* anomaly between the C$_a$ treatments (Fig. 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped. A linear mixed-effects model with C$_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C$_a$ on the *L* anomaly (P > 0.2). What difference between C$_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the e*C*$_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).

The dynamics of *L*τ were compared with litter fall, by estimating the absolute change in *L*τ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to *L*τ for each ring separately (similar to shown for τd in Fig. 1), and calculating the change in *L*τ for each period (Δ*L*τ). Litter production correlated significantly with Δ*L*τ when the latter was separated into periods where Δ*L*τ > 0 (net leaf growth) and Δ*L*τ <  0 (net leaf loss) (Fig. 5). During periods of net leaf loss, we found a significant correlation (with a slope of XX ± se, P < 0.001?, df or n?) close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (slope, P, df…) between Δ*L*τ and litter production (Fig. 5).

We found a significant correlation among FACE rings between litter production (averaged over the entire period) and mean *L*τ during the same period (Fig. 6a), which is expected if leaf lifespan is relatively constant. This allowed us estimate mean leaf lifespan from *L*τ and litter production as the ratio of *L*τ to annual litter production, yielding a mean lifespan of 0.95 yr, with no significant difference (P? F? df?) between C$_a$ treatments (Fig. 6b).

Finally, we quantified the relationship between timing of leaf flushing and environmental drivers. During the XXX month study period, we observed four leaf flushing events in the field, with one before estimates of *L*τ were available (Fig. 7). These events occurred in spring and summer, with no clear annual cycle (Oct. 2012, Jan. 2013, Dec. 2013 and Jan. 2015). The dynamics of *L*τ from the sensor record matched these events observed in the field quite well (Fig. 6a), as estimated from a generalized additive model fitted to the average *L*τ across the six FACE rings. From the fitted model, we estimated the numerical derivative of *L* (dL/dt), as well as its confidence interval. All four leaf flushing events corresponded to significant positive derivative dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an actual leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a heavy rainfall event (Fig. 4 and 6), when temperature was ???? and the soil water content in the top 80 cm increased to ca. > 0.2 m3 m-3. Two periods of sustained decrease in *L* (red lines in Fig. 7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

# Discussion

## Response of *L* to elevated [CO$_2$]
We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated CO$_2$ (e*C*$_a$) and 3 months of ramp-up in C$_a$. We had predicted that *L* would increase in response to e*C*$_a$ as a result of increased quantum yield for photosynthesis, enhanced CO$_2$ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to e*C*$_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011). Counter to our expectations but similar to Norby et al. (2003), we found no significant effect of elevated C$_a$ on *L*. It could be argued that the change in *L* in response to e*C*$_a$ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C$_a$ effect is not due to a lack of power, but a true lack of response to elevated C$_a$.

This lack of a response to elevated C$_a$ counters our expectations and those in previous publications (Norby et al. 2003 McCarthy et al. 2010 etc.), for a number of reasons. First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to e*C*$_a$. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. 1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013). Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under e*C*$_a$. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands (White et al., 2010).

A possible explanation for the lack of a *L* response to e*C*$_a$ is that leaf growth may not be limited by carbon availability. In other words, stimulation of net leaf photosynthesis by e*C*$_a$ (a common response, also observed at the EucFACE, (Gimeno et al. in review) may not lead to increased net leaf growth, due to water or nutrient limitations (Crous et al., 2015). Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by e*C*$_a$ (De Kauwe et al. 2014, refs on turnover). We found no support for differences in leaf lifespan estimates based on *L* and litter production between C$_a$ treatments (Fig. 6), suggesting that during the study period leaf turnover rate was not affected by e*C*$_a$. To further understand the response of biomass allocation to e*C*$_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (in review) reported increased soil respiration under e*C*$_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

##Continuous measurement of *L*
Studying canopy leaf area index in an evergreen natural woodland poses unique challenges. Unlike deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan is unknown. We can also not rely on estimated leaf area based on allometric relationships with stem diameter or height, since these equations are not available for E. tereticornis, and besides these would be useless since *L* is highly variable temporally (Figs. 1, 4). We found that indirect estimates of *L* based on canopy transmittance were very reliable.  Estimated *L* compared well to that from flat canopy photos (cf. (Macfarlane et al., 2007) (Fig. 2), and after calibration, to leaf litter production during a dry period. Unlike canopy photos, PAR sensors do not need thresholding or time-consuming post-processing. We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine (Chason et al., 1991)(Dufrêne and Bréda, 1995)(Nilson, 1999). When we used unfiltered canopy transmittance data, this yielded more distinct seasonal patterns in canopy transmittance (Fig. S1), because longer path lengths in the winter result in lower overall transmittance. Using diffuse PAR only avoids having to correct for these path length effects.

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, both during periods of decreasing and increasing *L*. During a sustained period without significant rainfall (42 mm over 4 months), we used the decrease in *L* as an opportunity to calibrate against litter production. During periods of increasing *L*, litter fall increased sharply, indicating replacement of old leaves by young leaves. Indeed we observed in the field that after the significant flushing event of Dec. 2013, very few old leaves were present in the canopy. This is in line with (Pook, 1984a) in Corymbia maculata (a genus closely related to Eucalyptus), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, (Stone and Bacon, 1995) showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability (Fig. 7; Pook, 1984a), net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing. The consequence is that leaf turnover is highly variable in time.

## Consequences for phenology models
Current-generation phenology models are heavily biased towards temperate conditions (Primack and Miller-Rushing, 2011), where leaf flushing occurs in spring in response to increase in temperature (Richardson et al., 2012; Savoy and Mackay, 2015). In contrast, phenology of evergreen Eucalyptus may be more opportunistic, and shows rapid response to water availability (Fig. 7). (Specht and Brouwer, 1975) demonstrated for four Eucalyptus species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth. Similarly, (Pook, 1984b) demonstrated leaf production during most months in Corymbia maculata, except during winter and suppression by dry periods (Pook, 1984a). Thus a minimum temperature is also required for leaf growth to occur. We also have not observed leaf flushing in winter time (Ellsworth et al., unpubl. data), but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig. 7).

Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response. In a comparison of six leading ecosystem and global vegetation models,  (Powell et al., 2013) showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model. Simulations from De Kauwe et al. (2014) at ambient and elevated C$_a$ also showed little response to inter-annual variation in rainfall (pers. comm. Martin De Kauwe). In most models  that provide prognostic estimates of *L* (i.e., excluding models that take *L* as an input parameter; e.g. (Kala et al., 2013)), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool. Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as e*C*$_a$ (Fig. 2). We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests, because not all important mechanisms are incorporated in these models. In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability. Water limitation is known to enhance leaf turnover (Pook et al.1997), possibly to maintain sufficient rates of water availability to the remaining leaves (Vilagrosa et al. 2003, Misson et al. 2010). Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g. G’DAY, see Corbeels et al. 2005), but this process has been poorly quantified.

# Conclusions
Leaf area index (*L*) of a natural Eucalyptus woodland was unresponsive to elevated [CO$_2$] after 27 months of treatment and 24 months at +150ppm CO$_2$ elevation. Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage). We found no clear annual pattern in onset of leaf growth, instead, leaf growth consistently commenced after heavy rainfall following a dry period. Both the dynamics of litter production and the rapid response to soil water availability challenge current-generation prognostic *L* and phenology models. High resolution time series such as the one we present in this study should be used to develop new models of phenology for evergreen woodlands.

# Acknowledgments {.unnumbered}
Steven Wohl, Vinod Kumar, Craig Barton, litter collection: Andrew Gherlenda, Julia Cooke, Catherine Beattie
EucFACE is an initiative supported by the Australian Government through the Department of Industry and Science in partnership with the University of Western Sydney. Facilities and instrumentation at EucFACE were assembled as an initiative of the Australian Government as part of the Nation-building Economic Stimulus Package.

# References {.unnumbered}

### 2015-04-24 manuscript.Rmd
equations, typesetting, some code

@@
 source("load.R")
+library(broom)
+options(scipen = 1, digits = 2)
+
+# Some pre-calculations
+caliblm <- calibrateToDrought(facegap_cloudy_byring)$lmfit
+
+# Some definitions of constants
+.gambase <- 18
 ```
@@
 The woodland ranges in density from 600-1000 trees ha-1.
-The vegetation consists of an open woodland (basal area before treatment, BA = 27.6 ± 2.7 m$^2$ ha‑1, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E.
+The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E.
 amplifolia* Naudin.
@@
 The vegetation within three randomly selected rings was exposed to a CO$_2$ mole fraction (C$_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO$_2$ treatment rings in infrastructure and instrumentation but received ambient C$_a$, and are used as control plots.
-The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol‑1 above ambient C$_a$ in full operation mode.
+The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol$^{-1}$ above ambient C$_a$ in full operation mode.
 However, C$_a$ was increased gradually for six months at a rate of ca.
-30 μmol CO$_2$ mol-1 month-1 until full operation mode was reached (Figure 1a).
+30 μmol CO$_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).
 
-Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP ® sensor (HMP 155 Vaisala, Vantaa, Finland).
+Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland).
 Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings.
 Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers.
-Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50 depth depending on the ring).
+Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).
 
@@
 
-"*L*" _p=-loga(τ_p)/k									(1)
+(1) $$L_p = -log(\tau_p)/k$$
 
@@
 
-τ_d=Q ̅_below/Q_above								(2)
+(2) $$\tau_d = \bar{Q}_{below} / Q_{above}$$
 
-where τ_d is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}$ s-1), and Q ̅_below is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
-We then calculated daily averages of τ_d, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of τ_d).
+where $\tau_d$ is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}s^{-1}$), and Q$_{below}$ is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
+We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of $\tau_d$).
 
@@
 
-τ_d=2∫_0^(π/2)▒〖e^(-LAI∙G(θ)/cosa(θ))  sina(θ)  cosa(θ)dθ〗						(3)
+(3) $$\tau_d = 2\int_0^{\pi/2} e^{-LG(\theta)/cos(\theta)}sin(\theta)cos(\theta)d\theta$$
 
@@
 3, we calibrated *L* against litter production.
-We estimated *L* from τ_d (*L*τ) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig.
+We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig.
 1).
-We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus τ_d increased steadily, see Fig.
+We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
 1).
@@
 
-c〖(LAI〗_τ-〖LAI〗_(t=0))=∫_(t=0)^t▒〖P_L dt〗								(4)
+(4) $$c(L_{\tau,t} - L_{\tau,t=0}) = \int_0^t P_Ldt$$
 
-where c is a calibration parameter, Lt=0 is *L*τ at the start of the drought period.
+where *c* is a calibration parameter, Lt=0 is $L_\tau$ at the start of the drought period.
 We used linear regression without intercept to estimate c in Eq.
-4 as 0.68 (SE = 0.035).
-It should be stressed that even after calibration with litter production, *L*τ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq.
+4 as `r calib` (SE = `r tidy(caliblm)$std.error`).
+It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq.
 4.
@@
 We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014).
-For all time-series, we used a base of 18 when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics.
+For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics.
 In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig.
@@
 For comparison between *L* from photos and PAR transmittance, we used simple linear regression.
-To test for effects of C$_a$ treatment on *L*τ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.
+To test for effects of C$_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.
 
-To assess effects of e*C*$_a$ on *L*τ, we accounted for pre-treatment differences in canopy structure, as follows.
-The average *L*τ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig.
+To assess effects of e*C*$_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
+The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig.
 3).
 Because the EucFACE woodland is well established, the differences in BA are not due to C$_a$ treatment effects, but instead represent natural variation within the woodland.
-We then calculated the deviation of actual *L*τ from the fitted relationship with BA shown in Fig.
+We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig.
 3, to yield a ‘*L* anomaly’.
@@
 
-Canopy transmittance (τ_d) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
+Canopy transmittance ($\tau_d$) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
 1).
@@
 2014 (for description of a similar outbreak in Eucalyptus moluccana, see (Hall et al., 2015)).
-This rapid decrease in *L* was evident in a quick increase in τ_d (Fig.
+This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
 1).
@@
 
-We compared calibrated *L*τ to an independent estimate of *L* from canopy photos (Lp), which were collected nine times between Oct.
+We compared calibrated $L_\tau$ to an independent estimate of *L* from canopy photos ($L_p$), which were collected nine times between Oct.
 2012 and Oct.
@@
 2, R2=0.71, RMSE=0.12).
-This is encouraging because the photos measure *L* near the zenith (+/- 15°), while *L*τ is integrated over the sky hemisphere (Eq.
+This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
 3), weighted towards the zenith.
@@
 The average half-width of the confidence interval on the *L* anomaly (Fig.
-4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the e*C*$_a$ treatment.
+4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment.
 This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).
 
-The dynamics of *L*τ were compared with litter fall, by estimating the absolute change in *L*τ for every FACE ring, for every period when litter was collected.
-We did this by fitting a generalized additive model to *L*τ for each ring separately (similar to shown for τd in Fig.
-1), and calculating the change in *L*τ for each period (Δ*L*τ).
-Litter production correlated significantly with Δ*L*τ when the latter was separated into periods where Δ*L*τ > 0 (net leaf growth) and Δ*L*τ <  0 (net leaf loss) (Fig.
+The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected.
+We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig.
+1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$).
+Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig.
 5).
 During periods of net leaf loss, we found a significant correlation (with a slope of XX ± se, P < 0.001?, df or n?) close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
-During periods of net leaf growth, we found a significant correlation (slope, P, df…) between Δ*L*τ and litter production (Fig.
+During periods of net leaf growth, we found a significant correlation (slope, P, df…) between $\Delta L_\tau$ and litter production (Fig.
 5).

### 2015-04-26 manuscript.Rmd
More refs and numbers. Added word template (not completely right though)

@@
 title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
-  atmospheric [CO$_2$] but tracks water availability
+  atmospheric [$CO_2$] but tracks water availability
 author: "Remko Duursma"
@@
 output:
+  word_document:
+    reference_docx: msstyle.docx
+  pdf_document: default
   html_document:
     number_sections: yes
-  pdf_document: default
-  word_document: default
 csl: global-change-biology.csl
@@
 
+# Main results
+df <- facegap_cloudy_byring
+
 # Some pre-calculations
-caliblm <- calibrateToDrought(facegap_cloudy_byring)$lmfit
+caliblm <- calibrateToDrought(df)$lmfit
+
+# Number of months of study
+r <- range(df$Date)
+nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))
+
+# Range mean LAI
+dfa <- summaryBy(LAI ~ Date, FUN=mean, na.rm=TRUE, data=df)
+rangeLAI <- round(range(dfa$LAI),1)
 
@@
 .gambase <- 18
+
 ```
@@
 Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
-Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [CO$_2$] and water availability of unmanaged forests have been poorly quantified.
-We studied canopy leaf area dynamics for 27 months in a native evergreen Eucalyptus woodland exposed to Free Air CO$_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
-We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [CO$_2$].
+Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [$CO_2$] and water availability of unmanaged forests have been poorly quantified.
+We studied canopy leaf area dynamics for `r nummon` months in a native evergreen Eucalyptus woodland exposed to Free Air $CO_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
+We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [$CO_2$].
 We estimated *L* from diffuse canopy transmittance and canopy photos, and measured monthly leaf litter production.
-Contrary to expectation, *L* did not respond to elevated [CO$_2$] in this timeframe.
-We found that *L* varied between 1.0 and 2.0, but did not show a clear phenological pattern.
+Contrary to expectation, *L* did not respond to elevated [$CO_2$] in this timeframe.
+We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear phenological pattern.
 Instead, increased *L* following leaf flushing occurred a few weeks after heavy rainfall, and *L* decreased steadily during periods of low rainfall.
@@
 
-Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, (Reich, 2012), and a key variable in land surface models (Kala et al., 2013; Walker et al., 2014).
-*L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments (Herbert and Fownes, 1995; Hubbard et al., 2004; White et al., 2010) and observational studies across rainfall gradients (Ellis and Hatton, 2008; Iio et al., 2013) [@Jin2013].
-Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO$_2$] (e*C*$_a$) due to increased net carbon uptake (De Kauwe et al., 2014; Hatton et al., 1992; Kergoat et al., 2002; Zhu et al., 2012) Also expected because increased quantum yield of photosynthesis in e*C*$_a$ would allow leaf retention into deeper shade (Norby et al., 1999), but little empirical evidence is available on the response of *L* to e*C*$_a$ in mature forests (Norby and Zak 2011).
-Quantifying the magnitude and direction of the e*C*$_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C$_a$ may provide an important feedback to climate change, and modify direct physiological responses to C$_a$ (Tor-ngern et al., 2015)(Murray et al., 2012).
+Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
+*L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
+Ecosystem models generally predict that *L* increases in response to elevated atmospheric [$CO_2$] ($eC_a$) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+Also expected because increased quantum yield of photosynthesis in $eC_a$ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to $eC_a$ in mature forests [@Norby2011].
+Quantifying the magnitude and direction of the $eC_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in $C_a$ may provide an important feedback to climate change, and modify direct physiological responses to $C_a$ [@Tor-ngern2015; @Murray2012].
 
-Direct effects of e*C*$_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012).
-Although numerous studies have been conducted on the response of physiology and plant growth to e*C*$_a$, very few of these inform the response of steady-state *L* to e*C*$_a$, because it requires a sufficiently mature forest to be exposed to free air CO$_2$ enrichment (FACE) (Norby and Zak, 2011).
-Increased carbon availability through enhanced photosynthesis under e*C*$_a$ does not necessarily lead to increases in *L*.
-Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to e*C*$_a$ *L* (Norby and Zak, 2011), although positive responses to e*C*$_a$ in terms of productivity were observed at all six sites.
-Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to e*C*$_a$, leading to the prediction that stands with low *L* will show a larger response to e*C*$_a$.
-Because diminished returns in increasing LAI in response to e*C*$_a$ when LAI is larger (Norby et al., 2005)
+Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012).
+Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) (Norby and Zak, 2011).
+Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*.
+Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* (Norby and Zak, 2011), although positive responses to $eC_a$ in terms of productivity were observed at all six sites.
+Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$.
+Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger (Norby et al., 2005)
 
-In ecosystems where water availability limits productivity, we could expect e*C*$_a$ to further affect *L* (Donohue et al., 2009), since e*C*$_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability.
+In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* (Donohue et al., 2009), since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability.
 These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance.
 Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.
-In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C$_a$.
-Based on the same water-savings principle, we predict that *L* should also increase under e*C*$_a$ in water-limited forests.
+In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$.
+Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.
 
@@
 
-We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to e*C*$_a$, the EucFACE experiment.
+We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment.
 The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
-We hypothesized that e*C*$_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO$_2$.
-In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to e*C*$_a$ (Norby and Zak, 2011).
+We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$.
+In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to $eC_a$ (Norby and Zak, 2011).
 Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
@@
 Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high.
-Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted CO$_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009).
-The vegetation within three randomly selected rings was exposed to a CO$_2$ mole fraction (C$_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO$_2$ treatment rings in infrastructure and instrumentation but received ambient C$_a$, and are used as control plots.
-The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol$^{-1}$ above ambient C$_a$ in full operation mode.
-However, C$_a$ was increased gradually for six months at a rate of ca.
-30 μmol CO$_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).
+Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009).
+The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots.
+The target diurnal $C_a$ for the rings exposed to $eC_a$ was 150 μmol $CO_2$ mol$^{-1}$ above ambient $C_a$ in full operation mode.
+However, $C_a$ was increased gradually for six months at a rate of ca.
+30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).
 
@@
 Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14).
-We did not find a significant effect of C$_a$ treatment on SLA (Ellsworth, unpublished data).
+We did not find a significant effect of $C_a$ treatment on SLA (Ellsworth, unpublished data).
 
@@
 For comparison between *L* from photos and PAR transmittance, we used simple linear regression.
-To test for effects of C$_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.
+To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.
 
-To assess effects of e*C*$_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
-The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig.
+To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
+The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the $eC_a$ treatment across the FACE rings (Fig.
 3).
-Because the EucFACE woodland is well established, the differences in BA are not due to C$_a$ treatment effects, but instead represent natural variation within the woodland.
+Because the EucFACE woodland is well established, the differences in BA are not due to $C_a$ treatment effects, but instead represent natural variation within the woodland.
 We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig.
@@
 
-We found no differences in the *L* anomaly between the C$_a$ treatments (Fig.
+We found no differences in the *L* anomaly between the $C_a$ treatments (Fig.
 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped.
-A linear mixed-effects model with C$_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C$_a$ on the *L* anomaly (P > 0.2).
-What difference between C$_a$ treatments could we have reasonably detected?
+A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (P > 0.2).
+What difference between $C_a$ treatments could we have reasonably detected?
 The average half-width of the confidence interval on the *L* anomaly (Fig.
-4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment.
+4) was 0.047 m$^2$ m$^{-2}$ for $aC_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment.
 This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).
@@
 F?
-df?) between C$_a$ treatments (Fig.
+df?) between $C_a$ treatments (Fig.
 6b).
@@
 
-## Response of *L* to elevated [CO$_2$]
-We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated CO$_2$ (e*C*$_a$) and 3 months of ramp-up in C$_a$.
-We had predicted that *L* would increase in response to e*C*$_a$ as a result of increased quantum yield for photosynthesis, enhanced CO$_2$ assimilation for the entire canopy, and increased soil water availability.
-We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to e*C*$_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011).
+## Response of *L* to elevated [$CO_2$]
+We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$.
+We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability.
+We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011).
 Counter to our expectations but similar to Norby et al.
-(2003), we found no significant effect of elevated C$_a$ on *L*.
-It could be argued that the change in *L* in response to e*C*$_a$ was too small to be detected.
+(2003), we found no significant effect of elevated $C_a$ on *L*.
+It could be argued that the change in *L* in response to $eC_a$ was too small to be detected.
 Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
@@
 Using this method, we estimated that we can detect a difference of only a 6% in *L*.
-Based on this small detectable difference, we conclude that the lack of a significant C$_a$ effect is not due to a lack of power, but a true lack of response to elevated C$_a$.
+Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.
 
-This lack of a response to elevated C$_a$ counters our expectations and those in previous publications (Norby et al.
+This lack of a response to elevated $C_a$ counters our expectations and those in previous publications (Norby et al.
 2003 McCarthy et al.
 2010 etc.), for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to e*C*$_a$.
+First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to $eC_a$.
 They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
@@
 1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013).
-Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under e*C*$_a$.
+Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under $eC_a$.
 It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE.
@@
 
-A possible explanation for the lack of a *L* response to e*C*$_a$ is that leaf growth may not be limited by carbon availability.
-In other words, stimulation of net leaf photosynthesis by e*C*$_a$ (a common response, also observed at the EucFACE, (Gimeno et al.
+A possible explanation for the lack of a *L* response to $eC_a$ is that leaf growth may not be limited by carbon availability.
+In other words, stimulation of net leaf photosynthesis by $eC_a$ (a common response, also observed at the EucFACE, (Gimeno et al.
 in review) may not lead to increased net leaf growth, due to water or nutrient limitations (Crous et al., 2015).
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.
-Either process may itself be affected by e*C*$_a$ (De Kauwe et al.
+Either process may itself be affected by $eC_a$ (De Kauwe et al.
 2014, refs on turnover).
-We found no support for differences in leaf lifespan estimates based on *L* and litter production between C$_a$ treatments (Fig.
-6), suggesting that during the study period leaf turnover rate was not affected by e*C*$_a$.
-To further understand the response of biomass allocation to e*C*$_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
+We found no support for differences in leaf lifespan estimates based on *L* and litter production between $C_a$ treatments (Fig.
+6), suggesting that during the study period leaf turnover rate was not affected by $eC_a$.
+To further understand the response of biomass allocation to $eC_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
 For example, Drake et al.
-(in review) reported increased soil respiration under e*C*$_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
+(in review) reported increased soil respiration under $eC_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
 
@@
 Simulations from De Kauwe et al.
-(2014) at ambient and elevated C$_a$ also showed little response to inter-annual variation in rainfall (pers.
+(2014) at ambient and elevated $C_a$ also showed little response to inter-annual variation in rainfall (pers.
 comm.
@@
 (Kala et al., 2013)), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
-Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as e*C*$_a$ (Fig.
+Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as $eC_a$ (Fig.
 2).
@@
 # Conclusions
-Leaf area index (*L*) of a natural Eucalyptus woodland was unresponsive to elevated [CO$_2$] after 27 months of treatment and 24 months at +150ppm CO$_2$ elevation.
+Leaf area index (*L*) of a natural Eucalyptus woodland was unresponsive to elevated [$CO_2$] after 27 months of treatment and 24 months at +150ppm $CO_2$ elevation.
 Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage).

### 2015-04-27 manuscript.Rmd
Lots of references fixed

@@
   word_document:
-    reference_docx: msstyle.docx
+    reference_docx: manuscriptstyle.docx
   pdf_document: default
@@
 
-Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012).
-Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) (Norby and Zak, 2011).
+Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
+Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) [@Norby2011].
 Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*.
-Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* (Norby and Zak, 2011), although positive responses to $eC_a$ in terms of productivity were observed at all six sites.
-Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$.
-Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger (Norby et al., 2005)
+Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* [@Norby2011], although positive responses to $eC_a$ in terms of productivity were observed at all six sites.
+Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$.
+Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger [@Norby2005].
 
-In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* (Donohue et al., 2009), since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability.
+In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
 These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance.
-Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.
-In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$.
+**Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al.
+[-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$.
 Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.
@@
 Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g.
-(Richardson et al., 2012; Savoy and Mackay, 2015) whereas *L* responses to fluctuations in water availability are qualitatively well-known (more cites), but poorly quantified.
-The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability (Pook, 1985; Specht and Specht, 1989; Whitehead and Beadle, 2004)(Pook et al., 1997)(Limousin et al., 2009).The strong relationship between *L* and rainfall observed across Australia (Ellis and Hatton, 2008; Specht and Specht, 1989), and other continents (Donohue et al., 2013; Grier and Running, 1977; [@Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
-However, local *L* is also highly responsive to water availability at shorter time scales (Pook, 1985; Pook et al., 1997), driven by leaf turnover (Misson et al., 2011) and leaf flushing and expansion (Muller et al., 2011), which are both highly responsive to soil water availability.
-Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology (Misson et al., 2011).
+[@Richardson2012; @Savoy2015]) whereas *L* responses to fluctuations in water availability are qualitatively well-known (**more cites**), but poorly quantified.
+The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009].The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
+However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
+Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].
 
@@
 We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$.
-In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to $eC_a$ (Norby and Zak, 2011).
+In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to $eC_a$ [@Norby2011].
 Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
@@
 
-We then estimated leaf area index (*L*) from τ_d (*L*τ), following (Campbell and Norman, 2000).
+We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following (Campbell and Norman, 2000).
 We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
@@
 
-where θ is the solar zenith angle, G(θ) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000).
+where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000).
 Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 (Stenberg, 2006).
 We solved numerically for *L* in Eq.
-3, using a smooth integration over θ, yielding *L*τ.
+3, using a smooth integration over $\theta$, yielding *L*τ.
 
@@
 During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period).
-A total of n=11 days with sufficiently diffuse conditions to estimate τ_d  were available during this period.
-In theory, cumulative *L* leaf litter production (PL, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance.
+A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
+In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance.
 We calibrated equation (4),
@@
 
-where *c* is a calibration parameter, Lt=0 is $L_\tau$ at the start of the drought period.
-We used linear regression without intercept to estimate c in Eq.
+where *c* is a calibration parameter, $L_{\tau,t=0}$ is $L_\tau$ at the start of the drought period.
+We used linear regression without intercept to estimate *c* in Eq.
 4 as `r calib` (SE = `r tidy(caliblm)$std.error`).
@@
 ## Data analysis
-We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014).
+We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries (Wood, 2006), using the *mgcv* package in R [@RCT2015].
 For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics.
@@
 
-An outbreak of psyllids (Cardiaspina sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept.
-2014 (for description of a similar outbreak in Eucalyptus moluccana, see (Hall et al., 2015)).
+An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept.
+2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see (Hall et al., 2015)).
 This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
@@
 ## Response of *L* to elevated [$CO_2$]
-We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$.
+We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$.
 We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability.
@@
 2003 McCarthy et al.
-2010 etc.), for a number of reasons.
+2007), for a number of reasons.
 First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to $eC_a$.
@@
 2013, very few old leaves were present in the canopy.
-This is in line with (Pook, 1984a) in Corymbia maculata (a genus closely related to Eucalyptus), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
+This is in line with (Pook, 1984a) in *Corymbia maculata* (a genus closely related to *Eucalyptus*), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
 Similarly, (Stone and Bacon, 1995) showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
@@
 (Specht and Brouwer, 1975) demonstrated for four Eucalyptus species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth.
-Similarly, (Pook, 1984b) demonstrated leaf production during most months in Corymbia maculata, except during winter and suppression by dry periods (Pook, 1984a).
+Similarly, (Pook, 1984b) demonstrated leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods (Pook, 1984a).
 Thus a minimum temperature is also required for leaf growth to occur.

### 2015-04-27 manuscript.Rmd
abstract minor fix

@@
 Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [$CO_2$] and water availability of unmanaged forests have been poorly quantified.
-We studied canopy leaf area dynamics for `r nummon` months in a native evergreen Eucalyptus woodland exposed to Free Air $CO_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
+We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air $CO_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
 We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [$CO_2$].
-We estimated *L* from diffuse canopy transmittance and canopy photos, and measured monthly leaf litter production.
+We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production.
 Contrary to expectation, *L* did not respond to elevated [$CO_2$] in this timeframe.
-We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear phenological pattern.
-Instead, increased *L* following leaf flushing occurred a few weeks after heavy rainfall, and *L* decreased steadily during periods of low rainfall.
+We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear annual pattern.
+Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady during periods of low rainfall.
 Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
+Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly increased under elevated [$CO_2$] (p = 0.051).
 We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.

### 2015-05-01 manuscript.Rmd
references

@@
 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay.
-The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated (Crous et al., 2015).
+The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015].
 Groundwater depth is 12.6 – 12.9 m below the soil surface.
-The woodland ranges in density from 600-1000 trees ha-1.
+The woodland ranges in density from 600-1000 trees $ha^{-1}$.
 The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E.
 amplifolia* Naudin.
-The understory stratum is composed mainly of grass species dominated by Microlaena stipoides (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst.
+The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst.
 & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg.
@@
 Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high.
-Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009).
+Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
 The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots.
@@
 ## Leaf litter production
-In each ring, 8 litter cloth baskets with an area of 0.1979 m$^2$ were installed at 0.5m above the forest floor.
+In each ring, 8 litter cloth baskets with an area of 0.1979 $m^2$ were installed at 0.5m above the forest floor.
 Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc.
 Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14).
-We did not find a significant effect of $C_a$ treatment on SLA (Ellsworth, unpublished data).
+We did not find a significant effect of $C_a$ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).
 
 ## Estimation of leaf area index from digital photography
-Leaf area index (*L*) was estimated from zenith canopy photographs following (Macfarlane et al., 2007).
+Leaf area index (*L*) was estimated from zenith canopy photographs following [@Macfarlane2007].
 From October 2012 to October 2013, we took ca.
@@
 Following recommendations by Jonckheere et al.
-(2005), we used the automated Ridler algorithm to threshold the canopy photos.
+[-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos.
 We then calculated canopy transmittance (τ_p) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.

### 2015-05-01 manuscript.Rmd
Lots of progress on manuscript, most references done, all (?) stats added. captions to be done and intro + abstract to be rewritten

@@
   atmospheric [$CO_2$] but tracks water availability
-author: "Remko Duursma"
-date: "April 24, 2015"
+author: ""
+date: ""
 output:
@@
 source("load.R")
+
 library(broom)
+library(lme4)
+library(lmerTest)
+library(car)
+
+# Number of digits in printing of numbers
 options(scipen = 1, digits = 2)
+
+# Some definitions of constants
+.gambase <- 18
 
@@
 
-# Some definitions of constants
-.gambase <- 18
+# Comparison flat canopy photo with transmittance
+lmphoto <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)
+
+# Significance CO2 on LAI anomaly
+df$Time <- as.factor(df$Date - min(df$Date))
+m3 <- lmer(LAIanomaly ~ treatment*Time + (1|Ring), data=df)
+LanomAOV <- Anova(m3)
+
+# Average CI width in Fig 4.
+fit <- smoothplot(Date, LAIanomaly, g=treatment, R="Ring", 
+                  data=facegap_cloudy_byring, plot=FALSE, kgam=.gambase)
+pa <- predict(fit$ambient, data.frame(X=as.numeric(facegap_cloudy_byring$Date)), se.fit=TRUE)
+pe <- predict(fit$elevated, data.frame(X=as.numeric(facegap_cloudy_byring$Date)), se.fit=TRUE)
+LanomCIamb <- mean(pa$se.fit)*2
+LanomCIele <- mean(pe$se.fit)*2
+
+# Detectable Ca effect on LAI
+Ldetect <- 100 * (LanomCIamb + LanomCIele) / mean(ba$LAI.mean)
+
+# Litter - dLAI correlations Figure 4
+dl <-dLAIlitter
+dl$Y <- with(dl, dLAI * 30.5/ndays)
+dl$X <- with(dl, dLAIlitter.mean * 30.5/ndays)
+dlup <- dl[dl$LAIchange == "increasing",]
+dldown <- dl[dl$LAIchange == "decreasing",]
+lmlitup <- lm(Y ~ X, data=dlup)
+lmlitdown <- lm(Y ~ X, data=dldown)
+
+# Leaf lifespan
+LLtest <- tidy(t.test(LL ~ treatment,data=ba))
 
 ```
+
+\
+$\mathrm{Remko~A.~Duursma}^1$, $\mathrm{Teresa~E.~Gimeno}^{1,2}$, $\mathrm{Matthias~M.~Boer}^1$, $\mathrm{Kristine~Y.~Crous}^1$, $\mathrm{Mark~G.~Tjoelker}^1$, $\mathrm{David~S.~Ellsworth}^1$
+
+\
+$~^1$ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia
+
+\
+$~^2$ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France
 
@@
 
+\
 *Keywords*: leaf area index, FACE, phenology, drought, litter production
@@
 
+\
 Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
@@
 Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$.
-Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger [@Norby2005].
+Because diminished returns in increasing *L* in response to $eC_a$ when LAI is larger [@Norby2005].
 
+\
 In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
@@
 
+\
 Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g.
@@
 
+\
 We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment.
@@
 
+\
 The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil.
@@
 
+\
 Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’).
@@
 However, $C_a$ was increased gradually for six months at a rate of ca.
-30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).
+30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Fig.
+1a).
 
+\
 Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland).
@@
 
-## Estimation of leaf area index from digital photography
-Leaf area index (*L*) was estimated from zenith canopy photographs following [@Macfarlane2007].
-From October 2012 to October 2013, we took ca.
-monthly canopy photographs in each FACE ring.
-Photographs were taken in the morning on cloudy days.
-We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring.
-Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded.
-The camera was aligned with the north-south axis and set at 1.25 m height.
-Camera lens was pointed directly upwards and levelled using a three axis bubble level.
-We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan).
-To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens.
-Following recommendations by Jonckheere et al.
-[-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos.
-We then calculated canopy transmittance (τ_p) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.
-*L* from canopy photos (Lp) was calculated with a simple Beer-Lambert law (Eq.
-1).
-
-(1) $$L_p = -log(\tau_p)/k$$
-
-where τ_p canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following (Macfarlane et al., 2007) recommendation for eucalypt forests.
-
 ## *L* from diffuse transmittance
 In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al.
-2003.
+[-@Norby2003].
 Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct.
@@
 
-We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following (Campbell and Norman, 2000).
+\
+We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following Campbell and Norman [-@Campbell2000].
 We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
@@
 
-where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000).
-Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 (Stenberg, 2006).
+where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000].
+Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 [@Stenberg2006].
 We solved numerically for *L* in Eq.
-3, using a smooth integration over $\theta$, yielding *L*τ.
+3, using a smooth integration over $\theta$, yielding $L_\tau$.
 
+\
 To overcome violations in the assumptions underlying Eq.
@@
 In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance.
-We calibrated equation (4),
+We calibrated Eq.
+4,
 
@@
 
+## Estimation of leaf area index from digital photography
+
+We took advantage of the detailed timeseries of *L* estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al.
+[-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests.
+From October 2012 to October 2013, we took ca.
+monthly canopy photographs in each FACE ring.
+Photographs were taken in the morning on cloudy days.
+We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring.
+Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded.
+The camera was aligned with the north-south axis and set at 1.25 m height.
+Camera lens was pointed directly upwards and levelled using a three axis bubble level.
+We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan).
+To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens.
+Following recommendations by Jonckheere et al.
+[-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos.
+We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.
+*L* from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq.
+1).
+
+(1) $$L_p = -log(\tau_p)/k$$
+
+where $\tau_p$ canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following Macfarlane et al.
+[-@Macfarlane2007] recommendation for eucalypt forests.
+
+\
+We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$).
+Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig.
+SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`).
+This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
+3), weighted towards the zenith.
+
 ## Data analysis
-We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries (Wood, 2006), using the *mgcv* package in R [@RCT2015].
+We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015].
 For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics.
@@
 7).
-We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014).
+We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson [@Simpson2014].
 For comparison between *L* from photos and PAR transmittance, we used simple linear regression.
@@
 
+\
 To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
@@
 
-Canopy transmittance ($\tau_d$) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
+Canopy transmittance ($\tau_d$) varied between `r min(df$Gapfraction.mean)` and `r max(df$Gapfraction.mean)`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
 1).
@@
 
+\
 An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept.
-2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see (Hall et al., 2015)).
+2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see [@Hall2015]).
 This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
 1).
-However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.
+However, it recovered rapidly after rainfall in December 2014, resulting in a lower $\tau_d$ than before the insect attack.
 
-We compared calibrated $L_\tau$ to an independent estimate of *L* from canopy photos ($L_p$), which were collected nine times between Oct.
-2012 and Oct.
-2013.
-Without further calibration, the two estimates were reasonably well correlated, though the canopy photos gave slightly higher estimates of *L* (Fig.
-2, R2=0.71, RMSE=0.12).
-This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
-3), weighted towards the zenith.
-
-We found no differences in the *L* anomaly between the $C_a$ treatments (Fig.
-4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped.
-A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (P > 0.2).
+\
+We found no differences in the *L* anomaly between the $C_a$ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig.
+3).
+A brief exception was the period where one of the rings was affected by an insect outbreak (see above).
+A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (p = `r LanomAOV[[3]][1]`).
 What difference between $C_a$ treatments could we have reasonably detected?
 The average half-width of the confidence interval on the *L* anomaly (Fig.
-4) was 0.047 m$^2$ m$^{-2}$ for $aC_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).
+4) was `r LanomCIamb` $m^2$ $m^{-2}$ for $aC_a$, and `r LanomCIele` $m^2$ $m^{-2}$ for the $eC_a$ treatment.
+This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` $m^2$ $m^{-2}$).
 
+\
 The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected.
@@
 Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig.
-5).
-During periods of net leaf loss, we found a significant correlation (with a slope of XX ± se, P < 0.001?, df or n?) close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
-During periods of net leaf growth, we found a significant correlation (slope, P, df…) between $\Delta L_\tau$ and litter production (Fig.
+4).
+During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
+During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
 5).
 
-We found a significant correlation among FACE rings between litter production (averaged over the entire period) and mean *L*τ during the same period (Fig.
-6a), which is expected if leaf lifespan is relatively constant.
-This allowed us estimate mean leaf lifespan from *L*τ and litter production as the ratio of *L*τ to annual litter production, yielding a mean lifespan of 0.95 yr, with no significant difference (P?
-F?
-df?) between $C_a$ treatments (Fig.
-6b).
+\
+We estimated gross leaf production from measured litter production ($P_L$) and temporal dynamics of $L_\tau$.
+Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to $P_L$ to arrive at an estimate of the leaf production.
+The temporal dynamics of litter and leaf production are shown in Fig.
+5, and demonstrate the irregular temporal pattern in leaf production and $P_L$.
+We found no significant difference in either litter or leaf production between $C_a$ treatments (p > 0.2 for both, assessed with linear-mixed effects models).
 
+\
+Next, we related $P_L$ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig.
+6a).
+If leaf lifespan is constant, $P_L$ would be linearly related to $L_\tau$.
+Instead, we found that in $eC_a$, $P_L$ was lower at a given $L_\tau$ as compared to $aC_a$.
+We estimated mean leaf lifespan from Fig.
+6a as the ratio of $L_\tau$ to annual $P_L$.
+We found a significant (p = `r LLtest$p.value`) increase in LL with elevated $C_a$ (Fig.
+6b), with LL = `r LLtest$estimate1`yr in $aC_a$ and `r LLtest$estimate2`yr in $eC_a$.
+
+\
 Finally, we quantified the relationship between timing of leaf flushing and environmental drivers.
-During the XXX month study period, we observed four leaf flushing events in the field, with one before estimates of *L*τ were available (Fig.
+During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $L_\tau$ were available (Fig.
 7).
@@
 2015).
-The dynamics of *L*τ from the sensor record matched these events observed in the field quite well (Fig.
-6a), as estimated from a generalized additive model fitted to the average *L*τ across the six FACE rings.
+The dynamics of $L_\tau$ from the sensor record matched these events observed in the field quite well (Fig.
+7a), as estimated from a generalized additive model fitted to the average $L_\tau$ across the six FACE rings.
 From the fitted model, we estimated the numerical derivative of *L* (dL/dt), as well as its confidence interval.
-All four leaf flushing events corresponded to significant positive derivative dL/dt.
+All four leaf flushing events corresponded to significant positive dL/dt.
 One brief significant positive dL/dt (June 2013) did not correspond to an actual leaf flushing event, but it was small in magnitude.
 It is also obvious that every leaf flush event occurred shortly after a heavy rainfall event (Fig.
-4 and 6), when temperature was ????
-and the soil water content in the top 80 cm increased to ca.
-> 0.2 m3 m-3.
+4 and 6), when the soil water content in the top 80 cm increased to ca.
+> 0.2 $m^3$ $m^{-3}$.
 Two periods of sustained decrease in *L* (red lines in Fig.
@@
 ## Response of *L* to elevated [$CO_2$]
-We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$.
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$.
 We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability.
-We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011).
+We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands [@Norby2011].
 Counter to our expectations but similar to Norby et al.
-(2003), we found no significant effect of elevated $C_a$ on *L*.
+[-@Norby2003], we found no significant effect of elevated $C_a$ on *L*.
 It could be argued that the change in *L* in response to $eC_a$ was too small to be detected.
@@
 To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate.
-Using this method, we estimated that we can detect a difference of only a 6% in *L*.
+Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*.
 Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.
 
-This lack of a response to elevated $C_a$ counters our expectations and those in previous publications (Norby et al.
-2003 McCarthy et al.
-2007), for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to $eC_a$.
+\
+This lack of a response to elevated $C_a$ counters our expectations and those in previous publications [@Norby2003; @Mccarthy2007], for a number of reasons.
+First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to $eC_a$.
 They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
 The mean *L* at the EucFACE was ca.
-1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013).
-Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under $eC_a$.
+`r mean(ba$LAI.mean)` $m^2$ $m^{-2}$, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
+[-@Iio2013].
+Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under $eC_a$.
 It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE.
-Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands (White et al., 2010).
+Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands [@White2010].
 
-A possible explanation for the lack of a *L* response to $eC_a$ is that leaf growth may not be limited by carbon availability.
-In other words, stimulation of net leaf photosynthesis by $eC_a$ (a common response, also observed at the EucFACE, (Gimeno et al.
-in review) may not lead to increased net leaf growth, due to water or nutrient limitations (Crous et al., 2015).
+\
+A possible explanation for the lack of a *L* response to $eC_a$ is that leaf growth may not be directly related to increased carbon availability as a result of $eC_a$.
+In other words, stimulation of net leaf photosynthesis by $eC_a$ (a common response, also observed at the EucFACE, Gimeno et al.
+in review) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.
-Either process may itself be affected by $eC_a$ (De Kauwe et al.
-2014, refs on turnover).
-We found no support for differences in leaf lifespan estimates based on *L* and litter production between $C_a$ treatments (Fig.
-6), suggesting that during the study period leaf turnover rate was not affected by $eC_a$.
+Either process may itself be affected by $eC_a$ [@DeKauwe2014].
+We found a significant effect of $eC_a$ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated $C_a$ (Fig.
+6).
+Because we found no difference in *L* with $C_a$ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
 To further understand the response of biomass allocation to $eC_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
@@
 
+\
+Other studies in FACE are consistent with longer leaf-lifespan in elevated Ca shown here.
+
 ##Continuous measurement of *L*
 Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
-Unlike deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan is unknown.
-We can also not rely on estimated leaf area based on allometric relationships with stem diameter or height, since these equations are not available for E.
-tereticornis, and besides these would be useless since *L* is highly variable temporally (Figs.
+Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan is unknown.
+We can also not rely on estimated leaf area based on allometric relationships with stem diameter or height, since these equations are not available for mature *E.
+tereticornis*, and besides these would be useless since *L* is highly variable temporally (Figs.
 1, 4).
@@
 Estimated *L* compared well to that from flat canopy photos (cf.
-(Macfarlane et al., 2007) (Fig.
-2), and after calibration, to leaf litter production during a dry period.
+[@Macfarlane2007] (Fig.
+SI2), and after calibration, to leaf litter production during a dry period (Fig.
+4).
 Unlike canopy photos, PAR sensors do not need thresholding or time-consuming post-processing.
-We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine (Chason et al., 1991)(Dufrêne and Bréda, 1995)(Nilson, 1999).
-When we used unfiltered canopy transmittance data, this yielded more distinct seasonal patterns in canopy transmittance (Fig.
-S1), because longer path lengths in the winter result in lower overall transmittance.
+We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999].
+As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig.
+S3), because longer path lengths in the winter result in lower overall transmittance.
 Using diffuse PAR only avoids having to correct for these path length effects.
@@
 We found a correlation between *L* dynamics and litter production, both during periods of decreasing and increasing *L*.
-During a sustained period without significant rainfall (42 mm over 4 months), we used the decrease in *L* as an opportunity to calibrate against litter production.
+During a four-month period in 2013 without significant rainfall, we used the decrease in *L* as an opportunity to calibrate against litter production.
 During periods of increasing *L*, litter fall increased sharply, indicating replacement of old leaves by young leaves.
@@
 2013, very few old leaves were present in the canopy.
-This is in line with (Pook, 1984a) in *Corymbia maculata* (a genus closely related to *Eucalyptus*), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
-Similarly, (Stone and Bacon, 1995) showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
-Since leaf flushing has been shown to be closely related to soil water availability (Fig.
-7; Pook, 1984a), net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing.
+This is in line with [@Pook1984] in *Corymbia maculata* (a genus closely related to *Eucalyptus*), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
+Similarly, [@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
+Since leaf flushing has been shown to be closely related to soil water availability [Fig.
+7; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig.
+5).
 The consequence is that leaf turnover is highly variable in time.
@@
 ## Consequences for phenology models
-Current-generation phenology models are heavily biased towards temperate conditions (Primack and Miller-Rushing, 2011), where leaf flushing occurs in spring in response to increase in temperature (Richardson et al., 2012; Savoy and Mackay, 2015).
-In contrast, phenology of evergreen Eucalyptus may be more opportunistic, and shows rapid response to water availability (Fig.
+Current-generation phenology models are heavily biased towards temperate conditions [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardsonet2012; @Savoy2015].
+In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig.
 7).
-(Specht and Brouwer, 1975) demonstrated for four Eucalyptus species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth.
-Similarly, (Pook, 1984b) demonstrated leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods (Pook, 1984a).
+Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth.
+Similarly, Pook  [-@Pook1984a] demonstrated leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods.
 Thus a minimum temperature is also required for leaf growth to occur.
-We also have not observed leaf flushing in winter time (Ellsworth et al., unpubl.
-data), but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig.
+We also have not observed leaf flushing in winter time, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig.
 7).
 
+\
 Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response.
-In a comparison of six leading ecosystem and global vegetation models,  (Powell et al., 2013) showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model.
+In a comparison of six leading ecosystem and global vegetation models,  Powell et al.
+[-@Powell2013] showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model.
 Simulations from De Kauwe et al.
-(2014) at ambient and elevated $C_a$ also showed little response to inter-annual variation in rainfall (pers.
+[-@DeKauwe2014] at ambient and elevated $C_a$ also showed little response to inter-annual variation in rainfall (pers.
 comm.
 Martin De Kauwe).
-In most models  that provide prognostic estimates of *L* (i.e., excluding models that take *L* as an input parameter; e.g.
-(Kala et al., 2013)), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
+In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g.
+@Kala2013), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
 Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as $eC_a$ (Fig.
@@
 In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability.
-Water limitation is known to enhance leaf turnover (Pook et al.1997), possibly to maintain sufficient rates of water availability to the remaining leaves (Vilagrosa et al.
-2003, Misson et al.
-2010).
+Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2010], and rates of leaf expansion [@Muller2011].
 Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g.
@@
 # Conclusions
-Leaf area index (*L*) of a natural Eucalyptus woodland was unresponsive to elevated [$CO_2$] after 27 months of treatment and 24 months at +150ppm $CO_2$ elevation.
+Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [$CO_2$] after 27 months of treatment and 24 months at +150ppm $CO_2$ elevation.
 Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage).
-We found no clear annual pattern in onset of leaf growth, instead, leaf growth consistently commenced after heavy rainfall following a dry period.
-Both the dynamics of litter production and the rapid response to soil water availability challenge current-generation prognostic *L* and phenology models.
-High resolution time series such as the one we present in this study should be used to develop new models of phenology for evergreen woodlands.
+Leaf lifespan, estimated from mean *L* and annual litter production, was `r round(abs(LLtest$estimate*12),1)` months longer in the elevated [$CO_2$] treatment.
+
+We found no clear annual pattern in dynamics of *L*, instead, a net increase in *L*, indicating leaf growth, consistently commenced after heavy rainfall following a dry period.
+The complex dynamics of litter production and the rapid response of litter production and leaf growth to soil water availability challenge current-generation prognostic *L* and phenology models.
+High resolution time series such as the one we present in this study can be used to develop new models of phenology for evergreen woodlands.
 
 # Acknowledgments {.unnumbered}
-Steven Wohl, Vinod Kumar, Craig Barton, litter collection: Andrew Gherlenda, Julia Cooke, Catherine Beattie
+We thank Steven Wohl, Vinod Kumar, Craig Barton and Craig McNamara for maintaining the FACE site and its wide array of sensors, and Andrew Gherlenda, Julia Cooke and Catherine Beattie for significant efforts in separating and weighing the litter samples.
+We also thank Gerry Devine for support in data management.
 EucFACE is an initiative supported by the Australian Government through the Department of Industry and Science in partnership with the University of Western Sydney.

### 2015-05-01 R/figures.R

@@ -157,8 +157,8 @@ figure4 <- function(df){
   
   abline(h=0, lty=5)
   abline(0,-1)
-  predline(lm(Y ~ X, data=df, subset=dLAI>0), col=Cols[2])
-  predline(lm(Y ~ X, data=df, subset=dLAI<0), col=Cols[1])
+  predline(lm(Y ~ X, data=dfup), col=Cols[2])
+  predline(lm(Y ~ X, data=dfdown), col=Cols[1])
   box()
   
   l <- legend("bottomright", c(expression(Delta*LAI < 0),expression(Delta*LAI > 0)), 

### 2015-05-01 R/stats-analysis.R

@@ -20,15 +20,11 @@ m2 <- lmer(LAI ~ treatment*Time + (1|Ring), data=df)
 Anova(m2)
 
 # LAI anomaly
-df <- facegap_cloudy_byring
-df$Time <- as.factor(df$Date - min(df$Date))
-m3 <- lmer(LAIanomaly ~ treatment*Time + (1|Ring), data=df)
-Anova(m3)
 
 # LAI production
 df <- dLAIlitter
 df$Time <- as.factor(df$Date - min(df$Date))
-df$leafprod <- with(df, dLAI + dLAI.mean)
+df$leafprod <- with(df, dLAI + dLAIlitter.mean)
 m4 <- lmer(leafprod ~ treatment*Time + (1|Ring), data=df)
 Anova(m4)
 

### 2015-05-01 load.R

@@ -14,6 +14,7 @@ library(gplots)
 library(mgcv)
 library(stringr)
 
+
 if(packageVersion('HIEv') < '0.6.4')
   stop("Please update the HIEv package!")
 

### 2015-05-01 manuscript.Rmd
fixed 4 missing refs

@@
 \
-This lack of a response to elevated $C_a$ counters our expectations and those in previous publications [@Norby2003; @Mccarthy2007], for a number of reasons.
+This lack of a response to elevated $C_a$ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons.
 First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to $eC_a$.
@@
 ## Consequences for phenology models
-Current-generation phenology models are heavily biased towards temperate conditions [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardsonet2012; @Savoy2015].
+Current-generation phenology models are heavily biased towards temperate conditions [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardson2012; @Savoy2015].
 In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig.
@@
 In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability.
-Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2010], and rates of leaf expansion [@Muller2011].
+Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and rates of leaf expansion [@Muller2011].
 Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g.

### 2015-05-01 manuscript.Rmd
added figure captions

@@
 
+# Figure captions {.unnumbered}
+
+Figure 1.
+(a) Target [CO2] (Ca) relative to the ambient Ca treatment at the EucFACE.
+The elevated Ca treatment proceeded in steps of 30 ppm, reaching +150 ppm in Feb.
+2013.
+(b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions (Fdiff > 0.98) only.
+The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model).
+Lines in blue colours denote ambient Ca treatment, red colours are elevated Ca (+150 ppm).
+Note steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
+The arrow labelled ‘calibration’ indicates a drought period when τ_dsteadily increased, used to calibrate *L* estimates to cumulative litter production.
+
+Figure 2.
+Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings (R2 = 0.87, L = 0.909 + 0.0175 * BA).
+This relationship was used to calculate the L anomaly in Fig.
+4, as the difference between L predicted from this regression and the actual $L_\tau$ at any given time.
+
+Figure 3.Time series of L after correcting for plot-to-plot differences in basal area (BA).
+At every time point, the L anomaly was calculated as Lτ minus L fitted from the relationship with BA shown in Fig.
+3.
+Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect.
+Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05, except for a brief period in late 2014.
+This difference was due to an insect attack that led to partial defoliation of the canopy in one FACE ring (see Fig.
+1).
+
+Figure 4.
+Comparison of absolute changes in L ($\Delta L$) to litter production (in units of *L*), calculated for each litter collection period.
+To calculate ΔL, we fit a generalized additive model fitted to data from each FACE ring separately, and estimated the change in L during the litter collection period (and expressed as monthly changes, by accounting for the differences in actual lengths of litter collection periods).
+Green symbols indicate periods during which ΔL was positive (net leaf production), orange symbols when ΔL was negative (net leaf loss).
+Solid black lines are 1:1 lines, coloured lines regression lines with 95% confidence intervals.
+The strong positive correlation between litter production and ΔL during periods of net leaf production indicates that during leaf flushing, old leaves are actively shed.
+
+Figure 5. Timeseries of litter production and estimated gross leaf production rate.
+
+Figure 6.
+(a) Total litter production for each of the FACE rings plotted against mean Lτ (p < 0.02).
+The ratio of mean L to annual litter production gives the mean leaf life span, shown in (b).
+Leaf lifespan was not different between Ca treatments (p = 0.11).
+
+Figure 7.
+Dynamics of L in relation to environmental drivers.
+(a) Smoothed time series of L estimated from τ_d, averaged across all six FACE rings.
+(b) From panel (a), we calculated the daily change in L (dL/dt), including the 95% confidence interval for the derivative of the time series.
+Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at α=0.05).
+Four of the five significant positive increases corresponded to leaf flushing (arrows) observed in the field (except the third brief significant increase).
+(c) Soil water content estimated from over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).
+(d) Average daily air temperature with a fitted smoother.
+
+Figure S1.
+Increase in gapfraction ($\tau_d$) and total litter production during the 2013 drought, calculated for each ring separately.
+Blue colours are $aC_a$, red colours eCa.
+(a) The change in τ_d and the 95% confidence interval was calculated from the slope of linear regression of τ_d with time (days since 8 July 2013)
+
+Figure S2.
+Correlation between L estimated from diffuse canopy transmittance (τ_d) and from canopy photos.
+L from τ_d was calibrated to litter production during the 2013 drought (see Fig.
+6?), but no further calibration was done for the comparison to canopy photos as shown here.
+Solid line is a 1:1 line, dashed line a regression line with a slope of 0.98 ± se (qirh , R2=0.71,P < 0.01), with the shading indicating the 95% confidence interval.
+
+Figure S3.
+Comparison of τ_d  calculated from canopy transmittance during highly diffuse conditions (Fdiff > 0.98) only, or using all PAR data without filtering.
+Shown are daily average τ_d  for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals).
+Note the underestimation of τ_(d )during southern hemisphere winter, when solar angles are low, leading to higher attenuation of the direct beam, due to a longer path length through the canopy.
+
 # References {.unnumbered}

### 2015-05-04 manuscript.Rmd
Manuscript format OK; captions all done; proper sub and superscripts; template uses pagebreak before Header1; Modified FigureSI2

--- 2015-05-01 manuscript.Rmd
@@
 title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
-  atmospheric [$CO_2$] but tracks water availability
-author: ""
-date: ""
+  atmospheric [CO~2~] but tracks water availability
+author: ''
+date: ''
 output:
@@
 library(car)
+library(reporttools)
+
+# shortcut to formatPval
+pval <- function(...)formatPval(..., includeEquality=TRUE)
 
@@
 .gambase <- 18
+.maxSD <- 0.03
 
@@
 rangeLAI <- round(range(dfa$LAI),1)
+
+# LAI ~ BA
+balm <- lm(LAI.mean ~ BA, data=ba)
 
@@
 
+# Canopy photo - Ltau comparison
+lmcanphot <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)
+
+# With optimized k
+flatcan_byring2 <- agg_flatcan(flatcan, by="Ring", k=kopt)
+flatcan_byring2 <- add_PARLAI_to_flatcan(facegap_cloudy_byring,flatcan_byring2)
+lmcanphot2 <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring2)
+
 ```
@@
 \
-$\mathrm{Remko~A.~Duursma}^1$, $\mathrm{Teresa~E.~Gimeno}^{1,2}$, $\mathrm{Matthias~M.~Boer}^1$, $\mathrm{Kristine~Y.~Crous}^1$, $\mathrm{Mark~G.~Tjoelker}^1$, $\mathrm{David~S.~Ellsworth}^1$
+Remko A. Duursma^1^, Teresa E. Gimeno^1,2^, Matthias M. Boer^1^, Kristine Y. Crous^1^, Mark G. Tjoelker^1^, David S. Ellsworth^1^.
 
 \
-$~^1$ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia
+^1^ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia
 
 \
-$~^2$ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France
+^2^ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France
 
@@
 Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
-Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [$CO_2$] and water availability of unmanaged forests have been poorly quantified.
-We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air $CO_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
-We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [$CO_2$].
+Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of unmanaged forests have been poorly quantified.
+We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
+We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [CO~2~].
 We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production.
-Contrary to expectation, *L* did not respond to elevated [$CO_2$] in this timeframe.
+Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe.
 We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear annual pattern.
@@
 Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
-Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly increased under elevated [$CO_2$] (p = 0.051).
+Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly increased under elevated [CO~2~] (p `r pval(LLtest$p.value)`).
 We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.
@@
 *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that *L* increases in response to elevated atmospheric [$CO_2$] ($eC_a$) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-Also expected because increased quantum yield of photosynthesis in $eC_a$ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to $eC_a$ in mature forests [@Norby2011].
-Quantifying the magnitude and direction of the $eC_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in $C_a$ may provide an important feedback to climate change, and modify direct physiological responses to $C_a$ [@Tor-ngern2015; @Murray2012].
+Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO~2~] (eC~a~) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+Also expected because increased quantum yield of photosynthesis in eC~a~ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
+Quantifying the magnitude and direction of the eC~a~ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
 
 \
-Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
-Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) [@Norby2011].
-Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*.
-Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* [@Norby2011], although positive responses to $eC_a$ in terms of productivity were observed at all six sites.
-Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$.
-Because diminished returns in increasing *L* in response to $eC_a$ when LAI is larger [@Norby2005].
+Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
+Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011].
+Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
+Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites.
+Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
+Because diminished returns in increasing *L* in response to eC~a~ when LAI is larger [@Norby2005].
 
 \
-In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
+In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
 These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance.
 **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al.
-[-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$.
-Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.
+[-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~.
+Based on the same water-savings principle, we predict that *L* should also increase under eC~a~ in water-limited forests.
 
@@
 \
-We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment.
+We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to eC~a~, the EucFACE experiment.
 The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
-We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$.
-In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to $eC_a$ [@Norby2011].
+We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~.
+In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [@Norby2011].
 Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
@@
 ## Site description and experimental design
-The EucFACE experiment is located on an ancient alluvial ﬂoodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l.
+The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l.
 elevation (33°37'S, 150°44'E).
@@
 Groundwater depth is 12.6 – 12.9 m below the soil surface.
-The woodland ranges in density from 600-1000 trees $ha^{-1}$.
-The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E.
+The woodland ranges in density from 600-1000 trees ha^-1^.
+The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E.
 amplifolia* Naudin.
@@
 Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high.
-Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
-The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots.
-The target diurnal $C_a$ for the rings exposed to $eC_a$ was 150 μmol $CO_2$ mol$^{-1}$ above ambient $C_a$ in full operation mode.
-However, $C_a$ was increased gradually for six months at a rate of ca.
-30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Fig.
+Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
+The vegetation within three randomly selected rings was exposed to a CO~2~ mole fraction (C~a~) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots.
+The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode.
+However, C~a~ was increased gradually for six months at a rate of ca.
+30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode was reached (Fig.
 1a).
@@
 ## Leaf litter production
-In each ring, 8 litter cloth baskets with an area of 0.1979 $m^2$ were installed at 0.5m above the forest floor.
+In each ring, 8 litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor.
 Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc.
-Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14).
-We did not find a significant effect of $C_a$ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).
+Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
+We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).
 
@@
 
-where $\tau_d$ is diffuse canopy transmittance, Qabove is above canopy PAR (µmol m$^{-2}s^{-1}$), and Q$_{below}$ is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
-We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was  > 0.03 (in units of $\tau_d$).
+where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR (µmol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
+We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was > `r .maxSD` (in units of $\tau_d$).
 
@@
 where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000].
-Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 [@Stenberg2006].
+Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006].
 We solved numerically for *L* in Eq.
@@
 For comparison between *L* from photos and PAR transmittance, we used simple linear regression.
-To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.
+To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as a factor.
 
 \
-To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
-The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the $eC_a$ treatment across the FACE rings (Fig.
+To assess effects of eC~a~ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
+The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig.
 3).
-Because the EucFACE woodland is well established, the differences in BA are not due to $C_a$ treatment effects, but instead represent natural variation within the woodland.
+Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.
 We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig.
@@
 \
-We found no differences in the *L* anomaly between the $C_a$ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig.
+We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig.
 3).
 A brief exception was the period where one of the rings was affected by an insect outbreak (see above).
-A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (p = `r LanomAOV[[3]][1]`).
-What difference between $C_a$ treatments could we have reasonably detected?
+A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the *L* anomaly (p `r pval(LanomAOV[[3]][1])`).
+What difference between C~a~ treatments could we have reasonably detected?
 The average half-width of the confidence interval on the *L* anomaly (Fig.
-4) was `r LanomCIamb` $m^2$ $m^{-2}$ for $aC_a$, and `r LanomCIele` $m^2$ $m^{-2}$ for the $eC_a$ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` $m^2$ $m^{-2}$).
+4) was `r LanomCIamb` m^2^ m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment.
+This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` m^2^m^-2^).
 
@@
 5, and demonstrate the irregular temporal pattern in leaf production and $P_L$.
-We found no significant difference in either litter or leaf production between $C_a$ treatments (p > 0.2 for both, assessed with linear-mixed effects models).
+We found no significant difference in either litter or leaf production between C~a~ treatments (p > 0.2 for both, assessed with linear-mixed effects models).
 
@@
 If leaf lifespan is constant, $P_L$ would be linearly related to $L_\tau$.
-Instead, we found that in $eC_a$, $P_L$ was lower at a given $L_\tau$ as compared to $aC_a$.
+Instead, we found that in eC~a~, $P_L$ was lower at a given $L_\tau$ as compared to aC~a~.
 We estimated mean leaf lifespan from Fig.
 6a as the ratio of $L_\tau$ to annual $P_L$.
-We found a significant (p = `r LLtest$p.value`) increase in LL with elevated $C_a$ (Fig.
-6b), with LL = `r LLtest$estimate1`yr in $aC_a$ and `r LLtest$estimate2`yr in $eC_a$.
+We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig.
+6b), with LL = `r LLtest$estimate1`yr in aC~a~ and `r LLtest$estimate2`yr in eC~a~.
 
@@
 
-## Response of *L* to elevated [$CO_2$]
-We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$.
-We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability.
-We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands [@Norby2011].
+## Response of *L* to elevated [CO~2~]
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~.
+We had predicted that *L* would increase in response to eC~a~ as a result of increased quantum yield for photosynthesis, enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability.
+We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011].
 Counter to our expectations but similar to Norby et al.
-[-@Norby2003], we found no significant effect of elevated $C_a$ on *L*.
-It could be argued that the change in *L* in response to $eC_a$ was too small to be detected.
+[-@Norby2003], we found no significant effect of elevated C~a~ on *L*.
+It could be argued that the change in *L* in response to eC~a~ was too small to be detected.
 Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
@@
 Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*.
-Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.
+Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.
 
 \
-This lack of a response to elevated $C_a$ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to $eC_a$.
+This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons.
+First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~.
 They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
 The mean *L* at the EucFACE was ca.
-`r mean(ba$LAI.mean)` $m^2$ $m^{-2}$, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
+`r mean(ba$LAI.mean)` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
 [-@Iio2013].
-Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under $eC_a$.
+Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~.
 It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE.
@@
 \
-A possible explanation for the lack of a *L* response to $eC_a$ is that leaf growth may not be directly related to increased carbon availability as a result of $eC_a$.
-In other words, stimulation of net leaf photosynthesis by $eC_a$ (a common response, also observed at the EucFACE, Gimeno et al.
+A possible explanation for the lack of a *L* response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~.
+In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
 in review) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.
-Either process may itself be affected by $eC_a$ [@DeKauwe2014].
-We found a significant effect of $eC_a$ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated $C_a$ (Fig.
+Either process may itself be affected by eC~a~ [@DeKauwe2014].
+We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig.
 6).
-Because we found no difference in *L* with $C_a$ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
-To further understand the response of biomass allocation to $eC_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
+Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
+To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
 For example, Drake et al.
-(in review) reported increased soil respiration under $eC_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
+(in review) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
 
@@
 Simulations from De Kauwe et al.
-[-@DeKauwe2014] at ambient and elevated $C_a$ also showed little response to inter-annual variation in rainfall (pers.
+[-@DeKauwe2014] at ambient and elevated C~a~ also showed little response to inter-annual variation in rainfall (pers.
 comm.
@@
 @Kala2013), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
-Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as $eC_a$ (Fig.
+Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as eC~a~ (Fig.
 2).
@@
 # Conclusions
-Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [$CO_2$] after 27 months of treatment and 24 months at +150ppm $CO_2$ elevation.
+Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [CO~2~] after 27 months of treatment and 24 months at +150ppm CO~2~ elevation.
 Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage).
-Leaf lifespan, estimated from mean *L* and annual litter production, was `r round(abs(LLtest$estimate*12),1)` months longer in the elevated [$CO_2$] treatment.
+Leaf lifespan, estimated from mean *L* and annual litter production, was `r round(abs(LLtest$estimate*12),1)` months longer in the elevated [CO~2~] treatment.
 
@@
 
+# Figure captions {.unnumbered}
+
+**Figure 1**.
+(a) Target [CO~2~] (C~a~) relative to the ambient C~a~ treatment at the EucFACE.
+The elevated C~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in Feb.
+2013.
+(b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions ($F_diff$ > 0.98) only.
+The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model).
+Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~ (+150 ppm).
+Note steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
+The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate *L* estimates to litter production (see Methods).
+
+**Figure 2**.
+Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA).
+This relationship was used to calculate the *L* anomaly in Fig.
+3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.
+
+**Figure 3**.Time series of *L* after correcting for plot-to-plot differences in basal area (BA).
+At every time point, the *L* anomaly was calculated as $L_\tau$ minus *L* fitted from the relationship with BA shown in Fig.
+2.
+Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect.
+Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05, except for a brief period in late 2014.
+This difference was due to an insect attack that led to partial defoliation of the canopy in one FACE ring (see Fig.
+1).
+
+**Figure 4**.
+Comparison of absolute changes in *L* ($\Delta L$) to litter production (in units of *L*), calculated for each litter collection period (approx.
+monthly).
+To calculate $\Delta L$, we fit a generalized additive model fitted to data from each FACE ring separately, and estimated the change in L during the litter collection period (and expressed as monthly changes, by accounting for the differences in actual lengths of litter collection periods).
+Green symbols indicate periods during which $\Delta L$ was positive (net leaf production), orange symbols when $\Delta L$ was negative (net leaf loss).
+Solid black lines are 1:1 lines, coloured lines regression lines with 95% confidence intervals.
+The strong positive correlation between litter production and $\Delta L$ during periods of net leaf production indicates that during leaf flushing, old leaves are actively shed.
+
+**Figure 5**.
+Timeseries of litter production and estimated gross leaf production rate.
+Leaf production rate was estimated as the sum of the net change in *L* and litter production (converted to a per area basis using specific leaf area estimates, see Methods).
+Error bars are omitted for clarity.
+We did not find any differences in litter production or leaf production between C~a~ treatments (see Results).
+
+**Figure 6**.
+(a) Total litter production (converted from data in Fig.
+5 to an annual estimate) for each of the FACE rings plotted against mean $L_\tau$ during the entire litter collection period.
+The ratio of mean $L_\tau$ to annual litter production gives the mean leaf life span.
+Grey lines indicate leaf lifespan isolines.
+(b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p = `r pval(LLtest$p.value)`).
+
+**Figure 7**.
+Dynamics of *L* in relation to environmental drivers.
+(a) Smoothed time series of *L_\tau* averaged across all six FACE rings.
+(b) From panel (a), we calculated the daily change in *L* (dL/dt), including the 95% confidence interval for the derivative of the time series.
+Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05).
+Four of the five significant positive increases corresponded to approximate dates of commencement of leaf flushing (horizontal lines) observed in the field (except the third brief significant increase).
+(c) Soil water content estimated from over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).
+(d) Average daily air temperature with a fitted smoother.
+
+**Figure S1**.
+Increase in gapfraction ($\tau_d$) and total litter production during the 2013 drought, calculated for each ring separately.
+Blue colours are ambient, red colours eleveated C~a~.
+(a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
+(b) Mean litter production for each ring.
+The confidence interval was calculated based on he 8 litter traps in each ring.
+Note the overlapping confidence intervals for all rings.
+
+**Figure S2**.
+Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
+Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
+(a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
+[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]`).
+
+**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions ($F_diff$ > 0.98), or using all PAR data without filtering.
+Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals).
+Note the much lower $\tau$ during southern hemisphere winter when using all data compared to diffuse only.
+This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.
+
 # References {.unnumbered}

### 2015-05-04 R/figures.R

@@ -382,18 +382,37 @@ figureSI1 <- function(litring, df){
   
 }  
 
+
 figureSI2 <- function(df){
   
-  par(mar=c(5,5,2,2), cex.axis=0.9)
+  par(mfrow=c(1,2), mar=c(5,5,2,2), cex.axis=0.9)
+  plotit <- function(df){
+    with(df, plot(LAI, LAI.PAR.mean, 
+                  ylab=expression(LAI~from~tau[d]~~(m^2~m^-2)),
+                  xlab=expression(LAI~from~canopy~photos~~(m^2~m^-2)),
+                  pch=19, col=my_ringcols()[Ring],
+                  panel.first=predline(lm(LAI.PAR.mean ~ LAI, data=df), lty=5),
+                  xlim=c(0.8,2.2), ylim=c(0.8,2.2)))
+    abline(0,1)
+    
+    ds <- split(df, df$Ring)
+    for(i in 1:6)predline(lm(LAI.PAR.mean ~ LAI, data=ds[[i]]), col=my_ringcols()[i], poly=FALSE)
+
+    box()
+  }
   
-  with(df, plot(LAI, LAI.PAR.mean, 
-                ylab=expression(LAI~from~tau[d]~~(m^2~m^-2)),
-                xlab=expression(LAI~from~canopy~photos~~(m^2~m^-2)),
-                pch=19, col=my_co2cols()[treatment],
-                xlim=c(0.8,2), ylim=c(0.8,2)))
-  abline(0,1)
-  predline(lm(LAI.PAR.mean ~ LAI, data=df), lty=5)
-  box()
+  plotit(df)
+  plotlabel("(a)", "topleft")
+  
+  legend("bottomright", as.character(1:6), lty=1, bty='n', pch=19,
+         col=my_ringcols(), title="Ring", cex=0.8, lwd=1)
+  
+  # Panel b : optimized
+  df <- agg_flatcan(flatcan, by="Ring", k=kopt)
+  df <- add_PARLAI_to_flatcan(facegap_cloudy_byring,df)
+  
+  plotit(df)
+  plotlabel("(b)", "topleft")
   
 }
 

### 2015-05-04 R/functions-figures.R

@@ -272,7 +272,8 @@ addpoly <- function(x,y1,y2,col=alpha("lightgrey",0.8),...){
   x <- x[ii]
   polygon(c(x,rev(x)), c(y1, rev(y2)), col=col, border=NA,...)
 }
-predline <- function(fit, from=NULL, to=NULL, ...){
+
+predline <- function(fit, from=NULL, to=NULL, poly=TRUE, ...){
   
   if(is.null(from))from <- min(fit$model[,2], na.rm=TRUE)
   if(is.null(to))to <- max(fit$model[,2], na.rm=TRUE)
@@ -282,9 +283,10 @@ predline <- function(fit, from=NULL, to=NULL, ...){
   nm <- names(coef(fit))
   names(newdat)[1] <- nm[length(nm)]
   
-  pred <- as.data.frame(predict(fit, newdat, se.fit=TRUE, interval="confidence")$fit)
-  
-  addpoly(newdat[[1]], pred$lwr, pred$upr)
+  if(poly){
+    pred <- as.data.frame(predict(fit, newdat, se.fit=TRUE, interval="confidence")$fit)
+    addpoly(newdat[[1]], pred$lwr, pred$upr)
+  }
   ablinepiece(fit, from=from, to=to, ...)
   
 }

### 2015-05-04 make_figures.R

@@ -5,6 +5,7 @@ source("R/derivSimulCI.R")
 
 
 maxdate <- as.Date("2015-3-1")
+# maxdate <- as.Date(today())
 
 to.pdf(figure1(subset(facegap_cloudy_byring,Date < maxdate), ramp),
        filename="output/figures/Figure1.pdf",
@@ -43,7 +44,7 @@ to.pdf(figureSI1(litring, subset(facegap_cloudy_byring, Date < maxdate)),
 
 to.pdf(figureSI2(flatcan_byring),
        filename="output/figures/FigureSI2.pdf",
-       width=5, height=5)
+       width=8, height=4)
 
 to.pdf(figureSI3(subset(facegap_cloudy_byring, Date < maxdate), subset(facegap_all_byring, Date < maxdate)),
        filename="output/figures/FigureSI3.pdf",

### 2015-05-04 run.R

@@ -80,6 +80,14 @@ flatcan_byring <- add_PARLAI_to_flatcan(facegap_cloudy_byring,flatcan_byring)
 
 flatcan_byCO2 <- agg_flatcan(flatcan, by="CO2")
 
+# find optimal k for flat canopy photos
+Ok <- function(k){
+  flatcan_byring <- agg_flatcan(flatcan, by="Ring", k)
+  flatcan_byring <- add_PARLAI_to_flatcan(facegap_cloudy_byring,flatcan_byring)
+  with(flatcan_byring, sum((LAI - LAI.PAR.mean)^2))
+}
+kopt <- optimize(Ok, c(0.2, 0.8))$minimum
+
 # Dataset with litter fall comparison to changes in LAI during same period.
 dLAIlitter <- make_dLAI_litter(facegap_cloudy_byring, litter, kgam=20)
 
@@ -101,8 +109,11 @@ m1 <- paste0("Most recent day with usable cloudy data: ",max(facegap_cloudy_byri
 message(m1)
 
 
+# Make manuscript
+render("manuscript.Rmd", "word_document", "manuscript.docx")
 
-
+# Make md like this:
+# render("manuscript.Rmd", md_document())
 
 
 

### 2015-05-04 manuscript.Rmd
More refs; bug fixes; author info moved to top

@@
 ---
+
+\
+Remko A. Duursma^1^, Teresa E. Gimeno^1,2^, Matthias M. Boer^1^, Kristine Y. Crous^1^, Mark G. Tjoelker^1^, David S. Ellsworth^1^.
+
+\
+^1^ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia
+
+\
+^2^ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France
 
@@
 
-\
-Remko A. Duursma^1^, Teresa E. Gimeno^1,2^, Matthias M. Boer^1^, Kristine Y. Crous^1^, Mark G. Tjoelker^1^, David S. Ellsworth^1^.
-
-\
-^1^ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia
-
-\
-^2^ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France
-
 # Abstract {.unnumbered}
@@
 
-Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
+Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
 *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO~2~] (eC~a~) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-Also expected because increased quantum yield of photosynthesis in eC~a~ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
-Quantifying the magnitude and direction of the eC~a~ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
+Ecosystem models generally predict that *L* increases in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+Other mechanisms are likely involved as well, including increased quantum yield of photosynthesis in eC~a~ allowing leaf retention into deeper shade [@Norby1999], and interactions with improved water relations under eC~a~  [@Wullschleger2002].
+Quantifying the magnitude and direction of the effect of elevated atmospheric [CO~2~] (eC~a~) on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
+To date, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
 
@@
 Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
-Because diminished returns in increasing *L* in response to eC~a~ when LAI is larger [@Norby2005].
 
@@
 In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
-These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance.
-**Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al.
+These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, since we observe a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008].
+In support of this prediction, Donohue et al.
 [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~.
-Based on the same water-savings principle, we predict that *L* should also increase under eC~a~ in water-limited forests.
+Based on the same water-savings principle, we predict that *L* should increase under eC~a~ in water-limited forests.
 
 \
-Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g.
-[@Richardson2012; @Savoy2015]) whereas *L* responses to fluctuations in water availability are qualitatively well-known (**more cites**), but poorly quantified.
-The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009].The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
+Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g.
+@Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
+The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
 However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
@@
 \
-We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to eC~a~, the EucFACE experiment.
-The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
+We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated woodland exposed to eC~a~, the EucFACE experiment.
+The woodland at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
 We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~.
@@
 All PAR data were averaged in 30 min time steps.
-At a nearby site (within 2 km), we measured the diffuse fraction (Fdiff) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy.
-For determination of *L*, we used only highly diffuse conditions, where Fdiff > 0.98.
+At a nearby site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy.
+For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
 We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles.
@@
 
-(2) $$\tau_d = \bar{Q}_{below} / Q_{above}$$
+\
+(1) 
+$$\tau_d = \bar{Q}_{below} / Q_{above}$$
 
@@
 
-(3) $$\tau_d = 2\int_0^{\pi/2} e^{-LG(\theta)/cos(\theta)}sin(\theta)cos(\theta)d\theta$$
+\
+(2)
+$$\tau_d = 2\int_0^{\pi/2} e^{-LG(\theta)/cos(\theta)}sin(\theta)cos(\theta)d\theta$$
 
@@
 We solved numerically for *L* in Eq.
-3, using a smooth integration over $\theta$, yielding $L_\tau$.
+2, using a smooth integration over $\theta$, yielding $L_\tau$.
 
@@
 To overcome violations in the assumptions underlying Eq.
-3, we calibrated *L* against litter production.
-We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig.
+2, we calibrated *L_\tau* against cumulative litter production during a drought period in 2013 (indicated in Fig.
 1).
-We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
+During this period (14 July 2013 – 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
 1).
-During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period).
+During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm).
 A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
-In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance.
+Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance.
 We calibrated Eq.
-4,
+3,
 
-(4) $$c(L_{\tau,t} - L_{\tau,t=0}) = \int_0^t P_Ldt$$
+\
+(3) 
+$$c(L_{\tau,t} - L_{\tau,t=0}) = \int_0^t P_Ldt$$
 
@@
 It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq.
-4.
+3.
+We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought (see Fig.
+S1).
 
@@
 *L* from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq.
-1).
+4).
 
-(1) $$L_p = -log(\tau_p)/k$$
+\
+(4) 
+$$L_p = -log(\tau_p)/k$$
 
-where $\tau_p$ canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following Macfarlane et al.
+where $\tau_p$ canopy transmission estimated from canopy photos.
+We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al.
 [-@Macfarlane2007] recommendation for eucalypt forests.
+Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig.
+SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`).
+
+, and compared this assumption with an optimized extinction that minimized the difference between the two estimates of *L*.
 
@@
 We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$).
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig.
-SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`).
 This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
@@
 # Results
-
 Canopy transmittance ($\tau_d$) varied between `r min(df$Gapfraction.mean)` and `r max(df$Gapfraction.mean)`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
@@
 4).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
-During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
+During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
 5).
@@
 4 and 6), when the soil water content in the top 80 cm increased to ca.
-> 0.2 $m^3$ $m^{-3}$.
+> 0.2 m^3^ m^-3^.
 Two periods of sustained decrease in *L* (red lines in Fig.
@@
 ## Response of *L* to elevated [CO~2~]
+
+** tie to objectives at end of intro! **
+
 We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~.
@@
 Other studies in FACE are consistent with longer leaf-lifespan in elevated Ca shown here.
+[@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
+Although this indicates the canopy stayed greener longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ due to overall higher productivity, leading to similar *L* in ambient C~a~ despite higher productivity.
+But not with [@Warren2011]: increased senescence in eC~a~ in response to a combined drought and heat wave.
 
@@
 2013.
-(b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions ($F_diff$ > 0.98) only.
+(b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions (F~diff~ > 0.98) only.
 The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model).
@@
 **Figure 2**.
-Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA).
+Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA).
 This relationship was used to calculate the *L* anomaly in Fig.
@@
 Grey lines indicate leaf lifespan isolines.
-(b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p = `r pval(LLtest$p.value)`).
+(b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p `r pval(LLtest$p.value)`).
 
@@
 Dynamics of *L* in relation to environmental drivers.
-(a) Smoothed time series of *L_\tau* averaged across all six FACE rings.
+(a) Smoothed time series of $L_\tau$ averaged across all six FACE rings.
 (b) From panel (a), we calculated the daily change in *L* (dL/dt), including the 95% confidence interval for the derivative of the time series.
@@
 (a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
-[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]`).
+[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
 
-**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions ($F_diff$ > 0.98), or using all PAR data without filtering.
+**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
 Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals).

### 2015-05-04 R/functions-figures.R
removed dependency on plantecophys and plotBy

@@ -79,6 +79,41 @@ adderrorbars <- function(x,y,SE,direction,barlen=0.04,...){
   
 }
 
+#'@title Add a line to a plot
+#'@description As \code{abline}, but with \code{from} and \code{to} arguments. 
+#'If a fitted linear regression model is used as asn argument, it uses the min and max values of the data used to fit the model.
+#'@param a Intercept (optional)
+#'@param b Slope (optional)
+#'@param reg A fitted linear regression model (output of \code{\link{lm}}).
+#'@param from Draw from this X value
+#'@param to Draw to this x value
+#'@param \dots Further parameters passed to \code{\link{segments}}
+#'@export
+ablinepiece <- function(a=NULL,b=NULL,reg=NULL,from=NULL,to=NULL,...){
+  
+  # Borrowed from abline
+  if (!is.null(reg)) a <- reg
+  
+  if (!is.null(a) && is.list(a)) {
+    temp <- as.vector(coefficients(a))
+    from <- min(a$model[,2], na.rm=TRUE)
+    to <- max(a$model[,2], na.rm=TRUE)
+    
+    if (length(temp) == 1) {
+      a <- 0
+      b <- temp
+    }
+    else {
+      a <- temp[1]
+      b <- temp[2]
+    }
+  }
+  
+  segments(x0=from,x1=to,
+           y0=a+from*b,y1=a+to*b,...)
+  
+}
+
 
 axisTitle <- function(text, side, ...){
   

### 2015-05-04 load.R

@@ -2,8 +2,6 @@
 library(plyr)
 library(dplyr)
 library(HIEv)
-library(doBy)
-library(plantecophys)
 library(canopyphoto)
 library(plotrix)
 library(RColorBrewer)
@@ -13,6 +11,8 @@ library(Hmisc)
 library(gplots)
 library(mgcv)
 library(stringr)
+
+library(rmarkdown)
 
 
 if(packageVersion('HIEv') < '0.6.4')

### 2015-05-04 manuscript.Rmd
abstract edits

@@
 
-Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
+Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
 Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of unmanaged forests have been poorly quantified.
 We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
-We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [CO~2~].
+We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, *L* should increase with elevated [CO~2~].
 We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production.
 Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe.
-We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear annual pattern.
+We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear annual pattern.
 Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady during periods of low rainfall.
 Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
-Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly increased under elevated [CO~2~] (p `r pval(LLtest$p.value)`).
-We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.
+Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly longer under elevated [CO~2~] (p `r pval(LLtest$p.value)`).
+Our main finding that *L* was not responsive to elevated CO~2~ are consistent with other forest FACE studies, but contrast the positive response commonly predicted by ecosystem models.
 

### 2015-05-07 R/data_processing.R
trimming package dependencies.

@@ -76,7 +76,7 @@ add_PARLAI_to_flatcan <- function(df1, df2){
     x2[[i]]$LAI.PAR.mean <- ap$y
   }
   
-  df2 <- rbind.fill(x2)
+  df2 <- bind_rows(x2)
 return(df2)
 }
 

### 2015-05-07 load.R

@@ -1,12 +1,12 @@
 
-library(plyr)
+# library(plyr)
 library(dplyr)
 library(HIEv)
 library(doBy)
 library(plantecophys)
-library(canopyphoto)
-library(plotrix)
-library(RColorBrewer)
+# library(canopyphoto)
+# library(plotrix)
+# library(RColorBrewer)
 library(plotBy)
 library(scales)
 library(Hmisc)

### 2015-05-07 load.R
trimmed package dependencies

@@ -1,20 +1,13 @@
 
-# library(plyr)
-library(dplyr)
-library(HIEv)
-library(canopyphoto)
-library(plotrix)
-library(RColorBrewer)
-library(doBy)
-library(plantecophys)
-library(plotBy)
-library(scales)
-library(Hmisc)
-library(gplots)
-library(mgcv)
-library(stringr)
 
-library(rmarkdown)
+library(dplyr)    # summarize, group_by
+library(HIEv)
+library(doBy)     # summaryBy
+library(gplots)   # plotCI
+library(mgcv)     # gam
+library(stringr)  # str_trim
+
+library(rmarkdown) # render
 
 
 if(packageVersion('HIEv') < '0.6.4')

### 2015-05-08 manuscript.Rmd
Manuscript updates; Bug fix (Hmisc needed after all); airt includes more met data.

@@
 *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that *L* increases in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-Other mechanisms are likely involved as well, including increased quantum yield of photosynthesis in eC~a~ allowing leaf retention into deeper shade [@Norby1999], and interactions with improved water relations under eC~a~  [@Wullschleger2002].
-Quantifying the magnitude and direction of the effect of elevated atmospheric [CO~2~] (eC~a~) on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
-To date, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
+Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+This potential change in *L* is of central importance to net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
+Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
 
@@
 Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
-Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites.
+Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites [@Norby2005].
 Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
@@
 In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
-These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, since we observe a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008].
+These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008].
 In support of this prediction, Donohue et al.
@@
 The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
-However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
-Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].
+However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
+Understanding dynamics of *L* in response to water availability and its potential interactions with C~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
 
@@
 We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~.
-In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [@Norby2011].
+In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [cf.
+@Norby2011].
 Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
@@
 The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l.
-elevation (33°37'S, 150°44'E).
-It is located within a remnant patch of native Cumberland Plain woodland (Tozer, 2003).
+elevation (33^o^ 37'S, 150^o^ 44'E).
+It is located within a remnant patch  of native Cumberland Plain woodland [@Tozer2003].
 The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca.
 70 years.
-The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 °C and mean annual precipitation of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
+The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
+The potential evapotranspiration (PET) is ca.
+1500mm (*to be refined!*), giving a ratio of P/PET of ca.
+0.53 (*quite dry - details to be added*).
 
 \
-The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil.
+The soil at the site is a loamy sand in the Clarendon Formation [@Barton2010; @Isbell2002], a slightly-acid, low-organic C, well-drained soil.
 The soil surface is a loamy sand with > 75% sand content, and from ca.
@@
 Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig.
-SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`).
-
-, and compared this assumption with an optimized extinction that minimized the difference between the two estimates of *L*.
-
-\
-We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$).
-This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
+SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig.
+S2).
+We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig.
+S2), giving *k* = `r kopt`.
+These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
 3), weighted towards the zenith.
@@
 ## Response of *L* to elevated [CO~2~]
-
-** tie to objectives at end of intro! **
-
-We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~.
-We had predicted that *L* would increase in response to eC~a~ as a result of increased quantum yield for photosynthesis, enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability.
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~.
+We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability.
 We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011].
@@
 Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~.
-It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE.
-Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands [@White2010].
+It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because compare to other forest FACE sites, the EucFACE is very strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
 
@@
 In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
-in review) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
+*in review*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.
@@
 For example, Drake et al.
-(in review) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
+(*in review*) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
 
 \
-Other studies in FACE are consistent with longer leaf-lifespan in elevated Ca shown here.
-[@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
-Although this indicates the canopy stayed greener longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ due to overall higher productivity, leading to similar *L* in ambient C~a~ despite higher productivity.
-But not with [@Warren2011]: increased senescence in eC~a~ in response to a combined drought and heat wave.
+Despite a lack of response of canopy *L*, we did find a significantly longer leaf lifespan (LL) in eC~a~, consistent with our second hypothesis.
+We could not determine whether there was an interaction with water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
+6).
+Although LL was increased in eC~a~, this was not reflected in a difference in leaf litter production at any time point (Fig.
+5), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
+There is little evidence from other FACE sites that LL is affected by eC~a~.
+Taylor et al.
+[-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
+Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~, leading to similar *L* in ambient C~a~ despite higher productivity.
+In contrast, Warren et al.
+[-@Warren2011] found increased senescence in eC~a~ at the ORNL FACE site (*Liquidambar styraciflua* plantation), in response to a combined drought and heat wave.
 
@@
 Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
-Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan is unknown.
-We can also not rely on estimated leaf area based on allometric relationships with stem diameter or height, since these equations are not available for mature *E.
-tereticornis*, and besides these would be useless since *L* is highly variable temporally (Figs.
-1, 4).
-We found that indirect estimates of *L* based on canopy transmittance were very reliable.
-Estimated *L* compared well to that from flat canopy photos (cf.
-[@Macfarlane2007] (Fig.
-SI2), and after calibration, to leaf litter production during a dry period (Fig.
-4).
-Unlike canopy photos, PAR sensors do not need thresholding or time-consuming post-processing.
+Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan (LL) is unknown (instead, we used *L* and litter production to estimate LL).
+We can also not rely on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
+tereticornis*, and would be useless anyway since *L* is highly variable temporally (Figs.
+1, 7).
+We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
+1, 5, 7).
 We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999].
@@
 
+We used the opportunity provided by the high quality timeseries of *L* to test a simple photography-based method suggested by Macfarlane et al.
+[-@Macfarlane2007].
+The two methods compared quite well (R^2^ = `r glance(lmcanphot)$r.squared`) (Fig.
+S2), suggesting the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
+
 ## Litter production and dynamics of *L*
-We found a correlation between *L* dynamics and litter production, both during periods of decreasing and increasing *L*.
-During a four-month period in 2013 without significant rainfall, we used the decrease in *L* as an opportunity to calibrate against litter production.
-During periods of increasing *L*, litter fall increased sharply, indicating replacement of old leaves by young leaves.
-Indeed we observed in the field that after the significant flushing event of Dec.
-2013, very few old leaves were present in the canopy.
-This is in line with [@Pook1984] in *Corymbia maculata* (a genus closely related to *Eucalyptus*), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
-Similarly, [@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
+We found a correlation between *L* dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing *L* (Fig.
+4).
+As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production.
+During periods of increasing *L*, litter fall increased sharply (Figs.
+4 and 5), indicating replacement of old leaves by young leaves.
+Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of Dec.
+2013.
+This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
+Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
 Since leaf flushing has been shown to be closely related to soil water availability [Fig.
@@
 5).
-The consequence is that leaf turnover is highly variable in time.
+The consequence is that leaf turnover is highly variable in time (Fig.
+5).
 
@@
 Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth.
-Similarly, Pook  [-@Pook1984a] demonstrated leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods.
+Similarly, Pook  [-@Pook1984a] demonstrated occurrence of leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods.
 Thus a minimum temperature is also required for leaf growth to occur.
@@
 In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g.
-@Kala2013), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
-Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as eC~a~ (Fig.
-2).
-We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests, because not all important mechanisms are incorporated in these models.
-In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability.
-Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and rates of leaf expansion [@Muller2011].
+@Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
+Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
+We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests.
+In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability, as we have demonstrated that both processes are highly dynamic.
+Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and rates of leaf expansion [e.g.
+@Muller2011].
 Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g.
-G’DAY, see Corbeels et al.
+G'DAY, see Corbeels et al.
 2005), but this process has been poorly quantified.
@@
 # Conclusions
-Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [CO~2~] after 27 months of treatment and 24 months at +150ppm CO~2~ elevation.
+Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [CO~2~] after `r nummon` months of treatment and `r nummon-3` months at +150ppm CO~2~ elevation.
 Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage).
@@
 
+\
 We found no clear annual pattern in dynamics of *L*, instead, a net increase in *L*, indicating leaf growth, consistently commenced after heavy rainfall following a dry period.
 The complex dynamics of litter production and the rapid response of litter production and leaf growth to soil water availability challenge current-generation prognostic *L* and phenology models.
-High resolution time series such as the one we present in this study can be used to develop new models of phenology for evergreen woodlands.
+High resolution time series such as the one we present in this study can be used as a basis to develop new models of phenology for evergreen woodlands.
 

### 2015-05-08 R/data_processing.R

@@ -394,7 +394,12 @@ get_rosTair <- function(){
   d <- downloadTOA5(c("ROS_WS","Table05min"), maxnfiles=500)
   
   d <- as.data.frame(dplyr::summarize(group_by(d, Date),
-                                      Tair=mean(AirTC_Avg, na.rm=TRUE)
+                                      Tair=mean(AirTC_Avg, na.rm=TRUE),
+                                      Tmin=min(AirTC_Avg, na.rm=TRUE),
+                                      Tmax=max(AirTC_Avg, na.rm=TRUE),
+                                      RH=mean(RH, na.rm=TRUE),
+                                      RHmin=min(RH, na.rm=TRUE),
+                                      RHmax=max(RH, na.rm=TRUE)
                                       ))
 return(d)  
 }

### 2015-05-08 R/figures.R

@@ -264,9 +264,7 @@ figure7 <- function(df, facesoilwater, faceraindaily, airt){
   axis(2)
   box()
   plotlabel("(b)","topleft", inset.x=xin)
-  
-#   arrows(x0=flushingdates(), x1=flushingdates(), y0=0.015, y1=0.01,
-#          length=0.05)
+
   segments(x0=flushingdates()-10, x1=flushingdates()+7,
            y0=0.014, y1=0.014, lwd=2) 
   
@@ -293,7 +291,6 @@ figure7 <- function(df, facesoilwater, faceraindaily, airt){
   par(mar=c(5,7,0,6))
   smoothplot(Date, Tair, data=subset(airt, Date > xl[1]), 
              kgam=25, pointcols=alpha("grey",0.8), linecols="black",axes=FALSE,
-#              panel.first=abline(h=seq(0,30,by=2), col="grey", lty=5),
              ylab=expression(T[air]~~(degree*C)), 
              xlab="",
              ylim=c(0,30), xlim=xl)

### 2015-05-08 load.R

@@ -6,6 +6,7 @@ library(doBy)     # summaryBy
 library(gplots)   # plotCI
 library(mgcv)     # gam
 library(stringr)  # str_trim
+library(Hmisc)    # approxExtrap
 
 library(rmarkdown) # render
 

### 2015-05-08 manuscript.Rmd
data analysis description improved

@@
 library(knitr)
+
 load("cache/lai_workspace.RData")
@@
 # Number of digits in printing of numbers
-options(scipen = 1, digits = 2)
+options(scipen = 1, digits = 3)
 
@@
 In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
-These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008].
+These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
 In support of this prediction, Donohue et al.
@@
 To overcome violations in the assumptions underlying Eq.
-2, we calibrated *L_\tau* against cumulative litter production during a drought period in 2013 (indicated in Fig.
+2, we calibrated $L_\tau$ against cumulative litter production during a drought period in 2013 (indicated in Fig.
 1).
@@
 [-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig.
+Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig.
 SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig.
@@
 ## Data analysis
-We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015].
-For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics.
-In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig.
+We used generalized additive models (GAM) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015].
+A GAM is a semi-parametric fitting method that allows the estimation of smoothed trends in the data, including confidence intervals.
+In all fitted GAMs, we used a cubic regression spline.
+For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed *L* dynamics.
+Random effects were included to account for the grouping of data within FACE rings, except when we analysed average *L* across all rings (Fig.
 7).
-We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson [@Simpson2014].
+For Fig.
+7, we estimated the derivative of the fitted GAM, including confidence intervals for the derivative, following Simpson [-@Simpson2014; see also @Curtis2014].
+
+\
 For comparison between *L* from photos and PAR transmittance, we used simple linear regression.
-To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as a factor.
+To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
 
@@
 This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.
+
+\
+The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXXXX), published on figshare (doi:XXXX).
+The repository includes the processed data needed to run the analyses and figures (*raw data needs to be shared too, have to figure out how though!*).
 
@@
 
-# Figure captions {.unnumbered}
+# Figures {.unnumbered}
+
+```{r}
+options(scipen = 1, digits = 5)
+```
+
+```{r fig.width=8, fig.height=6, echo=FALSE, message=FALSE}
+figure1(subset(facegap_cloudy_byring,Date < maxdate), ramp)
+```
 
@@
 
+```{r fig.width=4, fig.height=4, echo=FALSE, message=FALSE}
+figure2(ba)
+```
+
 **Figure 2**.
@@
 3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.
+
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+figure3(subset(facegap_cloudy_byring, Date < maxdate))
+```
 
@@
 
+```{r fig.width=6, fig.height=4, echo=FALSE, message=FALSE, echo=FALSE, message=FALSE}
+figure4(dLAIlitter)
+```
+
 **Figure 4**.
@@
 
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+figure5(dLAIlitter)
+```
+
 **Figure 5**.
@@
 We did not find any differences in litter production or leaf production between C~a~ treatments (see Results).
+
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+figure6(ba)
+```
 
@@
 (b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p `r pval(LLtest$p.value)`).
+
+```{r fig.width=10, fig.height=8, echo=FALSE, message=FALSE}
+figure7(subset(facegap_cloudy_byring, Date < maxdate),
+               subset(facesoilwater, Date < maxdate),
+               subset(faceraindaily, Date < maxdate),
+               subset(airt, Date < maxdate))
+```
 
@@
 
+```{r fig.width=5, fig.height=8, echo=FALSE, message=FALSE}
+figureSI1(litring, subset(facegap_cloudy_byring, Date < maxdate))
+```
+
 **Figure S1**.
@@
 
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+figureSI2(flatcan_byring)
+```
+
 **Figure S2**.
@@
 [-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
+
+```{r fig.width=8, fig.height=5, echo=FALSE, message=FALSE}
+figureSI3(subset(facegap_cloudy_byring, Date < maxdate), 
+          subset(facegap_all_byring, Date < maxdate))
+```
 

### 2015-05-08 manuscript.Rmd
Final tidying; LAI in figures replaced with L for consistency

@@
 r <- range(df$Date)
+r[2] <- maxdate
 nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))
@@
 Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-This potential change in *L* is of central importance to net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
+This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
 Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
@@
 Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig.
-SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig.
+SI2, R^2^=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig.
 S2).
@@
 S2), giving *k* = `r kopt`.
-These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq.
+These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
 3), weighted towards the zenith.
@@
 The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXXXX), published on figshare (doi:XXXX).
-The repository includes the processed data needed to run the analyses and figures (*raw data needs to be shared too, have to figure out how though!*).
+The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, with the rmarkdown package in Rstudio, which uses Pandoc for conversion).
 
@@
 An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept.
-2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see [@Hall2015]).
+2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015].
 This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
@@
 During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
-During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
+During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$std.error[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
 5).
@@
 \
-We estimated gross leaf production from measured litter production ($P_L$) and temporal dynamics of $L_\tau$.
-Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to $P_L$ to arrive at an estimate of the leaf production.
+We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $L_\tau$.
+Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to P~L~ to arrive at an estimate of the leaf production.
 The temporal dynamics of litter and leaf production are shown in Fig.
-5, and demonstrate the irregular temporal pattern in leaf production and $P_L$.
+5, and demonstrate the irregular temporal pattern in leaf production and P~L~.
 We found no significant difference in either litter or leaf production between C~a~ treatments (p > 0.2 for both, assessed with linear-mixed effects models).
@@
 \
-Next, we related $P_L$ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig.
+Next, we related P~L~ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig.
 6a).
-If leaf lifespan is constant, $P_L$ would be linearly related to $L_\tau$.
-Instead, we found that in eC~a~, $P_L$ was lower at a given $L_\tau$ as compared to aC~a~.
+If leaf lifespan is constant, P~L~ would be linearly related to $L_\tau$.
+Instead, we found that in eC~a~, P~L~ was lower at a given $L_\tau$ as compared to aC~a~.
 We estimated mean leaf lifespan from Fig.
-6a as the ratio of $L_\tau$ to annual $P_L$.
+6a as the ratio of $L_\tau$ to annual total P~L~.
 We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig.
@@
 
-```{r}
-options(scipen = 1, digits = 5)
+```{r, echo=FALSE, message=FALSE}
+options(scipen = 1, digits = 2)
 ```
@@
 **Figure 2**.
-Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA).
+Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
 This relationship was used to calculate the *L* anomaly in Fig.

### 2015-05-08 R/figures.R

@@ -96,7 +96,7 @@ figure2 <- function(df){
   par(mar=c(5,5,2,2), cex.lab=1.2,xaxs="i", yaxs="i")
   with(df, plot(BA, LAI.mean, pch=19, cex=1.2, col=my_ringcols(),
                 xlab=expression(Basal~area~~(m^2~ha^-1)),
-                ylab=expression(bar(LAI)~~(m^2~m^-2)),
+                ylab=expression(bar(italic(L))~~(m^2~m^-2)),
                 panel.first=predline(lm(LAI.mean ~ BA, data=ba)),
                 ylim=c(1.2,2.4), xlim=c(18,40)))
   legend("bottomright", as.character(1:6), pch=19, bty='n',
@@ -108,7 +108,7 @@ figure2 <- function(df){
 figure3 <- function(df,
                     xlim=NULL, ylim=NULL,
                     legend=TRUE,
-                    ylab=expression(LAI~anomaly~(m^2~m^-2)),
+                    ylab=expression(italic(L)~anomaly~(m^2~m^-2)),
                     cex.lab=1.1, cex.axis=0.8, cex.legend=0.7,
                     legendwhere="topleft",
                     setpar=TRUE,axisline=3,
@@ -123,7 +123,8 @@ figure3 <- function(df,
   if(is.null(xlim))xlim <- with(df, c(min(Date)-15, max(Date)+15))
   if(is.null(ylim))ylim <- c(-0.8,0.8)
   
-  smoothplot(Date, LAIanomaly, g=treatment, R="Ring", ylim=ylim, xlim=xlim, ylab=ylab, xlab="",
+  smoothplot(Date, LAIanomaly, g=treatment, R="Ring", ylim=ylim, xlim=xlim, 
+             ylab=ylab, xlab="",
              data=df, kgam=18, axes=FALSE,
              polycolor=c(alpha("royalblue",0.7),alpha("pink",0.7)))
   
@@ -152,7 +153,7 @@ figure4 <- function(df){
                xlab=expression(Leaf~litter~production~~(m^2~m^-2~mon^-1)),
                xlim=c(0,0.6),
                ylim=c(-0.3,0.5),
-               ylab=expression(Delta*LAI~from~tau[d]~(m^2~m^-2~mon^-1))))
+               ylab=expression(Delta*italic(L)~from~tau[d]~(m^2~m^-2~mon^-1))))
[…]

### 2015-05-08 R/data_processing.R
CUP analysis in R_other

@@ -478,8 +478,6 @@ makeCloudy <- function(df,
   # Toss nighttime data. 
   df <- subset(df, sunelevation > minSolarElevation)
     
-  df$PAR_Den_2_Avg[df$Ring == "R4" & df$Date == as.Date("2013-9-8")] <- NA
-  
   # Fdiff , fraction diffuse radiation.
   df$Fdiff <- with(df, DiffuseSS / TotalSS)
   

### 2015-05-08 R_other/cup.R

@@ -1,0 +1,37 @@
+source("load.R")
+
+parfiles <- searchHIEv("ternhect_underpar")
+
+
+makeCUPPAR <- function(){
+  
+  # CSV files are not actually CSV files, they are TOA5 
+  cup <- downloadTOA5("TERNHECT_UNDERPAR", allowedExtensions = c(".dat",".csv"))
+  
+  cupa <- dplyr::summarize(group_by(cup,DateTime=nearestTimeStep(DateTime,30)),
+                    PAR_Den_1_Avg=mean(PAR_Den_Avg.1.),
+                    PAR_Den_2_Avg=mean(PAR_Den_Avg.2.),
+                    PAR_Den_3_Avg=mean(PAR_Den_Avg.3.)
+                    )
+  facepar <- as.data.frame(makeFACEPAR(uploadnew=FALSE))
+  
+  facemean <- dplyr::summarize(group_by(facepar,DateTime),
+                               LI190SB_PAR_Den_Avg = mean(LI190SB_PAR_Den_Avg, na.rm=TRUE),
+                               TotalSS = mean(TotalSS),
+                               DiffuseSS = mean(DiffuseSS))
+                               
+  
+  cupa <- merge(cupa, facemean, all=FALSE)
+  
+return(cupa)
+}
+
+cupa <- makeCUPPAR()
+cupa$Date <- as.Date(cupa$DateTime)
+
+cup_cloudy <- makeCloudy(cupa, Fdiff=0.95
+
+cupgap <- summaryBy(Gapfraction ~ Date, data=cup_cloudy, FUN=mean, na.rm=TRUE)
+
+cupgap <- subset(cupgap, Date > as.Date("2014-7-20"))
+

### 2015-05-08 run.R

@@ -7,6 +7,9 @@ source("load.R")
 # 30minutely radiation data
 facepar <- makeFACEPAR(uploadnew=FALSE)
 
+# One bad row
+facepar$PAR_Den_2_Avg[facepar$Ring == "R4" & facepar$Date == as.Date("2013-9-8")] <- NA
+
 # Rainfall data, averaged across 3 rain gauges.
 facerain <- get_rain("rawmean")
 faceraindaily <- get_rain("daily")

### 2015-05-08 R_other/cup.R
CUP smoothed LAI

@@ -3,6 +3,9 @@ source("load.R")
 parfiles <- searchHIEv("ternhect_underpar")
 
 
+load("cache/lai_workspace.RData")
+
+
 makeCUPPAR <- function(){
   
   # CSV files are not actually CSV files, they are TOA5 
@@ -13,7 +16,7 @@ makeCUPPAR <- function(){
                     PAR_Den_2_Avg=mean(PAR_Den_Avg.2.),
                     PAR_Den_3_Avg=mean(PAR_Den_Avg.3.)
                     )
-  facepar <- as.data.frame(makeFACEPAR(uploadnew=FALSE))
+  # facepar <- as.data.frame(makeFACEPAR(uploadnew=FALSE))
   
   facemean <- dplyr::summarize(group_by(facepar,DateTime),
                                LI190SB_PAR_Den_Avg = mean(LI190SB_PAR_Den_Avg, na.rm=TRUE),
@@ -29,9 +32,24 @@ return(cupa)
 cupa <- makeCUPPAR()
 cupa$Date <- as.Date(cupa$DateTime)
 
-cup_cloudy <- makeCloudy(cupa, Fdiff=0.95
+cup_cloudy <- makeCloudy(cupa, Fdiff=0.95, bysensor=TRUE)
 
-cupgap <- summaryBy(Gapfraction ~ Date, data=cup_cloudy, FUN=mean, na.rm=TRUE)
+cupgap <- summaryBy(Gapfraction1 + Gapfraction2 + Gapfraction3 ~ Date, 
+                    data=cup_cloudy, FUN=mean, na.rm=TRUE)
 
-cupgap <- subset(cupgap, Date > as.Date("2014-7-20"))
+cupgap <- subset(cupgap, Date > as.Date("2014-7-30"))
+cupgap$LAI1 <- with(cupgap, -log(Gapfraction1.mean)/0.5)
+cupgap$LAI2 <- with(cupgap, -log(Gapfraction2.mean)/0.5)
+cupgap$LAI3 <- with(cupgap, -log(Gapfraction3.mean)/0.5)
+
+
+dfa <- summaryBy(LAI ~ Date, FUN=mean, na.rm=TRUE, data=facegap_cloudy_byring, keep.names=TRUE)
+dfa <- subset(dfa, Date > min(cupgap$Date))
+
+smoothplot(Date, LAI, data=dfa, kgam=18, ylim=c(1,2.5))
[…]

### 2015-05-08 make_figures.R

@@ -4,8 +4,8 @@ source("R/figures.R")
 source("R/derivSimulCI.R")
 
 
-maxdate <- as.Date("2015-3-1")
-# maxdate <- as.Date(today())
+#maxdate <- as.Date("2015-3-1")
+maxdate <- as.Date(today())
 
 to.pdf(figure1(subset(facegap_cloudy_byring,Date < maxdate), ramp),
        filename="output/figures/Figure1.pdf",

### 2015-05-10 R/data_processing.R
Now works with offline setting with new HIEv

@@ -13,7 +13,7 @@ return(ramp)
 
 get_solvars <- function(){
   
-  downloadCSV(c("P0037","SOLVARS"))
+  downloadCSV(c("P0037","SOLVARS"), tryoffline=TRUE)
   
 }
 
@@ -350,9 +350,9 @@ get_rain <- function(how=c("daily", "raw", "rawmean")){
   how <- match.arg(how)
   
   # Rain.
-  R1 <- downloadTOA5(filename="FACE_R1_T1_Rain")
-  R3 <- downloadTOA5(filename="FACE_R3_T1_Rain")
-  R4 <- downloadTOA5(filename="FACE_R4_T1_Rain")
+  R1 <- downloadTOA5(filename="FACE_R1_T1_Rain", tryoffline=TRUE)
+  R3 <- downloadTOA5(filename="FACE_R3_T1_Rain", tryoffline=TRUE)
+  R4 <- downloadTOA5(filename="FACE_R4_T1_Rain", tryoffline=TRUE)
   R1$Ring <- "R1"
   R3$Ring <- "R3"
   R4$Ring <- "R4"
@@ -370,7 +370,7 @@ get_rain <- function(how=c("daily", "raw", "rawmean")){
     names(rainagg)[3] <-"Rain"
     rainw <- reshape(rainagg, direction="wide", timevar="Ring", idvar="Date")
     
-    ros <- downloadTOA5("ROS_WS_Table15")
+    ros <- downloadTOA5("ROS_WS_Table15", tryoffline=TRUE)
     rosrain <- summaryBy(Rain_mm_Tot ~ Date, FUN=sum, data=ros, keep.names=TRUE)
     names(rosrain)[2] <- "Rain.ROS"
     
@@ -391,7 +391,7 @@ get_rain <- function(how=c("daily", "raw", "rawmean")){
 
 get_rosTair <- function(){
   
-  d <- downloadTOA5(c("ROS_WS","Table05min"), maxnfiles=500)
+  d <- downloadTOA5(c("ROS_WS","Table05min"), maxnfiles=500, tryoffline=TRUE)
   
   d <- as.data.frame(dplyr::summarize(group_by(d, Date),
                                       Tair=mean(AirTC_Avg, na.rm=TRUE),
@@ -415,7 +415,7 @@ get_soilwater <- function(how=c("mean","byring")){
     rowMeans(dfr, na.rm=TRUE)
   }
   
[…]

### 2015-05-10 load.R
now depends on HIEv > 0.7

@@ -11,7 +11,7 @@ library(Hmisc)    # approxExtrap
 library(rmarkdown) # render
 
 
-if(packageVersion('HIEv') < '0.6.4')
+if(packageVersion('HIEv') < '0.7')
   stop("Please update the HIEv package!")
 
 # set path for HIEv

### 2015-05-11 R/utils.R
Uses Library, install packages only when not available

@@ -1,0 +1,12 @@
+Library <- function(pkg, ...){
+  
+  PACK <- .packages(all.available=TRUE)
+  pkgc <- deparse(substitute(pkg))
+  
+  if(pkgc %in% PACK){
+    library(pkgc, character.only=TRUE)
+  } else {
+    install.packages(pkgc, ...)
+  }
+  
+}

### 2015-05-12 manuscript.Rmd
fix so manuscript can be actually compiled with figures
Uses Library, install packages only when not available
added todo list for GCB submission

@@
 source("load.R")
+
+source("R/functions-figures.R")
+source("R/figures.R")
+source("R/derivSimulCI.R")
 
@@
 ```
+
+<!--- 
+TO DO:
+Answers to the following questions (max 50 words per answer).
+Please take time to prepare your answers to these questions; this information may be used to determine if the manuscript should progress to stage two of the review process.
+
+What is the scientific question you are addressing?
+What is/are the key finding(s) that answers this question?
+What are the three most recently published papers that are relevant to this question?
+Why is this work important and timely?
+Does your paper falls within the scope of GCB; what biological AND global change aspects does it address?
+Provide information about suggested and non-preferred reviewers including their area of research and how it relates to your paper.
+Also provide justification for why you do not prefer certain reviewers.
+Provide justification for any non-conformance to author guidelines and/or formatting
+
+1. Title page
+      i. Title: this should be concise and informative            
+      ii. Running head: a shortened title with no more than 45 characters, including spaces
+      iii. List of authors
+      iv. Institute or laboratory of origin: Where authors have different addresses, use numbered superscripts to refer to each address provided
+      v. Corresponding author: include their telephone, fax and email details
+      vi.
+Keywords: 6 – 10 key words or short phrases to enable retrieval and indexing by searching techniques.
+Authors are encouraged to include scientific names, common names, and pseudonyms that are not mentioned in the title.
+      vii. Type of Paper
+
+      7. Conclusions
+Only Reviews may have conclusions sections. All other article types should incorporate any conclusions into the Discussion section, and should not include a “Conclusions” subheading.
+
+      13. Figure captions
+Figure captions should be listed one after the other, as part of the text document, separate from the figure files.
+Do not include captions below the figures.
+Enough detail should be given so that the figure can be understood without reference to the text.
+In the full-text online edition of the journal, figure captions may be truncated in abbreviated links to the full screen version.
+Therefore, the first 100 characters of any caption should inform the reader of key aspects of the figure.
+If you wish to publish figures in color online and grayscale in print, figure captions should be appropriate for both color and grayscale versions; do not refer to colors in the figure.
+
+15. Supporting information
+Appendicies will not be published in print, but may be submitted as Supporting Information to be published in the online version of the article.
+These files are not included in the typeset manuscript, but are downloadable and fully searchable from the HTML version of the article.
+This material should be submitted with the original manuscript in order that it can be included in the review process.
+The conclusions of an Author's manuscripts should not depend on the material supplied in Supporting Information.
+Supporting information will be made available in exactly the same form as originally provided; it will not be copyedited or typeset.
+All supporting information must be referred to in the manuscript with a leading capital S (e.g., fig.
+S2 for the second supporting information figure).
+Authors should include a Supporting Information section after the references in the main article.
+This section consists of short captions for each supporting information file.
+Full captions should be included within the online supporting information file.
+Please see Wiley-Blackwell's Supporting Information Guidelines for more information.
+
+Figure formatting
+http://goo.gl/iARuKc
+
+-->
 

### 2015-05-12 R/figures.R
number of points for prediction in smoothplot added as argument

@@ -67,7 +67,7 @@ figure1 <- function(df, ramp){
   
   smoothplot(Date, Gapfraction.mean, g=Ring, data=df, k=18,axes=FALSE,
              ylim=c(0,0.4),
-             xlab="",xlim=xl,
+             xlab="",xlim=xl, npred=201,
              ylab=expression(tau[d]~~("-")),
              pointcols=rep("grey",8), linecols=my_ringcols())
   

### 2015-05-12 R/functions-figures.R

@@ -180,6 +180,7 @@ smoothplot <- function(x,y,g=NULL,data,
                        xlab=NULL, ylab=NULL,
                        polycolor=alpha("lightgrey",0.7),
                        plotit=TRUE, add=FALSE,
+                       npred=101,
                        ...){
   
   fittype <- match.arg(fittype)
@@ -282,7 +283,7 @@ smoothplot <- function(x,y,g=NULL,data,
     for(i in 1:length(fits)){
       
       if(fittype == "gam"){
-        nd <- data.frame(X=seq(hran[[i]][1], hran[[i]][2], length=101))
+        nd <- data.frame(X=seq(hran[[i]][1], hran[[i]][2], length=npred))
         if(!inherits(fits[[i]], "try-error")){
           p <- predict(fits[[i]],nd,se.fit=TRUE)
           addpoly(nd$X, p$fit-2*p$se.fit, p$fit+2*p$se.fit, col=polycolor[i])

### 2015-05-19 manuscript.Rmd
Teresa's edits

--- 2015-05-08 manuscript.Rmd
@@
 ---
-title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
-  atmospheric [CO~2~] but tracks water availability
+title: Canopy leaf area of an evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability
 author: ''
@@
 *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that *L* increases with eC~a~, largely in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+Ecosystem models generally predict that *L* increases with eC~a~, largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
 This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
@@
 ## Site description and experimental design
-The EucFACE experiment is located on an ancient alluvial floodplain about 5 km from the Hawkesbury River in western Sydney (Australia) at 30 m a.s.l.
-elevation (33^o^ 37'S, 150^o^ 44'E).
-It is located within a remnant patch  of native Cumberland Plain woodland [@Tozer2003].
-The vegetation, maintained in this location since the 1880’s, retains some old-growth trees and some re-growth vegetation though it has been undisturbed for ca.
-70 years.
+The EucFACE experiment is located on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.).
+The site is part of a remnant patch  of native Cumberland Plain woodland [@Tozer2003].
+The vegetation consists of a native woodland undisturbed for ca.
+70 years and maintained in this location since the 1880’s.
 The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
@@
 \
-The soil at the site is a loamy sand in the Clarendon Formation [@Barton2010; @Isbell2002], a slightly-acid, low-organic C, well-drained soil.
-The soil surface is a loamy sand with > 75% sand content, and from ca.
+At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca.
 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay.
@@
 Groundwater depth is 12.6 – 12.9 m below the soil surface.
-The woodland ranges in density from 600-1000 trees ha^-1^.
-The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E.
+The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots).
+There is one single dominant species (*Eucalyptus tereticornis* Sm.) with some scattered individuals of *E.
 amplifolia* Naudin.
-The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst.
+The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst.
 & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg.
@@
 ## *L* from diffuse transmittance
-In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al.
-[-@Norby2003].
+In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
 Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct.
@@
 For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
-We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles.
+We further trimmed the data by excluding low solar e levation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles.
 Finally canopy transmittance was calculated for every 30-min time step as,
@@
 These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
-3), weighted towards the zenith.
+2), weighted towards the zenith.
 
@@
 The average half-width of the confidence interval on the *L* anomaly (Fig.
-4) was `r LanomCIamb` m^2^ m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment.
+4) was `r LanomCIamb` m^2^m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment.
 This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` m^2^m^-2^).
@@
 
-##Continuous measurement of *L*
+## Continuous measurement of *L*
 Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.

### 2015-05-19 manuscript.Rmd
some revisions done

@@
 ---
-title: Canopy leaf area of an evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability
+title: Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability
 author: ''
@@
 Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
-Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of unmanaged forests have been poorly quantified.
+Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of forests have been poorly quantified.
 We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
@@
 We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear annual pattern.
-Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady during periods of low rainfall.
+Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
 Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).

### 2015-05-20 manuscript.Rmd
revisions up to L from diff trans

@@
 # Litter - dLAI correlations Figure 4
-dl <-dLAIlitter
+dl <- dLAIlitter
 dl$Y <- with(dl, dLAI * 30.5/ndays)
@@
 Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
-Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of forests have been poorly quantified.
+Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified.
 We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
@@
 Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe.
-We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear annual pattern.
+We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear seasonal pattern.
 Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
 Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
-Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly longer under elevated [CO~2~] (p `r pval(LLtest$p.value)`).
-Our main finding that *L* was not responsive to elevated CO~2~ are consistent with other forest FACE studies, but contrast the positive response commonly predicted by ecosystem models.
+Leaf lifespan, estimated from mean *L* and total annual litter production, was up to two months longer under elevated [CO~2~] (`r LLtest$estimate2` vs.
+`r LLtest$estimate1` yrs; p `r pval(LLtest$p.value)`).
+Our main finding that *L* was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of *L* commonly predicted by many ecosystem models.
 
@@
 *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that *L* increases with eC~a~, largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].
+Ecosystem models generally predict that *L* increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy vegetation-atmosphere fluxes as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
 Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
@@
 Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
-Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011].
+Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
 Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
-Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites [@Norby2005].
+Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased *L* in response to eC~a~ [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
 Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
@@
 [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~.
-Based on the same water-savings principle, we predict that *L* should increase under eC~a~ in water-limited forests.
+Based on the same water-savings principle, we predicted that *L* should increase under eC~a~ in water-limited forests.
 
 \
-Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g.
+Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and photo period-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g.
 @Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
@@
 However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
-Understanding dynamics of *L* in response to water availability and its potential interactions with C~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
+Understanding dynamics of *L* in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
 
 \
-We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated woodland exposed to eC~a~, the EucFACE experiment.
-The woodland at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
-We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~.
-In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [cf.
-@Norby2011].
+We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
+The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration (long-term mean of 1500mm; BOM) and mean annual precipitation of 800mm.
+The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
+We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation.
+In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a positive response of *L* to eC~a~ [cf.
+@Norby2011], either via changes in rates of leaf production or leaf retention (leaf lifespan).
 Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
@@
 ## Site description and experimental design
-The EucFACE experiment is located on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.).
-The site is part of a remnant patch  of native Cumberland Plain woodland [@Tozer2003].
+The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.).
 The vegetation consists of a native woodland undisturbed for ca.
@@
 1500mm (*to be refined!*), giving a ratio of P/PET of ca.
-0.53 (*quite dry - details to be added*).
+0.53.
+Over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating very frequent water limitation.
 
@@
 Groundwater depth is 12.6 – 12.9 m below the soil surface.
-The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots).
-There is one single dominant species (*Eucalyptus tereticornis* Sm.) with some scattered individuals of *E.
+The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E.
 amplifolia* Naudin.
@@
 Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
-The vegetation within three randomly selected rings was exposed to a CO~2~ mole fraction (C~a~) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots.
+The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction (C~a~) higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots.
 The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode.
 However, C~a~ was increased gradually for six months at a rate of ca.
-30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode was reached (Fig.
+30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig.
 1a).
@@
 Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings.
-Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers.
+Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers (Campbell Scientific, Townsville, Australia).
 Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).
@@
 In each ring, 8 litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor.
-Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc.
+Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
 Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).

### 2015-05-20 manuscript.Rmd
fixed digits formatting; deleted conclusions

@@
 # Number of digits in printing of numbers
-options(scipen = 1, digits = 3)
+# options(scipen = 1, digits = 3)
 
-# Some definitions of constants
+# Shortcut to significant digits formatting
+f <- function(x, digits=2)sprintf(x, fmt=paste0("%.",digits,"f"))
+
+# Some definitirons of constants
 .gambase <- 18
@@
 Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe.
-We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear seasonal pattern.
+We found that *L* varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period, but did not show a clear seasonal pattern.
 Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
 Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
-Leaf lifespan, estimated from mean *L* and total annual litter production, was up to two months longer under elevated [CO~2~] (`r LLtest$estimate2` vs.
-`r LLtest$estimate1` yrs; p `r pval(LLtest$p.value)`).
+Leaf lifespan, estimated from mean *L* and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs.
+`r f(LLtest$estimate1)` yrs; p `r pval(LLtest$p.value)`).
 Our main finding that *L* was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of *L* commonly predicted by many ecosystem models.
@@
 Groundwater depth is 12.6 – 12.9 m below the soil surface.
-The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E.
+The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r f(mean(ba$BA))`, SE = `r f(sd(ba$BA)/sqrt(6))` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E.
 amplifolia* Naudin.
@@
 We used linear regression without intercept to estimate *c* in Eq.
-4 as `r calib` (SE = `r tidy(caliblm)$std.error`).
+4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`).
 It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq.
@@
 [-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)`% higher than $L_p$ (Fig.
-SI2, R^2^=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`) (Fig.
+Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (Fig.
+SI2, R^2^=`r f(glance(lmphoto)$r.squared)`, RMSE=`r f(glance(lmphoto)$sigma)`) (Fig.
 S2).
 We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig.
-S2), giving *k* = `r kopt`.
+S2), giving *k* = `r f(kopt)`.
 These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
@@
 We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig.
-3, to yield a ‘*L* anomaly’.
+3, to yield a '*L* anomaly'.
 This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.
@@
 # Results
-Canopy transmittance ($\tau_d$) varied between `r min(df$Gapfraction.mean)` and `r max(df$Gapfraction.mean)`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
+Canopy transmittance ($\tau_d$) varied between `r f(min(df$Gapfraction.mean))` and `r f(max(df$Gapfraction.mean))`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig.
 1).
@@
 The average half-width of the confidence interval on the *L* anomaly (Fig.
-4) was `r LanomCIamb` m^2^m^-2^ for aC~a~, and `r LanomCIele` m^2^m^-2^ for the eC~a~ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` m^2^m^-2^).
+4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment.
+This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` $m^2$ $m^{-2}$ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean *L* of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
 
@@
 4).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
-During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$std.error[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$estimate[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
+During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
 5).
@@
 We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig.
-6b), with LL = `r LLtest$estimate1`yr in aC~a~ and `r LLtest$estimate2`yr in eC~a~.
+6b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
 
@@
 The mean *L* at the EucFACE was ca.
-`r mean(ba$LAI.mean)` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
+`r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
 [-@Iio2013].
@@
 To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
-For example, Drake et al.
-(*in review*) reported increased soil respiration under eC~a~, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.
 
@@
 [-@Macfarlane2007].
-The two methods compared quite well (R^2^ = `r glance(lmcanphot)$r.squared`) (Fig.
+The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Fig.
 S2), suggesting the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
@@
 2005), but this process has been poorly quantified.
-
-# Conclusions
-Leaf area index (*L*) of a natural *Eucalyptus* woodland was unresponsive to elevated [CO~2~] after `r nummon` months of treatment and `r nummon-3` months at +150ppm CO~2~ elevation.
-Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage).
-Leaf lifespan, estimated from mean *L* and annual litter production, was `r round(abs(LLtest$estimate*12),1)` months longer in the elevated [CO~2~] treatment.
-
-\
-We found no clear annual pattern in dynamics of *L*, instead, a net increase in *L*, indicating leaf growth, consistently commenced after heavy rainfall following a dry period.
-The complex dynamics of litter production and the rapid response of litter production and leaf growth to soil water availability challenge current-generation prognostic *L* and phenology models.
-High resolution time series such as the one we present in this study can be used as a basis to develop new models of phenology for evergreen woodlands.
 

### 2015-05-21 manuscript.Rmd
Co-author revisions complete; More refs in leaf lifespan paragraph; fig6 updated

@@
 Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
-Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased *L* in response to eC~a~ [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
+Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased *L* in response to eC~a~ [@McCarthy2007; @Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
 Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
@@
 In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a positive response of *L* to eC~a~ [cf.
-@Norby2011], either via changes in rates of leaf production or leaf retention (leaf lifespan).
+@Norby2011], either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
 Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
@@
 \
-Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’).
+Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as 'rings').
 Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high.
@@
 ## Terminology
-In the following, we use the term ‘leaf’ area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e.
+In the following, we use the term 'leaf' area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e.
 woody stems and infrastructure.
@@
 ## *L* from diffuse transmittance
-In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
+In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
 Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct.
@@
 All PAR data were averaged in 30 min time steps.
-At a nearby site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy.
+At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca.
+5m above the canopy.
 For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
-We further trimmed the data by excluding low solar e levation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles.
+We further trimmed the data by excluding low solar e levation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles.
 Finally canopy transmittance was calculated for every 30-min time step as,
@@
 where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR (µmol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
-We then calculated daily averages of $\tau_d$, excluding days where less than 4 time steps were available, and days where the standard deviation among the three PAR sensors in a ring was > `r .maxSD` (in units of $\tau_d$).
+We then calculated daily averages of $\tau_d$, excluding days where fewer than 4 time steps (i.e.
+2 hours) were available, and days where the standard deviation among the three PAR sensors in a ring was > `r .maxSD` (in units of $\tau_d$).
 
@@
 To overcome violations in the assumptions underlying Eq.
-2, we calibrated $L_\tau$ against cumulative litter production during a drought period in 2013 (indicated in Fig.
+2, we calibrated $L_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
 1).
-During this period (14 July 2013 – 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
+During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
 1).
 During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm).
-A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
+A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
 Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance.
@@
 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`).
-It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq.
+It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology).
+However, this approximately constant area cancels out in the left-hand side of Eq.
 3.
-We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought (see Fig.
+We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought period (see Fig.
 S1).
@@
 [-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (Fig.
-SI2, R^2^=`r f(glance(lmphoto)$r.squared)`, RMSE=`r f(glance(lmphoto)$sigma)`) (Fig.
+Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^=`r f(glance(lmphoto)$r.squared)`, RMSE=`r f(glance(lmphoto)$sigma)`) (Fig.
 S2).
@@
 For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed *L* dynamics.
+Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
 Random effects were included to account for the grouping of data within FACE rings, except when we analysed average *L* across all rings (Fig.
@@
 \
-For comparison between *L* from photos and PAR transmittance, we used simple linear regression.
+For comparison between *L* from photos and PAR transmittance, we used linear regression.
 To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
@@
 \
-An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept.
+An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in Sept.
 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015].
@@
 \
-We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig.
+We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any point in time, Fig.
 3).
@@
 A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the *L* anomaly (p `r pval(LanomAOV[[3]][1])`).
-What difference between C~a~ treatments could we have reasonably detected?
+But what difference between C~a~ treatments could we have reasonably detected?
 The average half-width of the confidence interval on the *L* anomaly (Fig.
 4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` $m^2$ $m^{-2}$ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean *L* of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
+This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean *L* of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
 
@@
 The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected.
-We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig.
+We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to that shown for $\tau_d$ in Fig.
 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$).
@@
 4).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$estimate[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
 During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
@@
 We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $L_\tau$.
-Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to P~L~ to arrive at an estimate of the leaf production.
+For each litter collection period, the net change in $L_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries.
+Gross leaf production was then calculated as the sum of the net change in $L_\tau$ and the amount of leaf litter produced during that period.
 The temporal dynamics of litter and leaf production are shown in Fig.
 5, and demonstrate the irregular temporal pattern in leaf production and P~L~.
-We found no significant difference in either litter or leaf production between C~a~ treatments (p > 0.2 for both, assessed with linear-mixed effects models).
+We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).
 
@@
 If leaf lifespan is constant, P~L~ would be linearly related to $L_\tau$.
-Instead, we found that in eC~a~, P~L~ was lower at a given $L_\tau$ as compared to aC~a~.
+Instead, we found that in eC~a~, P~L~ was lower at a given $L_\tau$ as compared to aC~a~ (Fig.
+6a).
 We estimated mean leaf lifespan from Fig.
 6a as the ratio of $L_\tau$ to annual total P~L~.
-We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig.
+We found a significant (p `r pval(LLtest$p.value)`) increase in leaf lifespan (LL) with elevated C~a~ (Fig.
 6b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
@@
 7).
-These events occurred in spring and summer, with no clear annual cycle (Oct.
+These events occurred in spring and summer, with no clear seasonal cycle (Oct.
 2012, Jan.
@@
 One brief significant positive dL/dt (June 2013) did not correspond to an actual leaf flushing event, but it was small in magnitude.
-It is also obvious that every leaf flush event occurred shortly after a heavy rainfall event (Fig.
+It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig.
 4 and 6), when the soil water content in the top 80 cm increased to ca.
@@
 ## Response of *L* to elevated [CO~2~]
-We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~.
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from Nov.
+2012 to Jan.
+2013.
 We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability.
@@
 [-@Norby2003], we found no significant effect of elevated C~a~ on *L*.
+Because estimated leaf lifespan averaged `r round(mean(ba$LL),1)` yrs (Fig.
+6), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
 It could be argued that the change in *L* in response to eC~a~ was too small to be detected.
-Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
+Natural variation in *L* among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
 To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate.
@@
 This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~.
+First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*.
 They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
@@
 Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~.
-It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because compare to other forest FACE sites, the EucFACE is very strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
+In contrast to other forest FACE sites, the EucFACE is very strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
+Hence, it may be too early to draw general conclusions across FACE sites in terms of *L* response.
 
@@
 We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig.
-6).
+6) (an increase of `r f(100*with(LLtest,estimate2/estimate1 - 1))`%).
+Further years of observation are needed to determine if such a difference is sustained.
 Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
@@
 \
-Despite a lack of response of canopy *L*, we did find a significantly longer leaf lifespan (LL) in eC~a~, consistent with our second hypothesis.
-We could not determine whether there was an interaction with water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
+Despite a lack of response of canopy *L*, we did find a significantly longer leaf lifespan (LL) in eC~a~.
+We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
 6).
@@
 5), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
-There is little evidence from other FACE sites that LL is affected by eC~a~.
+There is mixed evidence from other FACE sites on the relatinship between LL and eC~a~.
+Norby et al.
+[-@Norby2003] reported inconsistent effects of eC~a~ on leaf area duration in a deciduous *Liquidambar* plantation (ORNL-FACE).
+At the same site, Warren et al.
+[-@Warren2011] found increased senescence in response to a combined drought and heat wave.
+McCarthy et al.
+[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of *L* by eC~a~.
 Taylor et al.
 [-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
-Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~, leading to similar *L* in ambient C~a~ despite higher productivity.
-In contrast, Warren et al.
-[-@Warren2011] found increased senescence in eC~a~ at the ORNL FACE site (*Liquidambar styraciflua* plantation), in response to a combined drought and heat wave.
+Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ at the PopFACE, leading to similar *L* in ambient C~a~ despite higher productivity.
+Li et al.
+[-@Li2000] reported delayed senescence of the oldest leaf cohort in evergreen *Quercus myrtifolia* exposed to eC~a~ in an open-top chamber experiment.
+Clearly, no consistent picture emerges from these comparisons.
 
@@
 Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan (LL) is unknown (instead, we used *L* and litter production to estimate LL).
-We can also not rely on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
-tereticornis*, and would be useless anyway since *L* is highly variable temporally (Figs.
+We cannot not rely either on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
+tereticornis*, and would not be valuable anyway since *L* is temporally highly variable (Figs.
 1, 7).
@@
 1, 5, 7).
-We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999].
+We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to those during periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999].
 As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig.
@@
 Using diffuse PAR only avoids having to correct for these path length effects.
-
+\
 We used the opportunity provided by the high quality timeseries of *L* to test a simple photography-based method suggested by Macfarlane et al.
@@
 The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Fig.
-S2), suggesting the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
+S2), suggesting both diffuse transmittanc and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
 
@@
 Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
-Since leaf flushing has been shown to be closely related to soil water availability [Fig.
+Since leaf flushing has been shown to be closely related to soil water availability in eucalypts [Fig.
 7; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig.
@@
 ## Consequences for phenology models
-Current-generation phenology models are heavily biased towards temperate conditions [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardson2012; @Savoy2015].
+Current-generation phenology models are heavily biased towards temperate conditions and deciduous rather than evergreen species [@Primack2011], where leaf flushing occurs in spring in response to increase in temperature [@Richardson2012; @Savoy2015].
 In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig.
@@
 7).
+We argue that both minimum temperature and high soil moisture are required  for leaf growth of subtropical eucalypts such as *E.
+tereticornis*.
 
@@
 Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
-We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests.
+We argue that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests.
 In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability, as we have demonstrated that both processes are highly dynamic.
@@
 # Acknowledgments {.unnumbered}
-We thank Steven Wohl, Vinod Kumar, Craig Barton and Craig McNamara for maintaining the FACE site and its wide array of sensors, and Andrew Gherlenda, Julia Cooke and Catherine Beattie for significant efforts in separating and weighing the litter samples.
+We thank Steven Wohl, Vinod Kumar, Craig Barton and Craig McNamara for maintaining the FACE site and its wide array of sensors, and Andrew Gherlenda, Cassie Mosdal, Tamara Leetham and Catherine Beattie for significant efforts in separating and weighing the litter samples.
 We also thank Gerry Devine for support in data management.
@@
 Facilities and instrumentation at EucFACE were assembled as an initiative of the Australian Government as part of the Nation-building Economic Stimulus Package.
+Additional support from the Australian Research Council is acknowledged.
 
@@
 3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.
+Red symbols are elevated C~a~, blue symbols ambient C~a~.
 
@@
 **Figure S1**.
-Increase in gapfraction ($\tau_d$) and total litter production during the 2013 drought, calculated for each ring separately.
-Blue colours are ambient, red colours eleveated C~a~.
+Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately.
+Blue colours are ambient, red colours elevated C~a~.
 (a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
 (b) Mean litter production for each ring.
-The confidence interval was calculated based on he 8 litter traps in each ring.
+The confidence interval was calculated based on the 8 litter traps in each ring.
 Note the overlapping confidence intervals for all rings.

### 2015-05-21 R/figures.R

@@ -203,15 +203,26 @@ figure6 <-  function(df){
 
   par(mar=c(5,5,2,2), cex.lab=1.2, xaxs="i", yaxs="i", mfrow=c(1,2))
   
+  LLs <- seq(0.8,2,by=0.1)
+  
   with(df, plot(LAI.mean.litterperiod, LAIlitter_annual, pch=19, cex=1.2, col=my_ringcols(),
                 ylab=expression(Litter~production~~(m^2~m^-2~yr^-1)),
                 xlab=expression(bar(italic(L))~~(m^2~m^-2)),
                 panel.first={
-                  for(z in seq(0.5,2,by=0.1))abline(0,z,col="grey", lty=5)
+                  for(z in LLs)abline(0,1/z,col="grey", lty=5)
                 },
                 xlim=c(1.2,2.2), ylim=c(1.2,2.2)))
   plotlabel("(a)", "topright")
+  legend("topleft", as.character(1:6), pch=19, bg="white", box.col="white",
+         col=my_ringcols(), title="Ring", cex=0.8)
   box()
+  x <- c(2.05, 2.05,2.05,1.965,1.83)
+  lls <- seq(1.3,0.9,by=-0.1)
+  b1 <- 1/lls
+  y <- b1*x
+  text(x,y,pos=4, labels=as.character(lls), cex=0.75, col="darkgrey",
+       font=3)
+  text(1.68, y[length(y)], "LL = ", font=3, cex=0.75, col="darkgrey", pos=4)
   
   with(df, plot(as.numeric(treatment), LL, axes=FALSE, xlim=c(0,3), ylim=c(0.5,1.5),
                 pch=19, col=my_ringcols(), xlab="",cex=1.2,
@@ -230,7 +241,7 @@ figure6 <-  function(df){
 }
 
 
-figure7 <- function(df, facesoilwater, faceraindaily, airt){
+figure7 <- function(df, facesoilwater, faceraindaily, airt, kgam=18){
     
   xin <- 0.02 # for panel label x inset
   
@@ -241,7 +252,7 @@ figure7 <- function(df, facesoilwater, faceraindaily, airt){
   par(mfrow=c(4,1), mar=c(0,7,5,6), cex.lab=1.3, las=1, mgp=c(4,1,0),yaxs="i",tcl=0.2)
   
   # panel a
-  smoothplot(Date, LAI, data=dfa, kgam=18, pointcols="dimgrey", linecols="black", 
[…]

### 2015-05-21 make_figures.R

@@ -4,8 +4,8 @@ source("R/figures.R")
 source("R/derivSimulCI.R")
 
 
-#maxdate <- as.Date("2015-3-1")
-maxdate <- as.Date(today())
+maxdate <- as.Date("2015-3-1")
+#maxdate <- as.Date(today())
 
 to.pdf(figure1(subset(facegap_cloudy_byring,Date < maxdate), ramp),
        filename="output/figures/Figure1.pdf",

### 2015-05-21 manuscript.Rmd
Fixed PET estimate from Zomer; moved R code in manuscript to manuscript_calcs.R

@@
 ```{r echo=FALSE, message=FALSE}
-library(knitr)
-
-load("cache/lai_workspace.RData")
-source("load.R")
-
-source("R/functions-figures.R")
-source("R/figures.R")
-source("R/derivSimulCI.R")
-
-library(broom)
-library(lme4)
-library(lmerTest)
-library(car)
-library(reporttools)
-
-# shortcut to formatPval
-pval <- function(...)formatPval(..., includeEquality=TRUE)
-
-# Number of digits in printing of numbers
-# options(scipen = 1, digits = 3)
-
-# Shortcut to significant digits formatting
-f <- function(x, digits=2)sprintf(x, fmt=paste0("%.",digits,"f"))
-
-# Some definitirons of constants
-.gambase <- 18
-.maxSD <- 0.03
-
-# Main results
-df <- facegap_cloudy_byring
-
-# Some pre-calculations
-caliblm <- calibrateToDrought(df)$lmfit
-
-# Number of months of study
-r <- range(df$Date)
-r[2] <- maxdate
-nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))
-
-# Range mean LAI
-dfa <- summaryBy(LAI ~ Date, FUN=mean, na.rm=TRUE, data=df)
-rangeLAI <- round(range(dfa$LAI),1)
-
-# LAI ~ BA
-balm <- lm(LAI.mean ~ BA, data=ba)
-
-# Comparison flat canopy photo with transmittance
-lmphoto <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)
-
-# Significance CO2 on LAI anomaly
-df$Time <- as.factor(df$Date - min(df$Date))
-m3 <- lmer(LAIanomaly ~ treatment*Time + (1|Ring), data=df)
-LanomAOV <- Anova(m3)
-
-# Average CI width in Fig 4.
-fit <- smoothplot(Date, LAIanomaly, g=treatment, R="Ring", 
-                  data=facegap_cloudy_byring, plot=FALSE, kgam=.gambase)
-pa <- predict(fit$ambient, data.frame(X=as.numeric(facegap_cloudy_byring$Date)), se.fit=TRUE)
-pe <- predict(fit$elevated, data.frame(X=as.numeric(facegap_cloudy_byring$Date)), se.fit=TRUE)
-LanomCIamb <- mean(pa$se.fit)*2
-LanomCIele <- mean(pe$se.fit)*2
-
-# Detectable Ca effect on LAI
-Ldetect <- 100 * (LanomCIamb + LanomCIele) / mean(ba$LAI.mean)
-
-# Litter - dLAI correlations Figure 4
-dl <- dLAIlitter
-dl$Y <- with(dl, dLAI * 30.5/ndays)
-dl$X <- with(dl, dLAIlitter.mean * 30.5/ndays)
-dlup <- dl[dl$LAIchange == "increasing",]
-dldown <- dl[dl$LAIchange == "decreasing",]
-lmlitup <- lm(Y ~ X, data=dlup)
-lmlitdown <- lm(Y ~ X, data=dldown)
-
-# Leaf lifespan
-LLtest <- tidy(t.test(LL ~ treatment,data=ba))
-
-# Canopy photo - Ltau comparison
-lmcanphot <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)
-
-# With optimized k
-flatcan_byring2 <- agg_flatcan(flatcan, by="Ring", k=kopt)
-flatcan_byring2 <- add_PARLAI_to_flatcan(facegap_cloudy_byring,flatcan_byring2)
-lmcanphot2 <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring2)
-
+source("manuscript_calcs.R")
 ```
@@
 We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
-The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration (long-term mean of 1500mm; BOM) and mean annual precipitation of 800mm.
+The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration [long-term mean of 1350mm; @Zomer2008] and mean annual precipitation of 800mm.
 The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
@@
 The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
-The potential evapotranspiration (PET) is ca.
-1500mm (*to be refined!*), giving a ratio of P/PET of ca.
-0.53.
-Over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating very frequent water limitation.
+An estimate of potential evapotranspiration (PET) at the site (1350mm) was from the Global-PET database [http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
+0.59.
+In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating very frequent water limitation.
 
@@
 Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
-We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).
+We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.
 

### 2015-05-21 manuscript_calcs.R

@@ -1,0 +1,98 @@
+
+#' This script is called by Manuscript.Rmd, and includes additional calculations 
+#' used in the manuscript.
+
+
+source("R/utils.R")  # Library()
+
+Library(knitr)
+Library(broom)
+Library(lme4)
+Library(lmerTest)
+Library(car)
+Library(reporttools)
+
+# Load analysis cache.
+load("cache/lai_workspace.RData")
+
+# More packages, functions, etc.
+source("load.R")
+
+# Functions for figures
+source("R/functions-figures.R")
+source("R/figures.R")
+
+# CI on GAM derivative (Gavin Simpson)
+source("R/derivSimulCI.R")
+
+
+# shortcut to formatPval
+pval <- function(...)formatPval(..., includeEquality=TRUE)
+
+# Number of digits in printing of numbers
+# options(scipen = 1, digits = 3)
+
+# Shortcut to significant digits formatting
+f <- function(x, digits=2)sprintf(x, fmt=paste0("%.",digits,"f"))
+
+# Some definitirons of constants
+.gambase <- 18
+.maxSD <- 0.03
+
+# Main results
+df <- facegap_cloudy_byring
+
+# Calibration constant to litter fall.
+caliblm <- calibrateToDrought(df)$lmfit
+
+# Number of months of study
+r <- range(df$Date)
+r[2] <- maxdate
+nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))
+
+# Range mean LAI
+dfa <- summaryBy(LAI ~ Date, FUN=mean, na.rm=TRUE, data=df)
+rangeLAI <- round(range(dfa$LAI),1)
+
+# LAI ~ BA
+balm <- lm(LAI.mean ~ BA, data=ba)
+
+# Comparison flat canopy photo with transmittance
+lmphoto <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)
+
+# Significance CO2 on LAI anomaly
+df$Time <- as.factor(df$Date - min(df$Date))
+m3 <- lmer(LAIanomaly ~ treatment*Time + (1|Ring), data=df)
+LanomAOV <- Anova(m3)
+
+# Average CI width in Fig 4.
+fit <- smoothplot(Date, LAIanomaly, g=treatment, R="Ring", 
[…]

### 2015-05-21 manuscript.Rmd
Some changes to account for GCB format requirements. Sent around to co-authors as version 3.

@@
 \
+*Running head* : Leaf area index in elevated [CO~2~]
+
+\
 Remko A. Duursma^1^, Teresa E. Gimeno^1,2^, Matthias M. Boer^1^, Kristine Y. Crous^1^, Mark G. Tjoelker^1^, David S. Ellsworth^1^.
@@
 
+\
+*Corresponding author*: 
+Remko Duursma
+E: remkoduursma@gmail.com
+T: +61(0)422096908
+
+\
+*Keywords*: 
+leaf area index, FACE, phenology, drought, litter production
+
+\
+*Type of Paper*:
+Primary Research Article
+
 ```{r echo=FALSE, message=FALSE}
 source("manuscript_calcs.R")
+
+includeFigs <- FALSE
 ```
@@
 Our main finding that *L* was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of *L* commonly predicted by many ecosystem models.
-
-\
-*Keywords*: leaf area index, FACE, phenology, drought, litter production
 
@@
 
-# Acknowledgments {.unnumbered}
+# Acknowledgements {.unnumbered}
 We thank Steven Wohl, Vinod Kumar, Craig Barton and Craig McNamara for maintaining the FACE site and its wide array of sensors, and Andrew Gherlenda, Cassie Mosdal, Tamara Leetham and Catherine Beattie for significant efforts in separating and weighing the litter samples.
@@
 
-```{r fig.width=8, fig.height=6, echo=FALSE, message=FALSE}
+```{r fig.width=8, fig.height=6, echo=FALSE, message=FALSE, eval=includeFigs}
 figure1(subset(facegap_cloudy_byring,Date < maxdate), ramp)
@@
 
-```{r fig.width=4, fig.height=4, echo=FALSE, message=FALSE}
+```{r fig.width=4, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
 figure2(ba)
@@
 
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
 figure3(subset(facegap_cloudy_byring, Date < maxdate))
@@
 
-```{r fig.width=6, fig.height=4, echo=FALSE, message=FALSE, echo=FALSE, message=FALSE}
+```{r fig.width=6, fig.height=4, echo=FALSE, message=FALSE, echo=FALSE, message=FALSE, eval=includeFigs}
 figure4(dLAIlitter)
@@
 
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
 figure5(dLAIlitter)
@@
 
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
 figure6(ba)
@@
 
-```{r fig.width=10, fig.height=8, echo=FALSE, message=FALSE}
+```{r fig.width=10, fig.height=8, echo=FALSE, message=FALSE, eval=includeFigs}
 figure7(subset(facegap_cloudy_byring, Date < maxdate),
@@
 
-```{r fig.width=5, fig.height=8, echo=FALSE, message=FALSE}
+```{r fig.width=5, fig.height=8, echo=FALSE, message=FALSE, eval=includeFigs}
 figureSI1(litring, subset(facegap_cloudy_byring, Date < maxdate))
@@
 
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
 figureSI2(flatcan_byring)
@@
 
-```{r fig.width=8, fig.height=5, echo=FALSE, message=FALSE}
+```{r fig.width=8, fig.height=5, echo=FALSE, message=FALSE, eval=includeFigs}
 figureSI3(subset(facegap_cloudy_byring, Date < maxdate), 

### 2015-05-21 R/figures.R

@@ -93,7 +93,7 @@ figure1 <- function(df, ramp){
 
 figure2 <- function(df){
   
-  par(mar=c(5,5,2,2), cex.lab=1.2,xaxs="i", yaxs="i")
+  par(mar=c(5,5,2,2), cex.lab=1.2,xaxs="i", yaxs="i", tcl=0.2, las=1)
   with(df, plot(BA, LAI.mean, pch=19, cex=1.2, col=my_ringcols(),
                 xlab=expression(Basal~area~~(m^2~ha^-1)),
                 ylab=expression(bar(italic(L))~~(m^2~m^-2)),
@@ -147,7 +147,7 @@ figure4 <- function(df){
   dfup <- df[df$LAIchange == "increasing",]
   dfdown <- df[df$LAIchange == "decreasing",]
   
-  par(mar=c(5,5,2,2), cex.lab=1.1, cex.axis=0.9, las=1, xaxs="i", yaxs="i")
+  par(mar=c(5,5,2,2), cex.lab=1.1, cex.axis=0.9, las=1, xaxs="i", yaxs="i", tcl=0.2)
   with(dfup, plot(X, Y, pch=c(19,21)[treatment],
                col=Cols[LAIchange],
                xlab=expression(Leaf~litter~production~~(m^2~m^-2~mon^-1)),
@@ -201,7 +201,7 @@ figure5 <- function(dLAIlitter){
 
 figure6 <-  function(df){
 
-  par(mar=c(5,5,2,2), cex.lab=1.2, xaxs="i", yaxs="i", mfrow=c(1,2))
+  par(mar=c(5,5,2,2), cex.lab=1.2, xaxs="i", yaxs="i", mfrow=c(1,2), tcl=0.2)
   
   LLs <- seq(0.8,2,by=0.1)
   
@@ -262,7 +262,7 @@ figure7 <- function(df, facesoilwater, faceraindaily, airt, kgam=18){
   plotlabel("(a)","topleft", inset.x=xin)
   
   # panel b
-  par(mar=c(0,7,1.5,6))
+  par(mar=c(0,7,1.5,6), tcl=0.2)
   
  
   df$Time <- as.numeric(df$Date - min(df$Date))
@@ -282,7 +282,7 @@ figure7 <- function(df, facesoilwater, faceraindaily, airt, kgam=18){
            y0=0.014, y1=0.014, lwd=2) 
   
   # panel c
-  par(mar=c(1.5,7,1.5,6))
[…]

### 2015-05-21 R/functions-figures.R

@@ -39,7 +39,7 @@ plotlabel <- function(txt, where, inset=0.08, inset.x=inset, inset.y=inset,...){
   if(grepl("bottom",where))y <- u[3] + inset.y*(u[4]-u[3])
   if(grepl("top",where))y <- u[4] - inset.y*(u[4]-u[3])
   
-  text(x,y,txt,...)
+  text(x,y,txt,font=2,...)
 }
 
 

### 2015-05-25 manuscript.Rmd
MT minor edits

@@
 Ecosystem models generally predict that *L* increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy vegetation-atmosphere fluxes as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
+This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes between vegetation and atmosphere as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
 Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
@@
 Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
-Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
+Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
 Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
@@
 We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
-The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration [long-term mean of 1350mm; @Zomer2008] and mean annual precipitation of 800mm.
-The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability.
+The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration [long-term mean of 1350 mm; @Zomer2008] and mean annual precipitation of 800 mm.
+The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L* and its components (litter fall and production) in response to water availability.
 We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation.
@@
 The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
-An estimate of potential evapotranspiration (PET) at the site (1350mm) was from the Global-PET database [http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
+An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
 0.59.
@@
 ## Leaf litter production
-In each ring, 8 litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor.
+In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor.
 Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
-Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
+Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
 We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.
@@
 In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
-Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct.
-2012, and automatically uploaded every day to a data management system.
+Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 October 2012, and automatically uploaded every day to a data management system.
 At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height.
-All PAR data were averaged in 30 min time steps.
-At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca.
-5m above the canopy.
+All PAR data were averaged in 30-min time steps.
+At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca.
+5 m above the canopy.
 For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
@@
 
-where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000].
+where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
 Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006].
@@
 The camera was aligned with the north-south axis and set at 1.25 m height.
-Camera lens was pointed directly upwards and levelled using a three axis bubble level.
+The camera lens was pointed directly upwards and leveled using a three axis bubble level.
 We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan).
@@
 [-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^=`r f(glance(lmphoto)$r.squared)`, RMSE=`r f(glance(lmphoto)$sigma)`) (Fig.
+Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig.
 S2).
@@
 \
-An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in Sept.
-2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015].
+An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015].
 This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
@@
 7).
-These events occurred in spring and summer, with no clear seasonal cycle (Oct.
-2012, Jan.
-2013, Dec.
-2013 and Jan.
-2015).
+These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
 The dynamics of $L_\tau$ from the sensor record matched these events observed in the field quite well (Fig.
@@
 All four leaf flushing events corresponded to significant positive dL/dt.
-One brief significant positive dL/dt (June 2013) did not correspond to an actual leaf flushing event, but it was small in magnitude.
+One brief significant positive dL/dt (June 2013) did not correspond to an observed leaf flushing event, but it was small in magnitude.
 It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig.
@@
 ## Response of *L* to elevated [CO~2~]
-We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from Nov.
-2012 to Jan.
-2013.
-We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability.
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013.
+We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability.
 We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011].
@@
 \
-This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007], for a number of reasons.
+This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons.
 First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*.
@@
 We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig.
-6) (an increase of `r f(100*with(LLtest,estimate2/estimate1 - 1))`%).
+6) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
 Further years of observation are needed to determine if such a difference is sustained.
@@
 5), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
-There is mixed evidence from other FACE sites on the relatinship between LL and eC~a~.
+There is mixed evidence from other FACE sites on the relationship between LL and eC~a~.
 Norby et al.
@@
 At the same site, Warren et al.
-[-@Warren2011] found increased senescence in response to a combined drought and heat wave.
+[-@Warren2011] found increased senescence in the eC~a~ treatment in response to a combined drought and heat wave.
 McCarthy et al.
@@
 S3), because longer path lengths in the winter result in lower overall transmittance.
-Using diffuse PAR only avoids having to correct for these path length effects.
+Using diffuse PAR alone avoids having to correct for these path length effects.
+
 \
@@
 The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Fig.
-S2), suggesting both diffuse transmittanc and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
+S2), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
 
@@
 4 and 5), indicating replacement of old leaves by young leaves.
-Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of Dec.
-2013.
+Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of December 2013.
 This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
@@
 \
-Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response.
+Although the response of *L* to water availability is qualitatively well known, current-generation phenology models fall short in incorporating this response.
 In a comparison of six leading ecosystem and global vegetation models,  Powell et al.
 [-@Powell2013] showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model.
-Simulations from De Kauwe et al.
-[-@DeKauwe2014] at ambient and elevated C~a~ also showed little response to inter-annual variation in rainfall (pers.
+Simulations at the EucFACE site using the set of ecosystem models in De Kauwe et al.
+[-@DeKauwe2014] also showed little response to inter-annual variation in rainfall (pers.
 comm.
-Martin De Kauwe).
+Martin De Kauwe; Medlyn et al.
+*in prep*).
 In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g.
-@Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool.
+@Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and then allocating some fraction to the leaf pool.
 Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
@@
 (a) Target [CO~2~] (C~a~) relative to the ambient C~a~ treatment at the EucFACE.
-The elevated C~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in Feb.
-2013.
+The elevated C~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in February 2013.
 (b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions (F~diff~ > 0.98) only.
@@
 Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~ (+150 ppm).
-Note steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
+Note the steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
 The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate *L* estimates to litter production (see Methods).
@@
 Four of the five significant positive increases corresponded to approximate dates of commencement of leaf flushing (horizontal lines) observed in the field (except the third brief significant increase).
-(c) Soil water content estimated from over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).
+(c) Time course of soil water content measured over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).
 (d) Average daily air temperature with a fitted smoother.
@@
 (a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
-(b) Mean litter production for each ring.
+(b) Mean litter production expressed in terms of *L* for each ring.
 The confidence interval was calculated based on the 8 litter traps in each ring.

### 2015-05-26 manuscript.Rmd
caption fixes

@@
 The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model).
-Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~ (+150 ppm).
+Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~.
 Note the steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
@@
 The ratio of mean $L_\tau$ to annual litter production gives the mean leaf life span.
-Grey lines indicate leaf lifespan isolines.
-(b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p `r pval(LLtest$p.value)`).
+Grey lines indicate leaf lifespan isolines (in years).
+(b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (t-test, p `r pval(LLtest$p.value)`).
 

### 2015-05-26 manuscript_calcs.R
maxdate needs to be set after loading Rdata

@@ -1,7 +1,7 @@
 
 #' This script is called by Manuscript.Rmd, and includes additional calculations 
 #' used in the manuscript.
-
+#
 
 source("R/utils.R")  # Library()
 
@@ -14,6 +14,9 @@ Library(reporttools)
 
 # Load analysis cache.
 load("cache/lai_workspace.RData")
+
+# Trim data to this date (note: has to be consistent with make_figures.R!) 
+maxdate <- as.Date("2015-3-1")
 
 # More packages, functions, etc.
 source("load.R")
@@ -40,7 +43,7 @@ f <- function(x, digits=2)sprintf(x, fmt=paste0("%.",digits,"f"))
 .maxSD <- 0.03
 
 # Main results
-df <- facegap_cloudy_byring
+df <- subset(facegap_cloudy_byring, Date < maxdate)
 
 # Calibration constant to litter fall.
 caliblm <- calibrateToDrought(df)$lmfit

### 2015-08-05 manuscript.Rmd
Replaced *L* with LAI

@@
 
-Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
-Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified.
+Canopy leaf area, quantified by the leaf area index (LAI), is a crucial driver of forest productivity, water use, and energy balance.
+Because LAI responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified.
 We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
-We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, *L* should increase with elevated [CO~2~].
-We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production.
-Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe.
-We found that *L* varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period, but did not show a clear seasonal pattern.
-Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
-Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
-Leaf lifespan, estimated from mean *L* and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs.
+We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, LAI should increase with elevated [CO~2~].
+We estimated LAI from diffuse canopy transmittance, and measured monthly leaf litter production.
+Contrary to expectation, LAI did not respond to elevated [CO~2~] in this timeframe.
+We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period, but did not show a clear seasonal pattern.
+Instead, dynamics of LAI showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
+Leaf litter production was correlated to changes in LAI, both during periods of decreasing LAI (when no leaf growth occurred), and during periods of increasing LAI (active shedding of old foliage when new leaf growth occurred).
+Leaf lifespan, estimated from mean LAI and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs.
 `r f(LLtest$estimate1)` yrs; p `r pval(LLtest$p.value)`).
-Our main finding that *L* was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of *L* commonly predicted by many ecosystem models.
+Our main finding that LAI was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of LAI commonly predicted by many ecosystem models.
 
@@
 
-Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
-*L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that *L* increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes between vegetation and atmosphere as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
-Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
+Leaf area index (LAI) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
+LAI is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
+Ecosystem models generally predict that LAI increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+This potential change in LAI is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes between vegetation and atmosphere as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
+Despite its importance, little empirical evidence is available on the response of LAI to eC~a~ in mature forests [@Norby2011].
 
@@
 Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
-Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
-Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
-Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased *L* in response to eC~a~ [@McCarthy2007; @Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
-Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
+Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state LAI to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
+Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in LAI.
+Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased LAI in response to eC~a~ [@McCarthy2007; @Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
+Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment LAI and the responsiveness to eC~a~, leading to the prediction that stands with low LAI will show a larger response to eC~a~.
 
 \
-In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
-These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
+In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect LAI [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
+These ‘water savings’ can be expected to lead to an increase in LAI in water-limited ecosystems, consistent with a tight coupling of LAI with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
 In support of this prediction, Donohue et al.
-[-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~.
-Based on the same water-savings principle, we predicted that *L* should increase under eC~a~ in water-limited forests.
+[-@Donohue2013] attributed the observed global increase in LAI in arid and semi-arid areas over the last 30 years to rising C~a~.
+Based on the same water-savings principle, we predicted that LAI should increase under eC~a~ in water-limited forests.
 
 \
-Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and photo period-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g.
-@Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
-The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
-However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
-Understanding dynamics of *L* in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
+Most previous studies on LAI responses to environmental drivers have focused on the seasonal, mostly temperature and photo period-driven, timing of LAI dynamics observed in N-hemisphere forests [e.g.
+@Richardson2012; @Savoy2015] whereas LAI responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
+The strong relationship between LAI and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
+However, local LAI is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
+Understanding dynamics of LAI in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
 
 \
-We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
+We estimated LAI from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
 The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration [long-term mean of 1350 mm; @Zomer2008] and mean annual precipitation of 800 mm.
-The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L* and its components (litter fall and production) in response to water availability.
-We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation.
-In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a positive response of *L* to eC~a~ [cf.
+The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of LAI and its components (litter fall and production) in response to water availability.
+We hypothesized that eC~a~ would result in an increase in LAI, particularly during periods of water limitation.
+In addition, we expected that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a positive response of LAI to eC~a~ [cf.
 @Norby2011], either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
-Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.
+Our second objective was to quantify the temporal dynamics of LAI, in particular in relation to fluctuating water availability.
 
@@
 ## Terminology
-In the following, we use the term 'leaf' area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e.
+In the following, we use the term 'leaf' area index (LAI) although the canopy and FACE rings contain non-leaf shading elements, i.e.
 woody stems and infrastructure.
-This avoids the need to switch between different terminologies when comparing temporal changes in *L* to litter production, and absolute values between different methods.
-The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.
+This avoids the need to switch between different terminologies when comparing temporal changes in LAI to litter production, and absolute values between different methods.
+The temporal changes in LAI are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.
 
@@
 Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
-Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
+Litter production was converted to units of LAI using an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
 We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.
 
-## *L* from diffuse transmittance
+## LAI from diffuse transmittance
 In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
@@
 5 m above the canopy.
-For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
+For determination of LAI, we used only highly diffuse conditions, where F~diff~ > 0.98.
 We further trimmed the data by excluding low solar e levation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles.
@@
 \
-We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following Campbell and Norman [-@Campbell2000].
+We then estimated leaf area index (LAI) from $\tau_d$ ($L_\tau$), following Campbell and Norman [-@Campbell2000].
 We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
-We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below.
-Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).
+We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the LAI estimate to leaf litter production as further detailed below.
+Diffuse transmittance of a canopy is a function of the LAI by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).
 
@@
 Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006].
-We solved numerically for *L* in Eq.
+We solved numerically for LAI in Eq.
 2, using a smooth integration over $\theta$, yielding $L_\tau$.
@@
 1).
-During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
+During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but LAI decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
 1).
@@
 A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
-Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance.
+Cumulative LAI leaf litter production ($P_L$, expressed in units of LAI) should match the observed decrease in LAI estimated from canopy transmittance.
 We calibrated Eq.
@@
 
-We took advantage of the detailed timeseries of *L* estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al.
+We took advantage of the detailed timeseries of LAI estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al.
 [-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests.
@@
 We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.
-*L* from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq.
+LAI from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq.
 4).
@@
 S2).
-We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig.
+We also estimated the extinction coefficient that minimized the difference between the two estimates of LAI (Fig.
 S2), giving *k* = `r f(kopt)`.
-These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
+These comparisons are encouraging because the photos measure LAI near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
 2), weighted towards the zenith.
@@
 ## Data analysis
-We used generalized additive models (GAM) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015].
+We used generalized additive models (GAM) for all analyses of LAI and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015].
 A GAM is a semi-parametric fitting method that allows the estimation of smoothed trends in the data, including confidence intervals.
 In all fitted GAMs, we used a cubic regression spline.
-For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed *L* dynamics.
+For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed LAI dynamics.
 Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
-Random effects were included to account for the grouping of data within FACE rings, except when we analysed average *L* across all rings (Fig.
+Random effects were included to account for the grouping of data within FACE rings, except when we analysed average LAI across all rings (Fig.
 7).
@@
 \
-For comparison between *L* from photos and PAR transmittance, we used linear regression.
+For comparison between LAI from photos and PAR transmittance, we used linear regression.
 To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
@@
 We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig.
-3, to yield a '*L* anomaly'.
+3, to yield a 'LAI anomaly'.
 This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.
@@
 An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015].
-This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
+This rapid decrease in LAI was evident in a quick increase in $\tau_d$ (Fig.
 1).
@@
 \
-We found no differences in the *L* anomaly between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any point in time, Fig.
+We found no differences in the LAI anomaly between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any point in time, Fig.
 3).
 A brief exception was the period where one of the rings was affected by an insect outbreak (see above).
-A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the *L* anomaly (p `r pval(LanomAOV[[3]][1])`).
+A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the LAI anomaly (p `r pval(LanomAOV[[3]][1])`).
 But what difference between C~a~ treatments could we have reasonably detected?
-The average half-width of the confidence interval on the *L* anomaly (Fig.
+The average half-width of the confidence interval on the LAI anomaly (Fig.
 4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean *L* of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
+This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean LAI of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
 
@@
 4).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where LAI was decreasing.
 During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
@@
 7a), as estimated from a generalized additive model fitted to the average $L_\tau$ across the six FACE rings.
-From the fitted model, we estimated the numerical derivative of *L* (dL/dt), as well as its confidence interval.
+From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval.
 All four leaf flushing events corresponded to significant positive dL/dt.
@@
 > 0.2 m^3^ m^-3^.
-Two periods of sustained decrease in *L* (red lines in Fig.
+Two periods of sustained decrease in LAI (red lines in Fig.
 7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.
@@
 
-## Response of *L* to elevated [CO~2~]
-We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013.
-We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability.
-We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011].
+## Response of LAI to elevated [CO~2~]
+We report on `r nummon` months of continuous leaf area index (LAI) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013.
+We predicted that LAI would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability.
+We also expected that this eucalypt woodland, with its inherently low LAI [@Whitehead2004], would be even more responsive to eC~a~ than higher-LAI and more closed-canopy stands [@Norby2011].
 Counter to our expectations but similar to Norby et al.
-[-@Norby2003], we found no significant effect of elevated C~a~ on *L*.
+[-@Norby2003], we found no significant effect of elevated C~a~ on LAI.
 Because estimated leaf lifespan averaged `r round(mean(ba$LL),1)` yrs (Fig.
 6), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
-It could be argued that the change in *L* in response to eC~a~ was too small to be detected.
-Natural variation in *L* among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
+It could be argued that the change in LAI in response to eC~a~ was too small to be detected.
+Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
 To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate.
-Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*.
+Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in LAI.
 Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.
@@
 This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*.
-They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
-The mean *L* at the EucFACE was ca.
-`r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
+First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low LAI stands were more responsive to eC~a~ than stands with high LAI.
+They argued that high LAI canopies would not benefit from further increases in LAI, because it would lead to only very small increases in total absorbed radiation.
+The mean LAI at the EucFACE was ca.
+`r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest LAI [@Iio2013], and the lowest LAI of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
 [-@Iio2013].
-Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in *L* under eC~a~.
+Based on this LAI estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in LAI under eC~a~.
 In contrast to other forest FACE sites, the EucFACE is very strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
-Hence, it may be too early to draw general conclusions across FACE sites in terms of *L* response.
+Hence, it may be too early to draw general conclusions across FACE sites in terms of LAI response.
 
 \
-A possible explanation for the lack of a *L* response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~.
+A possible explanation for the lack of a LAI response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~.
 In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
@@
 Further years of observation are needed to determine if such a difference is sustained.
-Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
+Because we found no difference in LAI with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
 To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
@@
 \
-Despite a lack of response of canopy *L*, we did find a significantly longer leaf lifespan (LL) in eC~a~.
-We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
+Despite a lack of response of canopy LAI, we did find a significantly longer leaf lifespan (LL) in eC~a~.
+We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean LAI and total accumulated leaf litter (Fig.
 6).
@@
 McCarthy et al.
-[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of *L* by eC~a~.
+[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of LAI by eC~a~.
 Taylor et al.
 [-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
-Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ at the PopFACE, leading to similar *L* in ambient C~a~ despite higher productivity.
+Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ at the PopFACE, leading to similar LAI in ambient C~a~ despite higher productivity.
 Li et al.
@@
 
-## Continuous measurement of *L*
+## Continuous measurement of LAI
 Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
-Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because leaf lifespan (LL) is unknown (instead, we used *L* and litter production to estimate LL).
+Unlike in deciduous canopies, litter production cannot be used to estimate LAI, because leaf lifespan (LL) is unknown (instead, we used LAI and litter production to estimate LL).
 We cannot not rely either on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
-tereticornis*, and would not be valuable anyway since *L* is temporally highly variable (Figs.
+tereticornis*, and would not be valuable anyway since LAI is temporally highly variable (Figs.
 1, 7).
-We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
+We found that indirect estimates of LAI based on canopy transmittance were very reliable because relatively small temporal changes in LAI could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
 1, 5, 7).
@@
 \
-We used the opportunity provided by the high quality timeseries of *L* to test a simple photography-based method suggested by Macfarlane et al.
+We used the opportunity provided by the high quality timeseries of LAI to test a simple photography-based method suggested by Macfarlane et al.
 [-@Macfarlane2007].
 The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Fig.
-S2), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
+S2), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate LAI.
 
-## Litter production and dynamics of *L*
-We found a correlation between *L* dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing *L* (Fig.
+## Litter production and dynamics of LAI
+We found a correlation between LAI dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing LAI (Fig.
 4).
-As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production.
-During periods of increasing *L*, litter fall increased sharply (Figs.
+As expected, during periods of decreasing LAI, the change in LAI was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of LAI against litter production.
+During periods of increasing LAI, litter fall increased sharply (Figs.
 4 and 5), indicating replacement of old leaves by young leaves.
 Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of December 2013.
-This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
+This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net LAI increase, but with a lag between litter fall and production of about 2 months.
 Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
@@
 \
-Although the response of *L* to water availability is qualitatively well known, current-generation phenology models fall short in incorporating this response.
+Although the response of LAI to water availability is qualitatively well known, current-generation phenology models fall short in incorporating this response.
 In a comparison of six leading ecosystem and global vegetation models,  Powell et al.
-[-@Powell2013] showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model.
+[-@Powell2013] showed that the steady decrease in LAI in an Amazon drought experiment could be matched by only one model.
 Simulations at the EucFACE site using the set of ecosystem models in De Kauwe et al.
@@
 *in prep*).
-In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g.
-@Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and then allocating some fraction to the leaf pool.
-Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
-We argue that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests.
+In most models  that provide prognostic estimates of LAI [i.e., excluding models that take LAI as an input parameter; e.g.
+@Kala2013], temporal changes in LAI are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and then allocating some fraction to the leaf pool.
+Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of LAI in response to environmental drivers such as eC~a~.
+We argue that this algorithm is not well suited to explain the dynamic nature of LAI in response to water availability observed in evergreen broad-leaved forests.
 In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability, as we have demonstrated that both processes are highly dynamic.
@@
 Note the steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
-The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate *L* estimates to litter production (see Methods).
+The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate LAI estimates to litter production (see Methods).
 
@@
 **Figure 2**.
-Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
-This relationship was used to calculate the *L* anomaly in Fig.
-3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.
+Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
+This relationship was used to calculate the LAI anomaly in Fig.
+3, as the difference between LAI predicted from this regression and the actual $L_\tau$ at any given time.
 Red symbols are elevated C~a~, blue symbols ambient C~a~.
@@
 
-**Figure 3**.Time series of *L* after correcting for plot-to-plot differences in basal area (BA).
-At every time point, the *L* anomaly was calculated as $L_\tau$ minus *L* fitted from the relationship with BA shown in Fig.
+**Figure 3**.Time series of LAI after correcting for plot-to-plot differences in basal area (BA).
+At every time point, the LAI anomaly was calculated as $L_\tau$ minus LAI fitted from the relationship with BA shown in Fig.
 2.
@@
 **Figure 4**.
-Comparison of absolute changes in *L* ($\Delta L$) to litter production (in units of *L*), calculated for each litter collection period (approx.
+Comparison of absolute changes in LAI ($\Delta L$) to litter production (in units of LAI), calculated for each litter collection period (approx.
 monthly).
@@
 Timeseries of litter production and estimated gross leaf production rate.
-Leaf production rate was estimated as the sum of the net change in *L* and litter production (converted to a per area basis using specific leaf area estimates, see Methods).
+Leaf production rate was estimated as the sum of the net change in LAI and litter production (converted to a per area basis using specific leaf area estimates, see Methods).
 Error bars are omitted for clarity.
@@
 **Figure 7**.
-Dynamics of *L* in relation to environmental drivers.
+Dynamics of LAI in relation to environmental drivers.
 (a) Smoothed time series of $L_\tau$ averaged across all six FACE rings.
-(b) From panel (a), we calculated the daily change in *L* (dL/dt), including the 95% confidence interval for the derivative of the time series.
+(b) From panel (a), we calculated the daily change in LAI (dL/dt), including the 95% confidence interval for the derivative of the time series.
 Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05).
@@
 (a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
-(b) Mean litter production expressed in terms of *L* for each ring.
+(b) Mean litter production expressed in terms of LAI for each ring.
 The confidence interval was calculated based on the 8 litter traps in each ring.
@@
 **Figure S2**.
-Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
+Correlation between LAI estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
 Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.

### 2015-08-05 manuscript.Rmd
Moved supporting information to separate Rmd

@@
 S1).
-
-## Estimation of leaf area index from digital photography
-
-We took advantage of the detailed timeseries of LAI estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al.
-[-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests.
-From October 2012 to October 2013, we took ca.
-monthly canopy photographs in each FACE ring.
-Photographs were taken in the morning on cloudy days.
-We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring.
-Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded.
-The camera was aligned with the north-south axis and set at 1.25 m height.
-The camera lens was pointed directly upwards and leveled using a three axis bubble level.
-We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan).
-To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens.
-Following recommendations by Jonckheere et al.
-[-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos.
-We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.
-LAI from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq.
-4).
-
-\
-(4) 
-$$L_p = -log(\tau_p)/k$$
-
-where $\tau_p$ canopy transmission estimated from canopy photos.
-We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al.
-[-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig.
-S2).
-We also estimated the extinction coefficient that minimized the difference between the two estimates of LAI (Fig.
-S2), giving *k* = `r f(kopt)`.
-These comparisons are encouraging because the photos measure LAI near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
-2), weighted towards the zenith.
 
@@
 
-```{r fig.width=5, fig.height=8, echo=FALSE, message=FALSE, eval=includeFigs}
-figureSI1(litring, subset(facegap_cloudy_byring, Date < maxdate))
-```
-
-**Figure S1**.
-Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately.
-Blue colours are ambient, red colours elevated C~a~.
-(a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
-(b) Mean litter production expressed in terms of LAI for each ring.
-The confidence interval was calculated based on the 8 litter traps in each ring.
-Note the overlapping confidence intervals for all rings.
-
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
-figureSI2(flatcan_byring)
-```
-
-**Figure S2**.
-Correlation between LAI estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
-Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
-(a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
-[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
-
-```{r fig.width=8, fig.height=5, echo=FALSE, message=FALSE, eval=includeFigs}
-figureSI3(subset(facegap_cloudy_byring, Date < maxdate), 
-          subset(facegap_all_byring, Date < maxdate))
-```
-
-**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
-Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals).
-Note the much lower $\tau$ during southern hemisphere winter when using all data compared to diffuse only.
-This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.
-
 # References {.unnumbered}

### 2015-08-05 manuscript_supportinginfo.Rmd
Moved supporting information to separate Rmd

---
title: Supporting Information
author: ''
date: ''
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

## Supporting Figures

```{r fig.width=4, fig.height=6, echo=FALSE, message=FALSE, eval=TRUE}
figureSI1(litring, subset(facegap_cloudy_byring, Date < maxdate))
```

**Figure S1**. Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately. Blue colours are ambient, red colours elevated C~a~. (a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013). (b) Mean litter production expressed in terms of LAI for each ring. The confidence interval was calculated based on the 8 litter traps in each ring. Note the overlapping confidence intervals for all rings.

\
```{r fig.width=8, fig.height=5, echo=FALSE, message=FALSE, eval=TRUE}
figureSI3(subset(facegap_cloudy_byring, Date < maxdate), 
          subset(facegap_all_byring, Date < maxdate))
```

**Figure S2.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering. Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals). Note the much lower $\tau$ during southern hemisphere winter when using all data compared to diffuse only. This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.

## Notes S1. Estimation of leaf area index from digital photography

We took advantage of the detailed timeseries of LAI estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al. [-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests. From October 2012 to October 2013, we took ca. monthly canopy photographs in each FACE ring. Photographs were taken in the morning on cloudy days. We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring. Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded. The camera was aligned with the north-south axis and set at 1.25 m height.  The camera lens was pointed directly upwards and leveled using a three axis bubble level. We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan). To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens. Following recommendations by Jonckheere et al. [-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos. We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold. LAI from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq. 4).

\
(4) 
$$L_p = -log(\tau_p)/k$$

  where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig. S2). We also estimated the extinction coefficient that minimized the difference between the two estimates of LAI (Fig. S2), giving *k* = `r f(kopt)`. These comparisons are encouraging because the photos measure LAI near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq. 2), weighted towards the zenith.

\
```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=TRUE}
figureSI2(flatcan_byring)
```

**Figure S2**. Correlation between LAI estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~). Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates. (a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al. [-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).

### 2015-08-05 manuscript_calcs.R
Added Table 1 with new analysis results

@@ -68,6 +68,15 @@ df$Time <- as.factor(df$Date - min(df$Date))
 m3 <- lmer(LAIanomaly ~ treatment*Time + (1|Ring), data=df)
 LanomAOV <- Anova(m3)
 
+# significance of LAI when using BA as actual covariate
+df2 <- merge(df, ba[,c("Ring","BA")])
+df2 <- subset(df2, Date < as.Date("2014-9-1"))
+m4 <- lmer(LAI ~ treatment * as.factor(Date) + BA + (1|Ring), data=df2)
+lmer_LAI <- Anova(m4, test="F")
+rownames(lmer_LAI) <- c("C~a~","Date","BA","C~a~ x Date")
+colnames(lmer_LAI) <- c("F","numDF","denDF","p-value")
+
+
 # Average CI width in Fig 4.
 fit <- smoothplot(Date, LAIanomaly, g=treatment, R="Ring", 
                   data=facegap_cloudy_byring, plot=FALSE, kgam=.gambase)
@@ -98,4 +107,6 @@ lmcanphot <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)
 flatcan_byring2 <- agg_flatcan(flatcan, by="Ring", k=kopt)
 flatcan_byring2 <- add_PARLAI_to_flatcan(facegap_cloudy_byring,flatcan_byring2)
 lmcanphot2 <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring2)
+
+
 

### 2015-08-06 R/figures-other.R
Two LAI timeseries plots useful for demoing

@@ -1,3 +1,34 @@
+figure3b <- function(df,
+                    xlim=NULL, ylim=NULL,
+                    legend=TRUE,
+                    ylab=expression(italic(L)~(m^2~m^-2)),
+                    cex.lab=1.1, cex.axis=0.8, cex.legend=0.7,
+                    legendwhere="topleft",
+                    setpar=TRUE,axisline=3,
+                    horlines=TRUE,
+                    greyrect=FALSE,
+                    addpoints=TRUE){
+  
+  if(setpar)par(cex.axis=cex.axis, mar=c(3,5,2,5), las=1, cex.lab=1.2, yaxs="i", tcl=0.2)
+  par(cex.lab=cex.lab)
+  palette(my_co2cols())
+  
+  if(is.null(xlim))xlim <- with(df, c(min(Date)-15, max(Date)+15))
+  if(is.null(ylim))ylim <- c(0,2.8)
+  
+  smoothplot(Date, LAI, g=treatment, R="Ring", ylim=ylim, xlim=xlim, 
+             ylab=ylab, xlab="",
+             data=df, kgam=18, axes=FALSE,
+             polycolor=c(alpha("royalblue",0.4),alpha("pink",0.4)))
+  
+  l <- legend("topleft", c("Ambient","Elevated"), title=expression(italic(C)[a]~treatment), 
+              fill=my_co2cols(), bty="n", cex=cex.legend)
+  axis(2) #, at=seq(-0.75,0.75,by=0.25))
+  box()
+  
+  timeseries_axis()  
+}
+
 
 
 #-----------------------------------------------------------------------------------------#
@@ -150,15 +181,17 @@ figure_timeseries <- function(df,flatcan_byCO2,
   
   adddata <- match.arg(adddata)
   
+  df <- df[order(df$Date),]
+  
   if(setpar)par(cex.axis=cex.axis, mar=c(3,5,2,5), las=1, cex.lab=1.2, yaxs="i")
   par(cex.lab=cex.lab)
   palette(my_co2cols())
   
   if(is.null(xlim))xlim <- with(df, c(min(Date)-15, max(Date)+15))
-  if(is.null(ylim))ylim <- with(df, c(0, 1.08*max(LAI.mean)))
+  if(is.null(ylim))ylim <- with(df, c(0, 1.08*max(LAI)))
   
   with(subset(df, treatment == "ambient"),
-       plot(Date, LAI.mean, col=palette()[1], type='l', lwd=2, 
[…]

### 2015-08-06 make_figures.R

@@ -51,3 +51,13 @@ to.pdf(figureSI3(subset(facegap_cloudy_byring, Date < maxdate), subset(facegap_a
        width=8, height=5)
 
 
+
+
+to.pdf(figure_timeseries(facegap_cloudy_byring, flatcan_byCO2),
+       filename="output/LAI_timeseries.pdf",
+       width=10, height=8)
+
+to.pdf(figure3b(facegap_cloudy_byring),
+       filename="output/figures/Figure3b.pdf",
+       width=8, height=4)
+

### 2015-08-06 manuscript.Rmd

@@
 \
-We then estimated leaf area index (LAI) from $\tau_d$ ($L_\tau$), following Campbell and Norman [-@Campbell2000].
+We then estimated leaf area index (LAI) from $\tau_d$ ($LAI_\tau$), following Campbell and Norman [-@Campbell2000].
 We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
@@
 We solved numerically for LAI in Eq.
-2, using a smooth integration over $\theta$, yielding $L_\tau$.
+2, using a smooth integration over $\theta$, yielding $LAI_\tau$.
 
@@
 To overcome violations in the assumptions underlying Eq.
-2, we calibrated $L_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
+2, we calibrated $LAI_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
 1).
@@
 
-where *c* is a calibration parameter, $L_{\tau,t=0}$ is $L_\tau$ at the start of the drought period.
+where *c* is a calibration parameter, $L_{\tau,t=0}$ is $LAI_\tau$ at the start of the drought period.
 We used linear regression without intercept to estimate *c* in Eq.
 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`).
-It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology).
+It should be stressed that even after calibration with litter production, $LAI_\tau$ still includes stem and infrastructure area (see Terminology).
 However, this approximately constant area cancels out in the left-hand side of Eq.
@@
 S1).
+
+\
+We used the opportunity provided by the high quality timeseries of LAI to test a simple photography-based method suggested by Macfarlane et al.
+[-@Macfarlane2007].
+The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Notes S1, Fig.
+S3), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate LAI.
 
@@
 For comparison between LAI from photos and PAR transmittance, we used linear regression.
-To test for effects of C~a~ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
+To test for effects of C~a~ treatment on $LAI_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
 
 \
-To assess effects of eC~a~ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
-The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig.
+To assess effects of eC~a~ on $LAI_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
+The average $LAI_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig.
 3).
 Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.
-We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig.
+We then calculated the deviation of actual $LAI_\tau$ from the fitted relationship with BA shown in Fig.
 3, to yield a 'LAI anomaly'.
@@
 \
-The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected.
-We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to that shown for $\tau_d$ in Fig.
-1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$).
+The dynamics of $LAI_\tau$ were compared with litter fall, by estimating the absolute change in $LAI_\tau$ for every FACE ring, for every period when litter was collected.
+We did this by fitting a generalized additive model to $LAI_\tau$ for each ring separately (similar to that shown for $\tau_d$ in Fig.
+1), and calculating the change in $LAI_\tau$ for each period ($\Delta L_\tau$).
 Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig.
@@
 \
-We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $L_\tau$.
-For each litter collection period, the net change in $L_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries.
-Gross leaf production was then calculated as the sum of the net change in $L_\tau$ and the amount of leaf litter produced during that period.
+We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $LAI_\tau$.
+For each litter collection period, the net change in $LAI_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries.
+Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period.
 The temporal dynamics of litter and leaf production are shown in Fig.
@@
 \
-Next, we related P~L~ to average $L_\tau$ (during the litter collection period) across the six FACE rings (Fig.
+Next, we related P~L~ to average $LAI_\tau$ (during the litter collection period) across the six FACE rings (Fig.
 6a).
-If leaf lifespan is constant, P~L~ would be linearly related to $L_\tau$.
-Instead, we found that in eC~a~, P~L~ was lower at a given $L_\tau$ as compared to aC~a~ (Fig.
+If leaf lifespan is constant, P~L~ would be linearly related to $LAI_\tau$.
+Instead, we found that in eC~a~, P~L~ was lower at a given $LAI_\tau$ as compared to aC~a~ (Fig.
 6a).
 We estimated mean leaf lifespan from Fig.
-6a as the ratio of $L_\tau$ to annual total P~L~.
+6a as the ratio of $LAI_\tau$ to annual total P~L~.
 We found a significant (p `r pval(LLtest$p.value)`) increase in leaf lifespan (LL) with elevated C~a~ (Fig.
@@
 Finally, we quantified the relationship between timing of leaf flushing and environmental drivers.
-During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $L_\tau$ were available (Fig.
+During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available (Fig.
 7).
 These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
-The dynamics of $L_\tau$ from the sensor record matched these events observed in the field quite well (Fig.
-7a), as estimated from a generalized additive model fitted to the average $L_\tau$ across the six FACE rings.
+The dynamics of $LAI_\tau$ from the sensor record matched these events observed in the field quite well (Fig.
+7a), as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings.
 From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval.
@@
 
-\
-We used the opportunity provided by the high quality timeseries of LAI to test a simple photography-based method suggested by Macfarlane et al.
-[-@Macfarlane2007].
-The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Fig.
-S2), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate LAI.
-
 ## Litter production and dynamics of LAI
@@
 
+# Tables {.unnumbered}
+
+**Table 1**.
+Results of a mixed-effects model of LAI with C~a~, Date and pre-treatment basal area (BA) as predictors.
+The model accounts for repeated measures by using FACE ring as the random effect.
+Sample size of LAI per ring was `r table(df$Ring)[1]`.
+
+```{r results='asis', echo=FALSE}
+lmer_LAI[,"p-value"] <- formatPval(lmer_LAI[,"p-value"])
+
+kable(lmer_LAI)
+```
+
 # Figures {.unnumbered}
@@
 **Figure 2**.
-Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
+Average $LAI_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
 This relationship was used to calculate the LAI anomaly in Fig.
-3, as the difference between LAI predicted from this regression and the actual $L_\tau$ at any given time.
+3, as the difference between LAI predicted from this regression and the actual $LAI_\tau$ at any given time.
 Red symbols are elevated C~a~, blue symbols ambient C~a~.
@@
 **Figure 3**.Time series of LAI after correcting for plot-to-plot differences in basal area (BA).
-At every time point, the LAI anomaly was calculated as $L_\tau$ minus LAI fitted from the relationship with BA shown in Fig.
+At every time point, the LAI anomaly was calculated as $LAI_\tau$ minus LAI fitted from the relationship with BA shown in Fig.
 2.
@@
 (a) Total litter production (converted from data in Fig.
-5 to an annual estimate) for each of the FACE rings plotted against mean $L_\tau$ during the entire litter collection period.
-The ratio of mean $L_\tau$ to annual litter production gives the mean leaf life span.
+5 to an annual estimate) for each of the FACE rings plotted against mean $LAI_\tau$ during the entire litter collection period.
+The ratio of mean $LAI_\tau$ to annual litter production gives the mean leaf life span.
 Grey lines indicate leaf lifespan isolines (in years).
@@
 Dynamics of LAI in relation to environmental drivers.
-(a) Smoothed time series of $L_\tau$ averaged across all six FACE rings.
+(a) Smoothed time series of $LAI_\tau$ averaged across all six FACE rings.
 (b) From panel (a), we calculated the daily change in LAI (dL/dt), including the 95% confidence interval for the derivative of the time series.

### 2015-08-07 manuscript.Rmd
R version auto-inserted; removed LAI anomaly methods description

@@
 ## Data analysis
-We used generalized additive models (GAM) for all analyses of LAI and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015].
+We used generalized additive models (GAM) for all analyses of LAI and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R version `r r_version` [@RCT2015].
 A GAM is a semi-parametric fitting method that allows the estimation of smoothed trends in the data, including confidence intervals.
@@
 \
-For comparison between LAI from photos and PAR transmittance, we used linear regression.
-To test for effects of C~a~ treatment on $LAI_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
+To test for effects of C~a~ treatment on LAI and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
+The random effect accounts for the repeated measures nature of the data.
+For LAI, we further used pre-treatment basal area (BA) of each FACE ring as a covariate.
+Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.
 
-\
-To assess effects of eC~a~ on $LAI_\tau$, we accounted for pre-treatment differences in canopy structure, as follows.
-The average $LAI_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the eC~a~ treatment across the FACE rings (Fig.
-3).
-Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.
-We then calculated the deviation of actual $LAI_\tau$ from the fitted relationship with BA shown in Fig.
-3, to yield a 'LAI anomaly'.
-This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.
+**More details here;package, which anova for Table 1; and now we use BA as a covariate  **
 
@@
 
+# Supporting information captions {.unnumbered}
+
+**Figure S1**. Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately.
+
+**Figure S2**. Correlation between L estimated from diffuse canopy transmittance (LAI) and from canopy photos.
+
+**Figure S3**. Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
+
 # References {.unnumbered}

### 2015-08-07 manuscript_calcs.R

@@ -15,6 +15,10 @@ Library(reporttools)
 # Load analysis cache.
 load("cache/lai_workspace.RData")
 
+# Current R version
+rv <- R.Version()
+r_version <- paste(rv$major, rv$minor, sep=".")
+
 # Trim data to this date (note: has to be consistent with make_figures.R!) 
 maxdate <- as.Date("2015-3-1")
 

### 2015-08-15 manuscript.Rmd
Litter methods moved from results; some notes added for revision

@@
 
+\ To compare the dynamics of LAI with litter production, we calculated the change in LAI for every time period when litter was collected.
+We did this by fitting a generalized additive model to LAI for each ring separately (similar to that shown for $\tau_d$ in Fig.
+1), and calculating the change in LAI for each period ($\Delta L_\tau$).
+Next, we estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $LAI_\tau$.
+For each litter collection period, the net change in $LAI_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries.
+Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period.
+
 \
@@
 \
-An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015].
+An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015; pers.
+comm.
+Andrew Gherlenda, Markus Riegler].
 This rapid decrease in LAI was evident in a quick increase in $\tau_d$ (Fig.
@@
 \
-We found no differences in the LAI anomaly between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any point in time, Fig.
+We found no differences in LAI between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig.
 3).
 A brief exception was the period where one of the rings was affected by an insect outbreak (see above).
-A linear mixed-effects model with C~a~ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on the LAI anomaly (p `r pval(LanomAOV[[3]][1])`).
+A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig.
+3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p **fix this**).
+** rethink the following; CI on least-square means accounting for BA?**
 But what difference between C~a~ treatments could we have reasonably detected?
@@
 \
-The dynamics of $LAI_\tau$ were compared with litter fall, by estimating the absolute change in $LAI_\tau$ for every FACE ring, for every period when litter was collected.
-We did this by fitting a generalized additive model to $LAI_\tau$ for each ring separately (similar to that shown for $\tau_d$ in Fig.
-1), and calculating the change in $LAI_\tau$ for each period ($\Delta L_\tau$).
 Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig.
@@
 \
-We estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $LAI_\tau$.
-For each litter collection period, the net change in $LAI_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries.
-Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period.
 The temporal dynamics of litter and leaf production are shown in Fig.

### 2015-08-15 review1.Rmd
Added first review from GCB and started on responses.

In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

Overall, it is my opinion that this manuscript is unsuitable for publication in Global Change Biology in its current form and should be rejected. I do not take this recommendation lightly, since data from forest CO2 enrichment experiments is highly valuable and in great demand. However, I have several issues with the field techniques and analyses addressed in this manuscript and the effort required to address them warrants the submission of an entirely new manuscript in my opinion.

The largest issue is that LAI anomaly (the main response analyzed) is essentially calculated from the mean of LAI over the experimental period. In an attempt to account for initial differences between plots, the authors correlated mean LAI and basal area for each plot. They then calculated the difference between the observed LAI and the expected value based on the LAI-BA relationship and defined it as “LAI anomaly.” However, because LAI and BA were so tightly correlated, this LAI anomaly is essentially the same as the difference between the mean LAI and the observed LAI. Since LAI means are the means of the observed values, all plots have a mean LAI anomaly of approximately zero (this can be seen in Fig. 3). This heavily restricts the ability of any statistical test to detect differences between plots or treatments. Thus, the finding of no response of LAI anomaly to elevated CO2 is not surprising.

**We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect. We only show the results when including BA as a covariate in the manuscript.**

The only statistically significant response of LAI that would be expected under the current statistical design would be from a CO2 x time interaction. While no statistics are reported, an interaction may have been detected but attributed exclusively to an insect outbreak, whereas it also has the potential to due to a real response. However, the inclusion of the post-outbreak data makes this difficult to discern. The outbreak confounds the effect of CO2 with that of the outbreak, and thus data from the infested plot should only be considered up until the outbreak occurred.

**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect. Although there was a brief period when the 95% confidence intervals were not overlapping, this brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect).**

The techniques used to collect litter are also concerning. The authors deployed eight 0.2 m2 litter traps in each ~123 m2 plot and used the litter collected from these traps to estimate litter production. However, considering the size of each individual trap and scale of the experiment, there is significant doubt that the traps provided a representative sample of each plot. Additionally, the analyses performed using litter data make several assumptions that are unlikely to hold.

*TODO* **We have included more background on the design of the litter collection methods. As the table below shows, the number and size (and fractional area of the plot covered) are in the upper range of six other FACE experiments. It is clear from this comparison that our methods are justified. Further, the standard errors for litter within a plot BLAH BLAH**

Another issue lies in the inclusion of an add-on analysis comparing two LAI estimation techniques. While useful, this comparison should be its own methods paper and does not align well with the scope of the rest of the manuscript. Currently, it adds to an already lengthy paper and distracts from the authors’ main points.

**We have moved this analysis, including the methods, to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a side-objective. We only briefly mention in the Methods that the two methods gave comparable results.**

Finally, the manuscript requires extensive editing for structure and flow. The introduction does not sufficiently establish the expectation of litter responses.

*TODO* **We have added a couple of sentences indicating the expectation of the litter response; it follows logically from the test of differences in LAI**

The methods are incomplete and unclear. Specifically, there are many portions of the Results section that should be included in Materials and Methods.

**We have moved all methods-related content from the Results to the Methods section. We respond further to specific comments related to the Methods below**

Additionally, it is unclear if the authors’ used repeated measures to analyze their data, which would be necessary considering the experimental design.

**Using a mixed-effects model as we have done (also in the original manuscript), with FACE ring as a random effect, accounts for the repeated measures nature of the data. We have clarified this in the Methods.**

Beyond these issues, the manuscript also should be revised for clarity and brevity.

**We have done so in response to specific comments listed below**

Specific Comments (line number given):
  Abstract
l.26 What do you mean by (hot summer), and is it necessary? This is not how you describe the climate in the methods.

**Reworded to 'subtropical', which is more fitting** *IS THIS CORRECT*

l.30 “this timeframe” is a stretched reference (ref. l.24). Reword

**Removed 'in this timeframe'.**

l.31 You don’t provide expectation of a seasonal pattern in the abstract, so explaining that there isn’t one is distracting here.
l.37 I would expect that GCB readers would know what FACE stands for, but it might be good to say free air CO2 enrichment to be safe.

**It was already defined both at the top of the Abstract and the second paragraph of the Introduction**

Introduction
l.40 Does it save a lot of space to abbreviate LAI as L or is there another reason for not using the conventional LAI? Unless there’s a strong reason to keep it as L, it would be more intuitive to have this as LAI.

**Changed to LAI throughout the manuscript**

l.55 drop the “the” in “the water-use efficiency”

**done**

l.60 this would make a better topic sentence for this paragraph than the current one

**todo**

l.69+ you can use stronger language here and throughout. If you expect X, say so. No need for “can” or “could.” Also, this is a very drawn-out sentence. The length of this sentence and others throughout weakens them and reduces their clarity. Be concise, particularly in your topic sentences.

**todo**

l.75 The meaning of Ca can be inferred from your definition of eCa, but it’s better to just write it out when you’re not using a defined term. So here, atmospheric [CO2]

**todo**

l.79 This sentence is awkwardly constructed. Again, too long and an unclear topic sentence. If the point of this sentence is to set up a paragraph about water, why start it off talking about temperature and photo period? It would be better to frame this as quantitative water effects being relatively unknown as opposed to qualitative effects being well-known. I would just drop l.79-81, and start with “The response of L to fluctuations in water availability….” Issues like this need to be addressed throughout the paper.

**todo**

l.94 the tense of this paragraph is weird. You’re moving between your hypotheses (which should be present tense) to your methods (in past tense). This is makes it unclear which statements you’re making a priori.

**todo**

l.95 to eCa in the EucFace experiment. EucFace needs to be defined in the body of the manuscript

**todo**

l.97 are these means over the same period? This type of detail should be in the methods, not the introduction

**todo**

l.99 Because Production – Litter = LAI? Currently, the introduction doesn’t address litter fall, so why is this expected to change?

**todo**

Materials and Methods
l.110 What do you mean by undisturbed? Be specific

**todo**

l.116 How accurate is this calculation? Are the P and PET values for the site or the region? Are they from different areas?

**todo**

l.136 unclear use of Ca

**todo**

l.160 “litter cloth basket” is odd. Where were they placed? This is a very small portion of the total plot area and indeed the area where litter would be expected to fall (an area much larger than the plot). So, these baskets are likely capturing only a tiny fraction of the litter from each plot and might even be catching litter from outside the plot as well, depending on where they are located. Can such a small and potentially contaminated sample be representative of such a large plot?

**We have responded to this issue above.**

l.163 provide this calculation. Was it LAI (cm2/cm2) = SLA  (cm2/g) x Litter (g/cm2)?

**totally obvious but yeah we'll add that**

l.206 The calibration is difficult to follow

**todo. which part is difficult???**

l.267 Was this set up as a repeated measures?

**toto. as mentioned above, yes. We have now clarified this.**

l.269 The calculation of the anomaly is problematic. Ideally, it would be done considering some pre-treatment value or, less desirably, using a starting condition value. Since there is a strong correlation between mean LAI and BA, and the anomaly is currently based on values from this correlation, the mean anomaly of each plot is inherently very close to zero (this can be seen in Fig. 3). Thus, using the mean value from the experimental period inherently removes almost any possibility of there being a meaningful main treatment effect, since all rings are forced to have roughly the same mean anomaly over the course of the experiment. In contrast, anomalies calculated based on a pre-treatment values are not defined as having the same mean and therefore can reveal significant main effect differences.

**We have responded to the data analysis concerns above.**

Results
l.290 The insect outbreak confounds your results and is not something that you really set out to examine. Consider excluding ring 6 data from the outbreak onward.

**todo. But it happened so why rmeove it.**

l.299 See above comments (a significant main Ca effect is almost impossible). A significant Ca x Date interaction is possible, but not reported here. However, if a significant interaction was found, it seems like it was likely attributed only to the insect outbreak in ring 6, but might also be due to Ca effects obscured by the current anomaly calculation.

**We have responded to the concerns about data analysis above.**

l.301+ move to the discussion

**todo**

l.307 move to methods

**todo**

l.314 Did you verify that there was no leaf growth, or is this an assumption based on drought?

**todo. field-verified - clarify**

l.318 move to methods. This calculation assumes litter is grown and lost in the same time period. This is unlikely, especially since a month is an ecologically arbitrary duration. So, this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production. However, the leaves lost as litter were likely produced over a variety of time points. This also ignores the problems of incomplete litter capture and non-experimental litter entering the plot.

**no idea what this is all about. Explain that leaf production must equal net increase of LAI (LAIt=2 - LAIt=1) PLUS the litter produced in that time period. It has nothing to do what happened previously. Incomplete litter capture was responded to above**

l.330 methods
l.333 methods

**todo. moved to methods**

Discussion
l.361 Its somewhat misleading to claim that basal area was included as a covariate, since it was not actually included in any statistical analysis of a response variable.

**The previous analysis was equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis**

l.368 There are a lot of places throughout the manuscript where you start a thought with “X showed Y response,” which puts the emphasis on the authors (X) and not the concept (Y). The points you’re trying to make are Y, not who showed them. This is not a huge problem, but the current emphases on authors and not concepts makes portions of the manuscript feel disjointed and list-y.

**todo. fixed throughout**

l.376 you originally expected that, but did not find it, right? Again, remove “could”

**todo**

l.377 “very” is redundant. How does this relate to the point you’re trying to make?

**todo**

l.378 The purpose of this manuscript was not to make broad generalizations on FACE L responses. This whole paragraph should be dropped.

**todo. I'd hate to remove this completely - maybe move some of it to introduction to show that the responses sofar are inconclusive so we need more info?**

l.381-5 pick one of these two sentences

**todo**

l.388 eCa. Be consistent

**todo**

l.389+ how does this support the message of this manuscript?

**todo**

l.395 define LL earlier in the ms and use throughout

**todo**

l.401+ Is the point what all these studies found specifically, or that there are mixed effects of eCa on LL?

**todo**

l.408 formatted as parenthetical reference

**todo**

Figures: Consider using other colors to avoid problems for red/blue colorblind readers. This is a lot of figures, evaluate which ones are essential to the main points of the manuscript and put non-essential figures in supplements.

**todo. Move LAI-BA to the SuppInfo, otherwise keep everything. Six is not a lot.**

3: include a 1:1 line for positive values too.

**todo**

4: Increase line thickness.

**todo**

### 2015-08-15 manuscript.Rmd
Made first revisions according to reviewer 1

@@
 Because LAI responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified.
-We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common.
+We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a subtropical climate where water limitation is common.
 We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, LAI should increase with elevated [CO~2~].
 We estimated LAI from diffuse canopy transmittance, and measured monthly leaf litter production.
-Contrary to expectation, LAI did not respond to elevated [CO~2~] in this timeframe.
+Contrary to expectation, LAI did not respond to elevated [CO~2~].
 We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period, but did not show a clear seasonal pattern.
@@
 \
-Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012].
+Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in water-use efficiency [@Ainsworth2005; @Ellsworth2012].
 Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state LAI to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].

### 2015-08-17 manuscript.Rmd
Moved LAI-BA to SuppInfo; fixed SuppInfo so it is standalone; changed Fig 2 to LAI instead of LAI anomaly; changed fig labels to LAI

@@
 source("manuscript_calcs.R")
+
+# Make figures
+# source("make_figures.R")
 
@@
 Random effects were included to account for the grouping of data within FACE rings, except when we analysed average LAI across all rings (Fig.
-7).
+6).
 For Fig.
-7, we estimated the derivative of the fitted GAM, including confidence intervals for the derivative, following Simpson [-@Simpson2014; see also @Curtis2014].
+6, we estimated the derivative of the fitted GAM, including confidence intervals for the derivative, following Simpson [-@Simpson2014; see also @Curtis2014].
 
@@
 We found no differences in LAI between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig.
-3).
-A brief exception was the period where one of the rings was affected by an insect outbreak (see above).
+2).
 A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig.
@@
 Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig.
-4).
+3).
 During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where LAI was decreasing.
 During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
-5).
+4).
 
@@
 The temporal dynamics of litter and leaf production are shown in Fig.
-5, and demonstrate the irregular temporal pattern in leaf production and P~L~.
+4, and demonstrate the irregular temporal pattern in leaf production and P~L~.
 We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).
@@
 Next, we related P~L~ to average $LAI_\tau$ (during the litter collection period) across the six FACE rings (Fig.
-6a).
+5a).
 If leaf lifespan is constant, P~L~ would be linearly related to $LAI_\tau$.
@@
 We estimated mean leaf lifespan from Fig.
-6a as the ratio of $LAI_\tau$ to annual total P~L~.
+5a as the ratio of $LAI_\tau$ to annual total P~L~.
 We found a significant (p `r pval(LLtest$p.value)`) increase in leaf lifespan (LL) with elevated C~a~ (Fig.
-6b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
+5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
 
@@
 During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available (Fig.
-7).
+6).
 These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
 The dynamics of $LAI_\tau$ from the sensor record matched these events observed in the field quite well (Fig.
-7a), as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings.
+6a), as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings.
 From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval.
@@
 It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig.
-4 and 6), when the soil water content in the top 80 cm increased to ca.
+3 and 5), when the soil water content in the top 80 cm increased to ca.
 > 0.2 m^3^ m^-3^.
 Two periods of sustained decrease in LAI (red lines in Fig.
-7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.
+6b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.
 
@@
 Because estimated leaf lifespan averaged `r round(mean(ba$LL),1)` yrs (Fig.
-6), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
+5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
 It could be argued that the change in LAI in response to eC~a~ was too small to be detected.
@@
 We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig.
-6) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
+5) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
 Further years of observation are needed to determine if such a difference is sustained.
@@
 We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean LAI and total accumulated leaf litter (Fig.
-6).
+5).
 Although LL was increased in eC~a~, this was not reflected in a difference in leaf litter production at any time point (Fig.
-5), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
+4), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
 There is mixed evidence from other FACE sites on the relationship between LL and eC~a~.
@@
 We found a correlation between LAI dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing LAI (Fig.
-4).
+3).
 As expected, during periods of decreasing LAI, the change in LAI was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of LAI against litter production.
@@
 Since leaf flushing has been shown to be closely related to soil water availability in eucalypts [Fig.
-7; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig.
-5).
+6; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig.
+4).
 The consequence is that leaf turnover is highly variable in time (Fig.
-5).
+4).
 
@@
 In contrast, phenology of evergreen *Eucalyptus* is more opportunistic, and shows rapid response to water availability (Fig.
-7).
+6).
 Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth.
@@
 We also have not observed leaf flushing in winter time, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig.
-7).
+6).
 We argue that both minimum temperature and high soil moisture are required  for leaf growth of subtropical eucalypts such as *E.
@@
 
-```{r fig.width=4, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
-figure2(ba)
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
+figure2(subset(facegap_cloudy_byring, Date < maxdate))
 ```
@@
 **Figure 2**.
-Average $LAI_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
-This relationship was used to calculate the LAI anomaly in Fig.
-3, as the difference between LAI predicted from this regression and the actual $LAI_\tau$ at any given time.
-Red symbols are elevated C~a~, blue symbols ambient C~a~.
+Time series of LAI estimated from canopy transmittance.
+Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect.
+Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05.
 
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
-figure3(subset(facegap_cloudy_byring, Date < maxdate))
+```{r fig.width=6, fig.height=4, echo=FALSE, message=FALSE, echo=FALSE, message=FALSE, eval=includeFigs}
+figure3(dLAIlitter)
 ```
 
-**Figure 3**.Time series of LAI after correcting for plot-to-plot differences in basal area (BA).
-At every time point, the LAI anomaly was calculated as $LAI_\tau$ minus LAI fitted from the relationship with BA shown in Fig.
-2.
-Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect.
-Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05, except for a brief period in late 2014.
-This difference was due to an insect attack that led to partial defoliation of the canopy in one FACE ring (see Fig.
-1).
-
-```{r fig.width=6, fig.height=4, echo=FALSE, message=FALSE, echo=FALSE, message=FALSE, eval=includeFigs}
-figure4(dLAIlitter)
-```
-
-**Figure 4**.
+**Figure 3**.
 Comparison of absolute changes in LAI ($\Delta L$) to litter production (in units of LAI), calculated for each litter collection period (approx.
@@
 ```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
-figure5(dLAIlitter)
+figure4(dLAIlitter)
 ```
 
-**Figure 5**.
+**Figure 4**.
 Timeseries of litter production and estimated gross leaf production rate.
@@
 ```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs}
-figure6(ba)
+figure5(ba)
 ```
 
-**Figure 6**.
+**Figure 5**.
 (a) Total litter production (converted from data in Fig.
@@
 ```{r fig.width=10, fig.height=8, echo=FALSE, message=FALSE, eval=includeFigs}
-figure7(subset(facegap_cloudy_byring, Date < maxdate),
+figure6(subset(facegap_cloudy_byring, Date < maxdate),
                subset(facesoilwater, Date < maxdate),
@@
 
-**Figure 7**.
+**Figure 6**.
 Dynamics of LAI in relation to environmental drivers.
@@
 
+**Figure S4**. Average $LAI_\tau$ over the entire study period plotted against the pre-treatment basal area (sum of stem cross-sectional areas) of the FACE rings.
+
 # References {.unnumbered}

### 2015-08-17 manuscript_supportinginfo.Rmd
Moved LAI-BA to SuppInfo; fixed SuppInfo so it is standalone; changed Fig 2 to LAI instead of LAI anomaly; changed fig labels to LAI

@@
 ---
+
+```{r echo=FALSE, warning=FALSE, message=FALSE}
+source("manuscript_calcs.R")
+```
 
@@
 This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.
+
+\
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=TRUE}
+figureSI2(flatcan_byring)
+```
+
+**Figure S3**.
+Correlation between LAI estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
+Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
+(a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
+[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
+
+\
+```{r fig.width=4, fig.height=4, echo=FALSE, message=FALSE}
+figureSI4(ba)
+```
+
+**Figure SI4**.
+Average $LAI_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
+This relationship was used to calculate the LAI anomaly in Fig.
+3, as the difference between LAI predicted from this regression and the actual $LAI_\tau$ at any given time.
+Red symbols are elevated C~a~, blue symbols ambient C~a~.
 
@@
 
-\
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=TRUE}
-figureSI2(flatcan_byring)
-```
-
-**Figure S2**.
-Correlation between LAI estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
-Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
-(a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
-[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
+## References

### 2015-08-17 R/figures.R

@@ -44,7 +44,6 @@ figure1 <- function(df, ramp){
              ylab=expression(tau[d]~~("-")),
              pointcols=rep("grey",8), linecols=my_ringcols())
   
-  
   timeseries_axis()
   
   axis(2)
@@ -62,55 +61,43 @@ figure1 <- function(df, ramp){
 }
 
 
-
-
-figure2 <- function(df){
-  
-  par(mar=c(5,5,2,2), cex.lab=1.2,xaxs="i", yaxs="i", tcl=0.2, las=1)
-  with(df, plot(BA, LAI.mean, pch=19, cex=1.2, col=my_ringcols(),
-                xlab=expression(Basal~area~~(m^2~ha^-1)),
-                ylab=expression(bar(italic(L))~~(m^2~m^-2)),
-                panel.first=predline(lm(LAI.mean ~ BA, data=ba)),
-                ylim=c(1.2,2.4), xlim=c(18,40)))
-  legend("bottomright", as.character(1:6), pch=19, bty='n',
-         col=my_ringcols(), title="Ring", cex=0.6, pt.cex=1)
-
-}
-
-
-figure3 <- function(df,
+figure2 <- function(df,
                     xlim=NULL, ylim=NULL,
                     legend=TRUE,
-                    ylab=expression(italic(L)~anomaly~(m^2~m^-2)),
+                    ylab=expression(italic(LAI)~~(m^2~m^-2)),
                     cex.lab=1.1, cex.axis=0.8, cex.legend=0.7,
                     legendwhere="topleft",
-                    setpar=TRUE,axisline=3,
+                    setpar=TRUE,
+                    axisline=3,
                     horlines=TRUE,
                     greyrect=FALSE,
                     addpoints=TRUE){
     
   if(setpar)par(cex.axis=cex.axis, mar=c(3,5,2,5), las=1, cex.lab=1.2, yaxs="i", tcl=0.2)
   par(cex.lab=cex.lab)
-  palette(my_co2cols())
+  palette(rev(my_co2cols()))
   
   if(is.null(xlim))xlim <- with(df, c(min(Date)-15, max(Date)+15))
-  if(is.null(ylim))ylim <- c(-0.8,0.8)
+  if(is.null(ylim))ylim <- c(0,2.8)
   
-  smoothplot(Date, LAIanomaly, g=treatment, R="Ring", ylim=ylim, xlim=xlim, 
[…]

### 2015-08-17 make_figures.R

@@ -11,32 +11,29 @@ to.pdf(figure1(subset(facegap_cloudy_byring,Date < maxdate), ramp),
        filename="output/figures/Figure1.pdf",
        width=8, height=6)
 
-to.pdf(figure2(ba), 
+to.pdf(figure2(subset(facegap_cloudy_byring, Date < maxdate)),
        filename="output/figures/Figure2.pdf",
-       width=4, height=4)
-
-to.pdf(figure3(subset(facegap_cloudy_byring, Date < maxdate)),
-       filename="output/figures/Figure3.pdf",
        width=8, height=4)
+
+to.pdf(figure3(dLAIlitter),
+       filename="output/figures/Figure3.pdf",
+       width=6, height=4)
 
 to.pdf(figure4(dLAIlitter),
        filename="output/figures/Figure4.pdf",
-       width=6, height=4)
+       width=8, height=4)
 
-to.pdf(figure5(dLAIlitter),
+to.pdf(figure5(ba), 
        filename="output/figures/Figure5.pdf",
        width=8, height=4)
 
-to.pdf(figure6(ba), 
-       filename="output/figures/Figure6.pdf",
-       width=8, height=4)
-
-to.pdf(figure7(subset(facegap_cloudy_byring, Date < maxdate),
+to.pdf(figure6(subset(facegap_cloudy_byring, Date < maxdate),
                subset(facesoilwater, Date < maxdate),
                subset(faceraindaily, Date < maxdate),
                subset(airt, Date < maxdate)),
-       filename="output/figures/Figure7.pdf",
+       filename="output/figures/Figure6.pdf",
        width=10, height=8)
+
 
 to.pdf(figureSI1(litring, subset(facegap_cloudy_byring, Date < maxdate)),
        filename="output/figures/FigureSI1.pdf",
@@ -50,14 +47,18 @@ to.pdf(figureSI3(subset(facegap_cloudy_byring, Date < maxdate), subset(facegap_a
        filename="output/figures/FigureSI3.pdf",
        width=8, height=5)
 
+to.pdf(figureSI4(ba), 
+       filename="output/figures/FigureSI4.pdf",
+       width=4, height=4)
 
 
-
-to.pdf(figure_timeseries(facegap_cloudy_byring, flatcan_byCO2),
-       filename="output/LAI_timeseries.pdf",
-       width=10, height=8)
-
-to.pdf(figure3b(facegap_cloudy_byring),
[…]

### 2015-08-17 manuscript.Rmd
Included litter basket table in response to reviewer.

@@
 ## Leaf litter production
-In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor.
+In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at random locations (**can I say this?**) at 0.5m above the forest floor.
 Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
@@
 
-where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR (µmol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
+where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR ($\mu$mol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring.
 We then calculated daily averages of $\tau_d$, excluding days where fewer than 4 time steps (i.e.

### 2015-08-17 review1.Rmd
Moved LAI-BA to SuppInfo; fixed SuppInfo so it is standalone; changed Fig 2 to LAI instead of LAI anomaly; changed fig labels to LAI
Included litter basket table in response to reviewer.

@@
+---
+output: word_document
+---
 In this manuscript, Duursma et al.
@@
 There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect.
-We only show the results when including BA as a covariate in the manuscript.**
+We only show the results when including BA as a covariate in the manuscript.
+Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**
 
@@
 **As noted above, we have redone the analysis and find no CO2 or CO2 x time effect.
-Although there was a brief period when the 95% confidence intervals were not overlapping, this brief period did not result in a significant overall CO2 x time effect.
+Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect.
 For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect).**
@@
 
-*TODO* **We have included more background on the design of the litter collection methods.
-As the table below shows, the number and size (and fractional area of the plot covered) are in the upper range of six other FACE experiments.
-It is clear from this comparison that our methods are justified.
-Further, the standard errors for litter within a plot BLAH BLAH**
+**and**
+
+l.160 “litter cloth basket” is odd.
+Where were they placed?
+This is a very small portion of the total plot area and indeed the area where litter would be expected to fall (an area much larger than the plot).
+So, these baskets are likely capturing only a tiny fraction of the litter from each plot and might even be catching litter from outside the plot as well, depending on where they are located.
+Can such a small and potentially contaminated sample be representative of such a large plot?
+
+**The size, number, and relative coverage of the litterfall traps is broadly consistent with other forest FACE studies (see Table below), with the largest trap size.
+The size and coverage of the traps is therefore fully justified.
+The litter traps were placed in somewhat random locations across the ring, but never within XXX meters of the plot edge, to minimize contamination.
+We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**
+
+```{r echo=FALSE}
+tabl <- read.csv(text="Experiment,	Reference,	Nr Traps per plot,	Trap size (m2),	Plot size (m2),	Coverage (%)
+Duke,Finzi et al. 2001,	12,	0.16,	908,	0.0021
+Oak Ridge,Norby et al. 2003, 	7,	0.19,	491,	0.0027
+PopFACE,Lagomarsino et al. 2013,	18,	0.13,	300, 0.0078
+Rhinelander,Liu et al. 2009,	18,	0.145,	707,	0.0037
+BangorFACE,Godbold et al. 2014,	7,	0.11,	57,	0.0136
+EucFACE,this paper,	8,	0.1979,	491,	0.0032", header=TRUE, check.names=FALSE)
+
+tabl <- tabl[,-ncol(tabl)]
+
+library(knitr)
+kable(tabl)
+```
 
@@
 **We have moved this analysis, including the methods, to the Supporting Information.
-We believe this is interesting information for others concerned with LAI estimation, but agree it is a side-objective.
+We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only.
 We only briefly mention in the Methods that the two methods gave comparable results.**
 
-Finally, the manuscript requires extensive editing for structure and flow. The introduction does not sufficiently establish the expectation of litter responses.
+Finally, the manuscript requires extensive editing for structure and flow.
+
+**We have made substantial edits to the text, in particular accounting for the reviewer's comments listed below.**
+
+The introduction does not sufficiently establish the expectation of litter responses.
 
@@
 
-**We have done so in response to specific comments listed below**
+**We have done so in response to specific comments listed below.**
 
@@
 **todo**
-
-l.160 “litter cloth basket” is odd.
-Where were they placed?
-This is a very small portion of the total plot area and indeed the area where litter would be expected to fall (an area much larger than the plot).
-So, these baskets are likely capturing only a tiny fraction of the litter from each plot and might even be catching litter from outside the plot as well, depending on where they are located.
-Can such a small and potentially contaminated sample be representative of such a large plot?
-
-**We have responded to this issue above.**
 
@@
 
-**todo**
+**Done.**
 

### 2015-08-17 R/figures.R
fixed dangerous switch of palette

@@ -75,7 +75,6 @@ figure2 <- function(df,
     
   if(setpar)par(cex.axis=cex.axis, mar=c(3,5,2,5), las=1, cex.lab=1.2, yaxs="i", tcl=0.2)
   par(cex.lab=cex.lab)
-  palette(rev(my_co2cols()))
   
   if(is.null(xlim))xlim <- with(df, c(min(Date)-15, max(Date)+15))
   if(is.null(ylim))ylim <- c(0,2.8)
@@ -86,6 +85,7 @@ figure2 <- function(df,
   smoothplot(Date, LAI, g=treatment, R="Ring", ylim=ylim, xlim=xlim, 
              ylab=ylab, xlab="",
              data=df, kgam=18, axes=FALSE,
+             pointcolor=rev(my_co2cols()),
              polycolor=rev(c(alpha("royalblue",0.7),alpha("pink",0.7))))
   
   l <- legend("topleft", c("Ambient","Elevated"), title=expression(italic(C)[a]~treatment), 

### 2015-08-17 R/data_processing.R
mean LAI in dLAIlitter

@@ -263,7 +263,7 @@ make_dLAI_litter <- function(dat, litter, kgam=15){
       d <- diff(z$LAIsm)
       dnegLAI <- sum(d[d < 0])
       dposLAI <- sum(d[d > 0])
-      return(data.frame(dLAI=dLAI, dnegLAI=dnegLAI, dposLAI=dposLAI))
+      return(data.frame(dLAI=dLAI, dnegLAI=dnegLAI, dposLAI=dposLAI, LAI=mean(z$LAIsm)))
     }))
   }
   

### 2015-08-17 run.R
reset digits option

@@ -111,10 +111,13 @@ save.image(file="cache/lai_workspace.RData")
 m1 <- paste0("Most recent day with usable cloudy data: ",max(facegap_cloudy_byring$Date))
 message(m1)
 
+olddigit <- options()$digits
 
 # Make manuscript
 render("manuscript.Rmd", "word_document", "manuscript.docx")
 render("manuscript_supportinginfo.Rmd", "word_document", "manuscript_SuppInfo.docx")
+
+options(digits=olddigit)
 
 # Make md like this:
 # render("manuscript.Rmd", md_document())

### 2015-08-17 R_other/lerp_and_lai.R
data and script for lerp data with LAI dynamics

@@ -1,0 +1,59 @@
+# data
+lerp <- read.csv("data/lerp_count.csv", stringsAsFactors = FALSE)
+lerp$ring <- paste0("R", lerp$ring)
+names(lerp)[1] <- "Ring"
+names(lerp)[4] <- "treatment"
+
+litdf <- dLAIlitter[,c("Date","Ring","LAI","dLAI","dLAIlitter.mean")]
+litdf$date <- format(litdf$Date, "%b-%y")
+
+lerplai <- merge(litdf, lerp)
+lerplai$species <- as.factor(lerplai$species)
+
+
+litprev <- do.call(rbind, lapply(split(litdf, litdf$Ring), function(x){
+  x <- x[order(x$Date),]
+  y <- x[c(1,1:(nrow(x)-1)),]
+  y[1,3:5] <- NA
+  y$Date <- x$Date
+  y$date <- NULL
+  
+  names(y)[3:5] <- paste(names(y)[3:5], "prev", sep="_")
+  
+  return(y)
+}))
+
+lerplai <- merge(lerplai, litprev)
+
+lerplai$leafgrowth <- with(lerplai, dLAI + dLAIlitter.mean)
+
+
+write.csv(lerplai, "output/data/lerplai.csv", row.names=FALSE)
+
+
+
+# figures
+palette(c("forestgreen","darkorange","royalblue"))
+pchs <- c(19,15,17)
+
+with(lerplai, plot(dLAIlitter.mean, log10(abundance_m2), col=as.factor(species),
+                   pch=pchs[species]))
+
+with(lerplai, plot(dLAIlitter.mean_prev, log10(abundance_m2), col=as.factor(species),
+                   pch=pchs[species]))
+
+
+windows(7,7)
+par(mfrow=c(3,1), mar=c(0,0,0,0), oma=c(5,5,2,2))
+with(lerplai, plot(Date, leafgrowth))
+with(lerplai, plot(Date, dLAIlitter.mean))
+with(lerplai, plot(Date, log10(abundance_30days), pch=19, col=species))
+
+
+
+
+
+
+
+
+

### 2015-08-18 manuscript.Rmd
Accounted for (nearly?) all minor comments from reviewer.

@@
 \
+Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in LAI.
 Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in water-use efficiency [@Ainsworth2005; @Ellsworth2012].
 Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state LAI to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
-Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in LAI.
 Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased LAI in response to eC~a~ [@McCarthy2007; @Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
@@
 \
-In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect LAI [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability.
-These ‘water savings’ can be expected to lead to an increase in LAI in water-limited ecosystems, consistent with a tight coupling of LAI with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
+In ecosystems where water availability limits productivity, we expect that LAI responds especially strong to eC~a~ [@Donohue2009].
+This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability.
+These 'water savings' can be expected to lead to an increase in LAI in water-limited ecosystems, consistent with a tight coupling of LAI with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
 In support of this prediction, Donohue et al.
-[-@Donohue2013] attributed the observed global increase in LAI in arid and semi-arid areas over the last 30 years to rising C~a~.
+[-@Donohue2013] attributed the observed global increase in LAI in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~].
 Based on the same water-savings principle, we predicted that LAI should increase under eC~a~ in water-limited forests.
@@
 \
-Most previous studies on LAI responses to environmental drivers have focused on the seasonal, mostly temperature and photo period-driven, timing of LAI dynamics observed in N-hemisphere forests [e.g.
-@Richardson2012; @Savoy2015] whereas LAI responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
+The responses of LAI to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
 The strong relationship between LAI and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
@@
 We estimated LAI from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
-The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration [long-term mean of 1350 mm; @Zomer2008] and mean annual precipitation of 800 mm.
 The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of LAI and its components (litter fall and production) in response to water availability.
-We hypothesized that eC~a~ would result in an increase in LAI, particularly during periods of water limitation.
-In addition, we expected that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a positive response of LAI to eC~a~ [cf.
-@Norby2011], either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
-Our second objective was to quantify the temporal dynamics of LAI, in particular in relation to fluctuating water availability.
+We hypothesize that eC~a~ would result in an increase in LAI, particularly during periods of water limitation.
+In addition, we expect that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of LAI to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
+Our second objective is to quantify the temporal dynamics of LAI, in particular in relation to fluctuating water availability.
 
@@
 The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.).
-The vegetation consists of a native woodland undisturbed for ca.
-70 years and maintained in this location since the 1880’s.
+The vegetation consists of a native woodland undisturbed (unmanaged) for ca.
+70 years and maintained in this location since the 1880's.
 The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
@@
 Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
-The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction (C~a~) higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots.
+The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots.
 The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode.
@@
 In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at random locations (**can I say this?**) at 0.5m above the forest floor.
+To minimize contamination from trees outside the rings, the baskets were installed at least 2m from the plot edge.
 Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
-Litter production was converted to units of LAI using an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
+Litter production was converted to units of LAI by multiplying it with an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
 We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.
@@
 Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
-Random effects were included to account for the grouping of data within FACE rings, except when we analysed average LAI across all rings (Fig.
+Random effects were included to account for the repeated measures structure of the data, except when we analysed average LAI across all rings (Fig.
 6).
-For Fig.
-6, we estimated the derivative of the fitted GAM, including confidence intervals for the derivative, following Simpson [-@Simpson2014; see also @Curtis2014].
 
@@
 Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period.
+We estimated mean leaf lifespan (LL) as the ratio of $LAI_\tau$ to annual total P~L~.
+
+\
+We quantified the relationship between timing of leaf flushing and environmental drivers.
+During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available.
+These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
+These leaf growth events were compared to the dynamics of LAI as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings.
+From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014; see also @Curtis2014].
 
@@
 3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p **fix this**).
-** rethink the following; CI on least-square means accounting for BA?**
-But what difference between C~a~ treatments could we have reasonably detected?
-The average half-width of the confidence interval on the LAI anomaly (Fig.
-4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean LAI of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
 
@@
 3).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where LAI was decreasing.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where LAI was decreasing (see also Fig.
+6).
 During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
@@
 5a).
-If leaf lifespan is constant, P~L~ would be linearly related to $LAI_\tau$.
+If leaf lifespan (LL) is constant, P~L~ would be linearly related to $LAI_\tau$.
 Instead, we found that in eC~a~, P~L~ was lower at a given $LAI_\tau$ as compared to aC~a~ (Fig.
 6a).
-We estimated mean leaf lifespan from Fig.
-5a as the ratio of $LAI_\tau$ to annual total P~L~.
-We found a significant (p `r pval(LLtest$p.value)`) increase in leaf lifespan (LL) with elevated C~a~ (Fig.
+We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig.
 5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
@@
 \
-Finally, we quantified the relationship between timing of leaf flushing and environmental drivers.
-During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available (Fig.
-6).
-These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
-The dynamics of $LAI_\tau$ from the sensor record matched these events observed in the field quite well (Fig.
-6a), as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings.
-From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval.
+The dynamics of $LAI_\tau$ from the sensor record matched leaf growth events observed in the field quite well (Fig.
+6a).
 All four leaf flushing events corresponded to significant positive dL/dt.
@@
 [-@Norby2003], we found no significant effect of elevated C~a~ on LAI.
-Because estimated leaf lifespan averaged `r round(mean(ba$LL),1)` yrs (Fig.
+Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig.
 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
@@
 Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.
+
+** rethink the following; CI on least-square means accounting for BA?**
+But what difference between C~a~ treatments could we have reasonably detected?
+The average half-width of the confidence interval on the LAI anomaly (Fig.
+4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment.
+This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean LAI of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
 
@@
 [-@Iio2013].
-Based on this LAI estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we could expect about a 30% increase in LAI under eC~a~.
-In contrast to other forest FACE sites, the EucFACE is very strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
+Based on this LAI estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in LAI under eC~a~.
+In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
 Hence, it may be too early to draw general conclusions across FACE sites in terms of LAI response.
@@
 \
-A possible explanation for the lack of a LAI response to eC~a~ is that leaf growth may not be directly related to increased carbon availability as a result of eC~a~.
-In other words, stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
-*in review*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
+Stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
+*in press*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.
 Either process may itself be affected by eC~a~ [@DeKauwe2014].
-We found a significant effect of eC~a~ on leaf lifespan, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in elevated C~a~ (Fig.
+We found a significant effect of eC~a~ on LL, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in eC~a~ (Fig.
 5) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
-Further years of observation are needed to determine if such a difference is sustained.
 Because we found no difference in LAI with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
@@
 \
-Despite a lack of response of canopy LAI, we did find a significantly longer leaf lifespan (LL) in eC~a~.
+Despite a lack of response of canopy LAI, we did find a significantly longer LL in eC~a~.
 We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean LAI and total accumulated leaf litter (Fig.
@@
 4), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
-There is mixed evidence from other FACE sites on the relationship between LL and eC~a~.
+There is a variety of reports from other FACE sites on the relationship between LL and eC~a~.
 Norby et al.
@@
 [-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
-Although this indicates the canopy stayed green longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ at the PopFACE, leading to similar LAI in ambient C~a~ despite higher productivity.
+Although this indicates the canopy stayed green longer in elevated C~a~, Gielen et al.
+[-@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ at the PopFACE, leading to similar LAI in ambient C~a~ despite higher productivity.
 Li et al.
@@
 Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
-Unlike in deciduous canopies, litter production cannot be used to estimate LAI, because leaf lifespan (LL) is unknown (instead, we used LAI and litter production to estimate LL).
+Unlike in deciduous canopies, litter production cannot be used to estimate LAI, because LL is unknown (instead, we used LAI and litter production to estimate LL).
 We cannot not rely either on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E.

### 2015-08-18 review1.Rmd
Accounted for (nearly?) all minor comments from reviewer.

@@
 l.31 You don’t provide expectation of a seasonal pattern in the abstract, so explaining that there isn’t one is distracting here.
+
+**todo**
+
 l.37 I would expect that GCB readers would know what FACE stands for, but it might be good to say free air CO2 enrichment to be safe.
 
-**It was already defined both at the top of the Abstract and the second paragraph of the Introduction**
+**It was already defined at the top of the Abstract, when the term EucFACE is introduced.**
 
@@
 
-**Changed to LAI throughout the manuscript**
+**Changed to LAI throughout the manuscript.**
 
@@
 
-**done**
+**Done.**
 
@@
 
-**todo**
+**This sentence is now the topic sentence of the paragraph.**
 
@@
 
-**todo**
+**Edited the paragraph throughout.**
 
@@
 
-**todo**
+**Edited the paragraph throughout (and deleted l.79-81).**
 
@@
 
-**todo**
+**Edited the paragraph throughout, with special attention to tense.**
 
@@
 
-**todo**
+**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is quite obvious from the statement that the woodland is Eucalyptus-dominated**
 
@@
 
-**todo**
+**Removed info on precipitation and PET here, as it is explained in more detail in the Methods already.**
 
@@
 
-**todo**
+**If LAI changes, this must be due to changes in either the production of new foliage, or the lifespan of existing foliage.
+Both components can be assessed with measurements of litter production in combination with the time dynamics of LAI.** *Expand intro on litter fall a bit!!*
 
@@
 
-**todo**
+**Added '(unmanaged)'.**
 
@@
 
-**todo**
+**The estimate of PET is used only here to give an estimate of the ratio P/PET as an indicator of long-term water limitation.
+The global PET database gives average PET based on 1950-2000 climatology, and is widely used to estimate PET across sites.
+It is beyond the scope of this manuscript to calculate more accurate estimates of PET, as we do not use PET in any of the subsequent analyses.**
 
@@
 
-**todo**
+**Removed Ca.**
 
@@
 
-**totally obvious but yeah we'll add that**
+**Changed the text to 'by multiplying it with an estimate of [...] SLA.**
 
@@
 
-**toto. as mentioned above, yes. We have now clarified this.**
+**The random ring effect accounts for the repeated measures structure of the data; we have now clarified this.**
 
@@
 
-**todo. But it happened so why rmeove it.**
+**We disagree that the outbreak confounded the results; we did not find effects of CO2 on LAI despite the outbreak.
+Although we did not set out to examine the outbreak, it occurred nonetheless and we think it is informative as the outbreak reflects natural dynamics of Eucalyptus-dominated woodlands.
+Also, we do not discuss the outbreak in any sort of detail, so removing it from the figure does not shorten the manuscript.**
 
@@
 
-**todo**
+**We have moved the discussion of discernable effect sizes to the discussion.**
 
@@
 
-**todo. field-verified - clarify**
+**We verified in the field that no leaf growth was occurring.
+We have clarified this in this section as well as made reference to Fig.
+6, where we have indicated when leaf growth was observed in the field.**
 
@@
 
-**no idea what this is all about.
-Explain that leaf production must equal net increase of LAI (LAIt=2 - LAIt=1) PLUS the litter produced in that time period.
-It has nothing to do what happened previously.
-Incomplete litter capture was responded to above**
+**Moved to methods.
+We disagree that this method 'assumes litter is grown and lost in the same time period'.
+Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shedded leaf area, as measured by litter production.
+And to  "this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production", this is only true if the reduction in LAI is less than the litter produced.
+In that case the canopy must have produced new leaves in that period to complete the sum of dLAI/dt = leafgrowth - litter.
+There are no assumptions on when the leaves were produced that show up in the litter traps at that time.**
+
+**We think that the term 'litter production' may be misleading, as the term 'production' seems to indicate that the leaves were 'produced' during that time period, which we certainly do not assume nor imply (in fact, we have shown that leaf lifespan is ca.
+1 year).**
 
 l.330 methods
+
+**Done.**
+
 l.333 methods
 
-**todo. moved to methods**
+**Done.**
 
@@
 
-**The previous analysis was equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis**
+**The previous analysis was essentially equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis.**
 
@@
 
-**todo. fixed throughout**
+**The manuscript was edited throughout to make statements more direct.**
 
@@
 
-**todo**
+**Removed 'could'.**
 
@@
 
-**todo**
+**Removed 'very'.**
 
@@
 
-**todo. I'd hate to remove this completely - maybe move some of it to introduction to show that the responses sofar are inconclusive so we need more info?**
+**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant discussion to the results of this study to the other FACE sites.
+This is informative because the results from the EucFACE do not fit well with a generalization across other FACE sites as presented by Norby and Zak (2011).**
 
@@
 
-**todo**
+**Deleted sentence on line 381.**
 
@@
 
-**todo**
+**Here we try to reconcile the apparent inconsistency : we found no effect of CO2 on LAI, but we did find a longer leaf lifespan in elevated CO2.
+This implies that the fractional allocation to leaves in elevated CO2 has decreased (through mass balance considerations).
+We have shortened this part of the paragraph.**
 
@@
 
-**todo**
+**We now define LL in the Methods and use it more consistently throughout.**
 
@@
 
-**todo**
+**Edited the sentence for clarity.**
 
@@
 
-**todo. Move LAI-BA to the SuppInfo, otherwise keep everything. Six is not a lot.**
+**We have moved Fig. 2 to the Supporting Information. We have retained the other figures as we feel they are essential to the manuscript (and six figures does not seem excessive to us).**
 

### 2015-08-18 review1.Rmd
edit to response

@@
 
+**We are surprised that the reviewer considers our paper 'lengthy', as the word count of ca. 5500 words is more than 30% lower than the allowed length for Global Change Biology (8000 words).**
+
 Finally, the manuscript requires extensive editing for structure and flow.
@@
 
-**We have done so in response to specific comments listed below.**
+**We have done so in response to specific comments listed below, and made more minor changes to increase clarity.**
 

### 2015-08-18 R/figures.R
fixed figure labels for consistency

@@ -64,7 +64,7 @@ figure1 <- function(df, ramp){
 figure2 <- function(df,
                     xlim=NULL, ylim=NULL,
                     legend=TRUE,
-                    ylab=expression(italic(LAI)~~(m^2~m^-2)),
+                    ylab=expression(LAI~~(m^2~m^-2)),
                     cex.lab=1.1, cex.axis=0.8, cex.legend=0.7,
                     legendwhere="topleft",
                     setpar=TRUE,
@@ -110,10 +110,10 @@ figure3 <- function(df){
   par(mar=c(5,5,2,2), cex.lab=1.1, cex.axis=0.9, las=1, xaxs="i", yaxs="i", tcl=0.2)
   with(dfup, plot(X, Y, pch=c(19,21)[treatment],
                col=Cols[LAIchange],
-               xlab=expression(Leaf~litter~production~~(m^2~m^-2~mon^-1)),
+               xlab=expression(P[L]~~(m^2~m^-2~mon^-1)),
                xlim=c(0,0.6),
                ylim=c(-0.3,0.5),
-               ylab=expression(Delta*italic(LAI)~from~tau[d]~(m^2~m^-2~mon^-1))))
+               ylab=expression(Delta*LAI~~(m^2~m^-2~mon^-1))))
   with(dfdown, points(X, Y, pch=c(17,24)[treatment],
                   col=Cols[LAIchange]))
   

### 2015-08-19 manuscript.Rmd
Many minor fixes

@@
 Remko Duursma
+\
 E: remkoduursma@gmail.com
+\
 T: +61(0)422096908
@@
 \
-In ecosystems where water availability limits productivity, we expect that LAI responds especially strong to eC~a~ [@Donohue2009].
+In ecosystems where water availability limits productivity, we expect an especially large response of LAI to eC~a~ [@Donohue2009].
 This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability.
@@
 The strong relationship between LAI and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
-However, local LAI is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability.
+However, local LAI is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011].
 Understanding dynamics of LAI in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
@@
 In addition, we expect that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of LAI to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
-Our second objective is to quantify the temporal dynamics of LAI, in particular in relation to fluctuating water availability.
+Our second objective is to quantify the temporal dynamics of LAI and its components (litter fall and leaf production), in particular in relation to fluctuating water availability.
 
@@
 70 years and maintained in this location since the 1880's.
-The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
-An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
+The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (1881 - 2014, Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
+An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [1950 - 2000 climatology, http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
 0.59.
-In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating very frequent water limitation.
+In addition, over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating frequent water limitation.
 
@@
 The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015].
-Groundwater depth is 12.6 – 12.9 m below the soil surface.
+Groundwater depth is 12.6 - 12.9 m below the soil surface.
 The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r f(mean(ba$BA))`, SE = `r f(sd(ba$BA)/sqrt(6))` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E.
@@
 Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
-The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots.
-The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode.
+The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received aC~a~, and are used as control plots.
+The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above aC~a~ in full operation mode.
 However, C~a~ was increased gradually for six months at a rate of ca.
@@
 ## Terminology
-In the following, we use the term 'leaf' area index (LAI) although the canopy and FACE rings contain non-leaf shading elements, i.e.
+In the following, we use the term leaf area index (LAI) although the canopy and FACE rings contain non-leaf shading elements, i.e.
 woody stems and infrastructure.
@@
 At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height.
-All PAR data were averaged in 30-min time steps.
+All PAR data were averaged at 30-min time steps.
 At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca.
@@
 For determination of LAI, we used only highly diffuse conditions, where F~diff~ > 0.98.
-We further trimmed the data by excluding low solar e levation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles.
+We further trimmed the data by excluding low solar elevation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles.
 Finally canopy transmittance was calculated for every 30-min time step as,
@@
 \
-We then estimated leaf area index (LAI) from $\tau_d$ ($LAI_\tau$), following Campbell and Norman [-@Campbell2000].
+We then estimated leaf area index (LAI) from $\tau_d$ following Campbell and Norman [-@Campbell2000].
 We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
@@
 
-where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
+where L is LAI, $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
 Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006].
 We solved numerically for LAI in Eq.
-2, using a smooth integration over $\theta$, yielding $LAI_\tau$.
+2, using a smooth integration over $\theta$, yielding LAI.
 
@@
 To overcome violations in the assumptions underlying Eq.
-2, we calibrated $LAI_\tau$ against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
+2, we calibrated LAI against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
 1).
@@
 
-where *c* is a calibration parameter, $L_{\tau,t=0}$ is $LAI_\tau$ at the start of the drought period.
+where *c* is a calibration parameter, $L_{\tau,t=0}$ is LAI at the start of the drought period.
 We used linear regression without intercept to estimate *c* in Eq.
 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`).
-It should be stressed that even after calibration with litter production, $LAI_\tau$ still includes stem and infrastructure area (see Terminology).
+It should be stressed that even after calibration with litter production, LAI still includes stem and infrastructure area (see Terminology).
 However, this approximately constant area cancels out in the left-hand side of Eq.
@@
 The random effect accounts for the repeated measures nature of the data.
-For LAI, we further used pre-treatment basal area (BA) of each FACE ring as a covariate.
+For LAI, we further used pre-treatment basal area (BA) of each FACE ring as a covariate (see Fig.
+S4).
 Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.
 
-**More details here;package, which anova for Table 1; and now we use BA as a covariate  **
-
-\ To compare the dynamics of LAI with litter production, we calculated the change in LAI for every time period when litter was collected.
-We did this by fitting a generalized additive model to LAI for each ring separately (similar to that shown for $\tau_d$ in Fig.
-1), and calculating the change in LAI for each period ($\Delta L_\tau$).
-Next, we estimated gross leaf production from measured litter production (P~L~) and temporal dynamics of $LAI_\tau$.
-For each litter collection period, the net change in $LAI_\tau$ was estimated using fitted generalized additive models to allow interpolation from a smoothed timeseries.
-Gross leaf production was then calculated as the sum of the net change in $LAI_\tau$ and the amount of leaf litter produced during that period.
-We estimated mean leaf lifespan (LL) as the ratio of $LAI_\tau$ to annual total P~L~.
+\
+To compare the dynamics of LAI with litter production, we calculated the change in LAI for every time period when litter was collected.
+We did this by fitting a GAM to LAI for each ring separately (similar to that shown for $\tau_d$ in Fig.
+1), and calculating the change in LAI for each litter fall collection period ($\Delta$LAI).
+Next, we estimated gross leaf production (G~L~) as the sum of the net change in LAI ($\Delta$LAI) and P~L~, because mass balance requires that $\Delta$LAI = G~L~ - P~L~.
+We estimated mean leaf lifespan (LL, yr) as the ratio of mean LAI (m^2^ m^-2^) for the entire litter fall collection period to total P~L~ (m^2^ m^-2^ yr^-1^).
 
@@
 We quantified the relationship between timing of leaf flushing and environmental drivers.
-During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $LAI_\tau$ were available.
+During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of LAI were available.
 These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
-These leaf growth events were compared to the dynamics of LAI as estimated from a generalized additive model fitted to the average $LAI_\tau$ across the six FACE rings.
-From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014; see also @Curtis2014].
+These leaf growth events were compared to the dynamics of LAI as estimated from a GAM fitted to the average daily LAI across the six FACE rings.
+From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014, see also @Curtis2014].
 
 \
-The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXXXX), published on figshare (doi:XXXX).
-The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, with the rmarkdown package in Rstudio, which uses Pandoc for conversion).
+The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXX), published on figshare (doi:XXX).
+The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, using the 'rmarkdown' R package for conversion).
 
@@
 Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.
-
-\
 An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings (ring 6), starting in September 2014 [for description of a similar outbreak in *Eucalyptus moluccana*, see @Hall2015; pers.
@@
 A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig.
-3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p **fix this**).
+3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p > 0.5).
 
 \
-Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig.
+Litter production (P~L~) correlated significantly with $\Delta$LAI when the latter was separated into periods where $\Delta$LAI > 0 (net leaf growth) and $\Delta$LAI <  0 (net leaf loss) (Fig.
 3).
@@
 6).
-During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig.
+During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta$LAI and litter production (Fig.
 4).
@@
 The temporal dynamics of litter and leaf production are shown in Fig.
-4, and demonstrate the irregular temporal pattern in leaf production and P~L~.
+4, and demonstrate the irregular temporal pattern in both leaf production and P~L~.
 We found no significant difference in either litter or leaf production between C~a~ treatments at any time point (p > 0.2 for both, assessed with linear-mixed effects models).
@@
 \
-Next, we related P~L~ to average $LAI_\tau$ (during the litter collection period) across the six FACE rings (Fig.
+Next, we related total P~L~ to average LAI (during the litter collection period) across the six FACE rings (Fig.
 5a).
-If leaf lifespan (LL) is constant, P~L~ would be linearly related to $LAI_\tau$.
-Instead, we found that in eC~a~, P~L~ was lower at a given $LAI_\tau$ as compared to aC~a~ (Fig.
+If leaf lifespan (LL) is constant, P~L~ would be linearly related to LAI.
+Instead, we found that in eC~a~, P~L~ was lower at a given LAI as compared to aC~a~ (Fig.
 6a).
-We found a significant (p `r pval(LLtest$p.value)`) increase in LL with elevated C~a~ (Fig.
+We found a significant (p `r pval(LLtest$p.value)`) increase in LL with eC~a~ (Fig.
 5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
@@
 \
-The dynamics of $LAI_\tau$ from the sensor record matched leaf growth events observed in the field quite well (Fig.
+The dynamics of LAI from the sensor record matched leaf growth events observed in the field quite well (Fig.
 6a).
@@
 Counter to our expectations but similar to Norby et al.
-[-@Norby2003], we found no significant effect of elevated C~a~ on LAI.
+[-@Norby2003], we found no significant effect of eC~a~ on LAI.
 Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig.
@@
 Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
-To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate.
+To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate.
 Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in LAI.
-Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.
+Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.
 
@@
 \
-This lack of a response to elevated C~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons.
+This lack of a response to eC~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons.
 First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low LAI stands were more responsive to eC~a~ than stands with high LAI.
@@
 Stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
-*in press*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations [@Crous2015].
+*in press*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations.
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.
@@
 [-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
-Although this indicates the canopy stayed green longer in elevated C~a~, Gielen et al.
-[-@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ at the PopFACE, leading to similar LAI in ambient C~a~ despite higher productivity.
+Although this indicates the canopy stayed green longer in eC~a~, Gielen et al.
+[-@Gielen2001] reported faster turnover rate of leaves in eC~a~ at the PopFACE, leading to similar LAI in aC~a~ despite higher productivity.
 Li et al.
@@
 Clearly, no consistent picture emerges from these comparisons.
-
-## Continuous measurement of LAI
-Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
-Unlike in deciduous canopies, litter production cannot be used to estimate LAI, because LL is unknown (instead, we used LAI and litter production to estimate LL).
-We cannot not rely either on leaf area estimates from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
-tereticornis*, and would not be valuable anyway since LAI is temporally highly variable (Figs.
-1, 7).
-We found that indirect estimates of LAI based on canopy transmittance were very reliable because relatively small temporal changes in LAI could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
-1, 5, 7).
-We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to those during periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999].
-As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig.
-S3), because longer path lengths in the winter result in lower overall transmittance.
-Using diffuse PAR alone avoids having to correct for these path length effects.
 
@@
 4).
-The consequence is that leaf turnover is highly variable in time (Fig.
-4).
+The consequence is that leaf turnover is highly variable in time.
+
+## Continuous measurement of LAI
+Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
+Unlike in deciduous canopies, litter production cannot be used to estimate LAI, because LL is unknown (instead, we used LAI and litter production to estimate LL).
+We could also not estimate leaf area from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
+tereticornis*, and would not be valuable anyway since LAI is temporally variable (Fig.
+6).
+We found that indirect estimates of LAI based on canopy transmittance were very reliable because relatively small temporal changes in LAI could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
+1, 4, 6).
+We do argue that the raw data must be filtered by diffuse conditions, as the spatial and within-day variance are negligible compared to those during periods of direct sunshine [@Chason1991; @Dufrene1995; @Nilson1999].
+As expected, when we used unfiltered canopy transmittance data, the seasonal patterns in canopy transmittance were much more distinct (Fig.
+S3), because longer path lengths in the winter result in lower overall transmittance.
+Using diffuse PAR alone avoids having to correct for these path length effects.
 
@@
 Thus a minimum temperature is also required for leaf growth to occur.
-We also have not observed leaf flushing in winter time, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig.
+We also have not observed leaf flushing in winter time during the study period, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig.
 6).
-We argue that both minimum temperature and high soil moisture are required  for leaf growth of subtropical eucalypts such as *E.
-tereticornis*.
 
@@
 Simulations at the EucFACE site using the set of ecosystem models in De Kauwe et al.
-[-@DeKauwe2014] also showed little response to inter-annual variation in rainfall (pers.
-comm.
-Martin De Kauwe; Medlyn et al.
-*in prep*).
+[-@DeKauwe2014] also showed little response to inter-annual variation in rainfall (Medlyn et al.
+*in review*).
 In most models  that provide prognostic estimates of LAI [i.e., excluding models that take LAI as an input parameter; e.g.
@@
 lmer_LAI[,"p-value"] <- formatPval(lmer_LAI[,"p-value"])
-
+lmer_LAI[,"F"] <- round(lmer_LAI[,"F"],2)
 kable(lmer_LAI)
@@
 **Figure 1**.
-(a) Target [CO~2~] (C~a~) relative to the ambient C~a~ treatment at the EucFACE.
-The elevated C~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in February 2013.
+(a) Target [CO~2~] (C~a~) relative to the aC~a~ treatment at the EucFACE.
+The eC~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in February 2013.
 (b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions (F~diff~ > 0.98) only.
 The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model).
-Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~.
+Lines in blue colours denote aC~a~ treatment, red colours are eC~a~.
 Note the steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
-The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate LAI estimates to litter production (see Methods).
+The arrow labelled 'calibration' indicates a drought period when $\tau_d$ steadily increased, used to calibrate LAI estimates to litter production (see Methods).
 
@@
 **Figure 3**.
-Comparison of absolute changes in LAI ($\Delta L$) to litter production (in units of LAI), calculated for each litter collection period (approx.
-monthly).
-To calculate $\Delta L$, we fit a generalized additive model fitted to data from each FACE ring separately, and estimated the change in L during the litter collection period (and expressed as monthly changes, by accounting for the differences in actual lengths of litter collection periods).
-Green symbols indicate periods during which $\Delta L$ was positive (net leaf production), orange symbols when $\Delta L$ was negative (net leaf loss).
+Comparison of absolute changes in LAI ($\Delta$LAI) to litter production (P~L~), calculated for each litter collection period.
+Green symbols (circles) indicate periods during which $\Delta L$ was positive (net leaf production), orange symbols (triangles) when $\Delta$LAI was negative (net leaf loss).
 Solid black lines are 1:1 lines, coloured lines regression lines with 95% confidence intervals.
-The strong positive correlation between litter production and $\Delta L$ during periods of net leaf production indicates that during leaf flushing, old leaves are actively shed.
 
@@
 **Figure 4**.
-Timeseries of litter production and estimated gross leaf production rate.
+Timeseries of litter production (P~L~) and estimated leaf production rate.
 Leaf production rate was estimated as the sum of the net change in LAI and litter production (converted to a per area basis using specific leaf area estimates, see Methods).
@@
 (a) Total litter production (converted from data in Fig.
-5 to an annual estimate) for each of the FACE rings plotted against mean $LAI_\tau$ during the entire litter collection period.
-The ratio of mean $LAI_\tau$ to annual litter production gives the mean leaf life span.
+5 to an annual estimate) for each of the FACE rings plotted against mean LAI during the entire litter collection period.
+The ratio of mean LAI to annual litter production gives the mean leaf life span.
 Grey lines indicate leaf lifespan isolines (in years).
@@
 Dynamics of LAI in relation to environmental drivers.
-(a) Smoothed time series of $LAI_\tau$ averaged across all six FACE rings.
-(b) From panel (a), we calculated the daily change in LAI (dL/dt), including the 95% confidence interval for the derivative of the time series.
+(a) Smoothed time series of LAI averaged across all six FACE rings.
+(b) The daily change in LAI (dL/dt), including the 95% confidence interval.
 Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05).
@@
 
-**Figure S2**. Correlation between L estimated from diffuse canopy transmittance (LAI) and from canopy photos.
+**Figure S2**. Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
 
-**Figure S3**. Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
+**Figure S3**. Correlation between LAI estimated from diffuse canopy transmittance and from canopy photos.
 
-**Figure S4**. Average $LAI_\tau$ over the entire study period plotted against the pre-treatment basal area (sum of stem cross-sectional areas) of the FACE rings.
+**Figure S4**. Average LAI over the entire study period is correlated with pre-treatment basal area (sum of stem cross-sectional areas) of the FACE rings.
 

### 2015-08-19 manuscript_supportinginfo.Rmd
Many minor fixes

@@
 Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately.
-Blue colours are ambient, red colours elevated C~a~.
+Blue colours are ambient (nrs 2,3,6), red colours elevated C~a~ (nrs 1,4,5).
 (a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
-(b) Mean litter production expressed in terms of LAI for each ring.
+(b) Total litter production during the drought expressed in terms of LAI for each ring.
 The confidence interval was calculated based on the 8 litter traps in each ring.
@@
 We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.
-LAI from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq.
+LAI from canopy photos (LAI~p~) was calculated with a simple Beer-Lambert law (Eq.
 4).
@@
 [-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig.
-S2).
+Without further calibration, the two estimates were reasonably well correlated, though on average LAI$_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig.
+S3).
 We also estimated the extinction coefficient that minimized the difference between the two estimates of LAI (Fig.

### 2015-08-19 manuscript.Rmd
New SI figure on detectable difference.

@@
 To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate.
-Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in LAI.
+Using this method, we estimated that we can detect a difference of only a 6.1% in LAI (see Fig.
+S5).
 Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.
-
-** rethink the following; CI on least-square means accounting for BA?**
-But what difference between C~a~ treatments could we have reasonably detected?
-The average half-width of the confidence interval on the LAI anomaly (Fig.
-4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment.
-This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean LAI of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).
 
@@
 Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05.
+Variation between rings was substantial, and largely due to pre-treatment differences in stand basal area (see Fig.
+S3 and S5).
 
@@
 
+**Figure SI5**. Detectable differences in LAI between C~a~ treatments after accounting for pre-treatment variation in basal area.
+
 # References {.unnumbered}

### 2015-08-19 manuscript_supportinginfo.Rmd
New SI figure on detectable difference.

@@
 source("manuscript_calcs.R")
+opts_chunk$set(dpi=300)
 ```
@@
 
+\
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
+res <- figureSI5(df, ba)
+```
+
+**Figure SI5**.
+Detectable differences in LAI between C~a~ treatments after accounting for pre-treatment variation in basal area.
+We fit LAI with a fitted GAM using C~a~, time and pre-treatment basal area (BA) as the covariates.
+In the GAM, time was entered as a smooth term (estimated by a cubic spline), and BA as a linear term.
+Shown are the predictions (smooth lines) from the fitted GAM when BA was set to the mean value across all plots.
+Shaded areas indicate 95% confidence intervals (CI).
+Note they are much narrower than the CI across plots when BA was allowed to vary (Fig.
+2 in main text).
+The CI here thus accounts for the pre-treatment differences between rings, and visualizes how using BA as a covariate in the mixed-effects model increases the ability to resolve small difference in LAI that may develop over time.
+The minimum detectable difference in LAI between C~a~ treatments can be estimated from the average half-width of the CI of fitted LAI at a common BA.
+A significant effect would could be observed (provided the error variance remains the same) when the two CIs (for aC~a~ and eC~a~) would just not overlap (so that p < 0.05), or twice the half-width CI.
+This difference was estimated as `r round(res$halfci,2)` m^2^ m^-2^, which when expressed relative to the mean LAI of `r round(mean(ba$LAI.mean),2)` is a difference of only `r round(100*res$minobsdiff,1)`%.
+
 ## Notes S1. Estimation of leaf area index from digital photography

### 2015-08-19 R/figures.R

@@ -347,8 +347,8 @@ figureSI2 <- function(df){
   par(mfrow=c(1,2), mar=c(5,5,2,2), cex.axis=0.9, tcl=0.2)
   plotit <- function(df){
     with(df, plot(LAI, LAI.PAR.mean, 
-                  ylab=expression(italic(L)~from~tau[d]~~(m^2~m^-2)),
-                  xlab=expression(italic(L)~from~canopy~photos~~(m^2~m^-2)),
+                  ylab=expression(LAI~from~tau[d]~~(m^2~m^-2)),
+                  xlab=expression(LAI~from~canopy~photos~~(m^2~m^-2)),
                   pch=19, col=my_ringcols()[Ring],
                   panel.first=predline(lm(LAI.PAR.mean ~ LAI, data=df), lty=5),
                   xlim=c(0.8,2.2), ylim=c(0.8,2.2)))
@@ -406,12 +406,68 @@ figureSI4 <- function(df){
   par(mar=c(5,5,2,2), cex.lab=1.2,xaxs="i", yaxs="i", tcl=0.2, las=1)
   with(df, plot(BA, LAI.mean, pch=19, cex=1.2, col=my_ringcols(),
                 xlab=expression(Basal~area~~(m^2~ha^-1)),
-                ylab=expression(bar(italic(LAI))~~(m^2~m^-2)),
+                ylab=expression(bar(LAI)~~(m^2~m^-2)),
                 panel.first=predline(lm(LAI.mean ~ BA, data=ba)),
                 ylim=c(1.2,2.4), xlim=c(18,40)))
   legend("bottomright", as.character(1:6), pch=19, bty='n',
          col=my_ringcols(), title="Ring", cex=0.6, pt.cex=1)
   
+}
+
+figureSI5 <- function(df, ba){
+  
+  df$Date <- as.numeric(df$Date) # for gamm
+  df <- merge(df, ba)
+  
+  # Fit full model
+  g <- gamm(LAI ~ treatment + BA + s(Date, k=18), random = list(Ring=~1), data=df)
+  
+  
+  # Now predict from model at mean basal area.
+  nd <- expand.grid(Date=seq(min(df$Date), max(df$Date), by=1),
+                    treatment=c("ambient","elevated"))
+  nd$BA <- mean(ba$BA)
+  
+  p <- predict(g[[2]], nd, se.fit=TRUE)
[…]

### 2015-08-19 make_figures.R

@@ -51,6 +51,10 @@ to.pdf(figureSI4(ba),
        filename="output/figures/FigureSI4.pdf",
        width=4, height=4)
 
+to.pdf(figureSI5(df,ba), 
+       filename="output/figures/FigureSI5.pdf",
+       width=8, height=4)
+
 
 # 
 # 

### 2015-08-19 manuscript.Rmd
Edit to reviewer's response; abstract minor edit; figure 4 thicker lines

@@
 Contrary to expectation, LAI did not respond to elevated [CO~2~].
-We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period, but did not show a clear seasonal pattern.
-Instead, dynamics of LAI showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
+We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period.
+The dynamics of LAI showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
 Leaf litter production was correlated to changes in LAI, both during periods of decreasing LAI (when no leaf growth occurred), and during periods of increasing LAI (active shedding of old foliage when new leaf growth occurred).

### 2015-08-19 review1.Rmd
Edit to reviewer's response; abstract minor edit; figure 4 thicker lines

--- 2015-08-18 review1.Rmd
@@
 ---
+
+**We thank the reviewer for the detailed and constructive comments.
+We have redone the data analysis to account for the reviewer's main concern, and provide detailed justification for the litter trap sampling design in the response below.
+We have accounted for the vast majority of detailed comments and suggestions, and have substantially edited the text for consistency, clarity and brevity.**
+
 In this manuscript, Duursma et al.
@@
 There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect.
-We only show the results when including BA as a covariate in the manuscript.
+On this point, we have also added a new figure in the supporting information (Fig.
+S5) that addresses the minimum detectable difference with a new method.
 Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**
@@
 **As noted above, we have redone the analysis and find no CO2 or CO2 x time effect.
-Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect.
-For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect).**
+Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly in the previous manuscript, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect.
+For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect), since it does not confound the overall findings.**
 
@@
 
-**The size, number, and relative coverage of the litterfall traps is broadly consistent with other forest FACE studies (see Table below), with the largest trap size.
+**The size, number, and relative coverage of the litterfall traps at EucFACE is consistent with the design of other forest FACE studies (see Table below), with in fact the largest trap size compared to other sites.
 The size and coverage of the traps is therefore fully justified.
-The litter traps were placed in somewhat random locations across the ring, but never within XXX meters of the plot edge, to minimize contamination.
+The litter traps were placed in somewhat random locations across the ring, but never within 2 meters of the plot edge, to minimize contamination.
 We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**
@@
 
-tabl <- tabl[,-ncol(tabl)]
+tabl[,ncol(tabl)] <- 100*tabl[,ncol(tabl)]
 
@@
 
-**We have moved this analysis, including the methods, to the Supporting Information.
+**We have moved this analysis, including the methods, in its entirety to the Supporting Information.
 We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only.
@@
 
-*TODO* **We have added a couple of sentences indicating the expectation of the litter response; it follows logically from the test of differences in LAI**
+**Since the reviewer deems the manuscript too lengthy already, we hesitate to include a full introduction of litter fall and our expectations.
+We have edited the final paragraph to make it more clear that we are also interested in the components of LAI dynamics, namely, leaf growth and litter production.**
 
@@
 
-**We have moved all methods-related content from the Results to the Methods section. We respond further to specific comments related to the Methods below**
+**We have moved all methods-related content from the Results to the Methods section, and substantially edited the entire section for brevity and clarity.
+We respond further to specific comments related to the Methods below.**
 
@@
 
-**Reworded to 'subtropical', which is more fitting** *IS THIS CORRECT*
+**Reworded to 'subtropical', to be consistent with Methods.**
 
@@
 
-**todo**
+**Removed from abstract.**
 
@@
 
-**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is quite obvious from the statement that the woodland is Eucalyptus-dominated**
+**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is hopefully obvious from the statement that the woodland is Eucalyptus-dominated.**
 
@@
 
-**Removed info on precipitation and PET here, as it is explained in more detail in the Methods already.**
+**Removed info on precipitation and PET here, as it is explained in more detail in the Methods already.
+In the Methods, we have indicated the period over which P and PET are estimated (both are longterm).**
 
@@
 
-**If LAI changes, this must be due to changes in either the production of new foliage, or the lifespan of existing foliage.
-Both components can be assessed with measurements of litter production in combination with the time dynamics of LAI.** *Expand intro on litter fall a bit!!*
+**If LAI changes over time between treatments, this must be due to changes in either the production of new foliage, or the lifespan of existing foliage.
+Both components can be assessed with measurements of litter production in combination with the time dynamics of LAI.**
 
@@
 
-**todo. which part is difficult???**
+**I HAVE NOT DONE THIS - CAN SOMEONE TELL ME WHAT IS DIFFICULT ABOUT IT???**
 
@@
 
-**We have moved the discussion of discernable effect sizes to the discussion.**
+**We have moved the discussion of discernable effect sizes to the discussion, and the details of that analysis to the Supporting Information (new Fig. S5).**
 
@@
 
-**Moved to methods.
+**Moved to methods (and edited for clarity).
 We disagree that this method 'assumes litter is grown and lost in the same time period'.
@@
 In that case the canopy must have produced new leaves in that period to complete the sum of dLAI/dt = leafgrowth - litter.
-There are no assumptions on when the leaves were produced that show up in the litter traps at that time.**
-
-**We think that the term 'litter production' may be misleading, as the term 'production' seems to indicate that the leaves were 'produced' during that time period, which we certainly do not assume nor imply (in fact, we have shown that leaf lifespan is ca.
-1 year).**
+There are no assumptions on when the leaves were produced that show up in the litter traps at that time.
+We now explain this mass balance consideration in the Methods.**
 
@@
 
-**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant discussion to the results of this study to the other FACE sites.
+**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant to compare the results of this study to the other FACE sites.
 This is informative because the results from the EucFACE do not fit well with a generalization across other FACE sites as presented by Norby and Zak (2011).**
@@
 
-**We have moved Fig. 2 to the Supporting Information. We have retained the other figures as we feel they are essential to the manuscript (and six figures does not seem excessive to us).**
+**We have moved Fig. 2 to the Supporting Information. We have retained the other figures as we feel they are essential to the manuscript.**
 
@@
 
-**todo**
+**Done.**

### 2015-08-19 R/figures.R

@@ -141,13 +141,14 @@ figure4 <- function(dLAIlitter){
   
   par(mar=c(3,5,2,2), cex.lab=1.1,tcl=0.2,las=1)
   with(subset(da, treatment == "ambient"), plot(Date, laprod, type='l', col="blue",
+                                                lwd=2,
                                                 axes=FALSE,xlab="",
                                                 ylab=expression("Leaf or litter production"~(m^2~m^-2~mon^-1)),
                                                 ylim=c(-0.05,0.8)))
-  with(subset(da, treatment == "elevated"), lines(Date, laprod, col="red"))
+  with(subset(da, treatment == "elevated"), lines(Date, laprod, col="red", lwd=2))
   
-  with(subset(da, treatment == "ambient"), lines(Date, lit, col="blue", lty=5))
-  with(subset(da, treatment == "elevated"), lines(Date, lit, col="red", lty=5))
+  with(subset(da, treatment == "ambient"), lines(Date, lit, col="blue", lty=5, lwd=2))
+  with(subset(da, treatment == "elevated"), lines(Date, lit, col="red", lty=5, lwd=2))
   abline(h=0, col="darkgrey")
   axis(2)
   box()

### 2015-08-19 manuscript.Rmd
refer to Table1; gimeno in press reference added

@@
 A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig.
-3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p > 0.5).
+3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p > 0.5) (Table 1).
 
@@
 Stimulation of net leaf photosynthesis by eC~a~ (a common response, also observed at the EucFACE, Gimeno et al.
-*in press*) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations.
+[-@gimeno2015]) may not lead to increased net leaf growth, due to other constraints such as water or nutrient limitations.
 Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool.

### 2015-08-19 review1.Rmd
refer to Table1; gimeno in press reference added

@@
 It would be better to frame this as quantitative water effects being relatively unknown as opposed to qualitative effects being well-known.
-I would just drop l.79-81, and start with “The response of L to fluctuations in water availability….” Issues like this need to be addressed throughout the paper.
+I would just drop l.79-81, and start with “The response of L to fluctuations in water availability.” Issues like this need to be addressed throughout the paper.
 

### 2015-08-20 R_other/lerp_and_lai.R
couple figures of lerp abundance compared with LAI; sent to andrew and markus

@@ -26,7 +26,7 @@ litprev <- do.call(rbind, lapply(split(litdf, litdf$Ring), function(x){
 lerplai <- merge(lerplai, litprev)
 
 lerplai$leafgrowth <- with(lerplai, dLAI + dLAIlitter.mean)
-
+lerplai$leafgrowth_prev <- with(lerplai, dLAI_prev + dLAIlitter.mean_prev)
 
 write.csv(lerplai, "output/data/lerplai.csv", row.names=FALSE)
 
@@ -43,17 +43,66 @@ with(lerplai, plot(dLAIlitter.mean_prev, log10(abundance_m2), col=as.factor(spec
                    pch=pchs[species]))
 
 
-windows(7,7)
-par(mfrow=c(3,1), mar=c(0,0,0,0), oma=c(5,5,2,2))
-with(lerplai, plot(Date, leafgrowth))
-with(lerplai, plot(Date, dLAIlitter.mean))
-with(lerplai, plot(Date, log10(abundance_30days), pch=19, col=species))
+
+speccols <- c("forestgreen","darkorange","royalblue")
+treatcols <- c("blue","red")
+lerplai$treatment <- as.factor(lerplai$treatment)
+
+windows(7,9)
+par(mfrow=c(4,1), mar=c(0,0,0,0), oma=c(5,5,2,2))
+with(lerplai, plot(Date, LAI, pch=19, col=treatcols[treatment], axes=FALSE))
+timeseries_axis(labels=FALSE)
+axis(2)
+box()
+CO2legend("topleft")
+with(lerplai, plot(Date, leafgrowth, pch=19, col=treatcols[treatment], axes=FALSE))
+timeseries_axis(labels=FALSE)
+axis(2)
+box()
+with(lerplai, plot(Date, dLAIlitter.mean,pch=19, col=treatcols[treatment], axes=FALSE))
+timeseries_axis(labels=FALSE)
+axis(2)
+box()
+with(lerplai, plot(Date, log10(abundance_30days), pch=15, col=speccols[species],
+                   axes=FALSE))
+timeseries_axis(labels=TRUE)
+axis(2)
+box()
+legend("topleft", levels(lerplai$species), col=speccols, pch=15, cex=0.8)
+
+mtext(side=2, line=3, at=0.125, text="log10 Abundance", outer=TRUE)
[…]

### 2015-08-20 manuscript.Rmd
changed LAI back to L

@@
 
-Canopy leaf area, quantified by the leaf area index (LAI), is a crucial driver of forest productivity, water use, and energy balance.
-Because LAI responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified.
+Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance.
+Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified.
 We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a subtropical climate where water limitation is common.
-We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, LAI should increase with elevated [CO~2~].
-We estimated LAI from diffuse canopy transmittance, and measured monthly leaf litter production.
-Contrary to expectation, LAI did not respond to elevated [CO~2~].
-We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period.
-The dynamics of LAI showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
-Leaf litter production was correlated to changes in LAI, both during periods of decreasing LAI (when no leaf growth occurred), and during periods of increasing LAI (active shedding of old foliage when new leaf growth occurred).
-Leaf lifespan, estimated from mean LAI and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs.
+We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, *L* should increase with elevated [CO~2~].
+We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production.
+Contrary to expectation, *L* did not respond to elevated [CO~2~].
+We found that *L* varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period.
+The dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall.
+Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred).
+Leaf lifespan, estimated from mean *L* and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs.
 `r f(LLtest$estimate1)` yrs; p `r pval(LLtest$p.value)`).
-Our main finding that LAI was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of LAI commonly predicted by many ecosystem models.
+Our main finding that *L* was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of *L* commonly predicted by many ecosystem models.
 
@@
 
-Leaf area index (LAI) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
-LAI is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
-Ecosystem models generally predict that LAI increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
-This potential change in LAI is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes between vegetation and atmosphere as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
-Despite its importance, little empirical evidence is available on the response of LAI to eC~a~ in mature forests [@Norby2011].
+Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014].
+*L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].
+Ecosystem models generally predict that *L* increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012].
+This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes between vegetation and atmosphere as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012].
+Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].
 
 \
-Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in LAI.
+Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*.
 Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in water-use efficiency [@Ainsworth2005; @Ellsworth2012].
-Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state LAI to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
-Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased LAI in response to eC~a~ [@McCarthy2007; @Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
-Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment LAI and the responsiveness to eC~a~, leading to the prediction that stands with low LAI will show a larger response to eC~a~.
+Although numerous studies have examined the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011].
+Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased *L* in response to eC~a~ [@McCarthy2007; @Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005].
+Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.
 
 \
-In ecosystems where water availability limits productivity, we expect an especially large response of LAI to eC~a~ [@Donohue2009].
+In ecosystems where water availability limits productivity, we expect an especially large response of *L* to eC~a~ [@Donohue2009].
 This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability.
-These 'water savings' can be expected to lead to an increase in LAI in water-limited ecosystems, consistent with a tight coupling of LAI with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
+These 'water savings' can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
 In support of this prediction, Donohue et al.
-[-@Donohue2013] attributed the observed global increase in LAI in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~].
-Based on the same water-savings principle, we predicted that LAI should increase under eC~a~ in water-limited forests.
+[-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~].
+Based on the same water-savings principle, we predicted that *L* should increase under eC~a~ in water-limited forests.
 
 \
-The responses of LAI to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
-The strong relationship between LAI and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
-However, local LAI is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011].
-Understanding dynamics of LAI in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
+The responses of *L* to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified.
+The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions.
+However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011].
+Understanding dynamics of *L* in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].
 
 \
-We estimated LAI from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
-The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of LAI and its components (litter fall and production) in response to water availability.
-We hypothesize that eC~a~ would result in an increase in LAI, particularly during periods of water limitation.
-In addition, we expect that the low LAI that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of LAI to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
-Our second objective is to quantify the temporal dynamics of LAI and its components (litter fall and leaf production), in particular in relation to fluctuating water availability.
+We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
+The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L* and its components (litter fall and production) in response to water availability.
+We hypothesize that eC~a~ would result in an increase in *L*, particularly during periods of water limitation.
+In addition, we expect that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of *L* to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
+Our second objective is to quantify the temporal dynamics of *L* and its components (litter fall and leaf production), in particular in relation to fluctuating water availability.
 
@@
 ## Terminology
-In the following, we use the term leaf area index (LAI) although the canopy and FACE rings contain non-leaf shading elements, i.e.
+In the following, we use the term leaf area index (*L*) although the canopy and FACE rings contain non-leaf shading elements, i.e.
 woody stems and infrastructure.
-This avoids the need to switch between different terminologies when comparing temporal changes in LAI to litter production, and absolute values between different methods.
-The temporal changes in LAI are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.
+This avoids the need to switch between different terminologies when comparing temporal changes in *L* to litter production, and absolute values between different methods.
+The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.
 
@@
 Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
-Litter production was converted to units of LAI by multiplying it with an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
+Litter production was converted to units of *L* by multiplying it with an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
 We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.
 
-## LAI from diffuse transmittance
+## *L* from diffuse transmittance
 In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
@@
 5 m above the canopy.
-For determination of LAI, we used only highly diffuse conditions, where F~diff~ > 0.98.
+For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
 We further trimmed the data by excluding low solar elevation (< 10^o^), because measured canopy transmittance was more variable at these low solar angles.
@@
 \
-We then estimated leaf area index (LAI) from $\tau_d$ following Campbell and Norman [-@Campbell2000].
+We then estimated leaf area index (*L*) from $\tau_d$ following Campbell and Norman [-@Campbell2000].
 We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
-We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the LAI estimate to leaf litter production as further detailed below.
-Diffuse transmittance of a canopy is a function of the LAI by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).
+We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below.
+Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).
 
@@
 
-where L is LAI, $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
+where L is *L*, $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
 Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006].
-We solved numerically for LAI in Eq.
-2, using a smooth integration over $\theta$, yielding LAI.
+We solved numerically for *L* in Eq.
+2, using a smooth integration over $\theta$, yielding *L*.
 
@@
 To overcome violations in the assumptions underlying Eq.
-2, we calibrated LAI against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
+2, we calibrated *L* against cumulative litter production during a 4-month drought period in 2013 (indicated in Fig.
 1).
-During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but LAI decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
+During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
 1).
@@
 A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
-Cumulative LAI leaf litter production ($P_L$, expressed in units of LAI) should match the observed decrease in LAI estimated from canopy transmittance.
+Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance.
 We calibrated Eq.
@@
 
-where *c* is a calibration parameter, $L_{\tau,t=0}$ is LAI at the start of the drought period.
+where *c* is a calibration parameter, $L_{\tau,t=0}$ is *L* at the start of the drought period.
 We used linear regression without intercept to estimate *c* in Eq.
 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`).
-It should be stressed that even after calibration with litter production, LAI still includes stem and infrastructure area (see Terminology).
+It should be stressed that even after calibration with litter production, *L* still includes stem and infrastructure area (see Terminology).
 However, this approximately constant area cancels out in the left-hand side of Eq.
@@
 \
-We used the opportunity provided by the high quality timeseries of LAI to test a simple photography-based method suggested by Macfarlane et al.
+We used the opportunity provided by the high quality timeseries of *L* to test a simple photography-based method suggested by Macfarlane et al.
 [-@Macfarlane2007].
 The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Notes S1, Fig.
-S3), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate LAI.
+S3), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
 
 ## Data analysis
-We used generalized additive models (GAM) for all analyses of LAI and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R version `r r_version` [@RCT2015].
+We used generalized additive models (GAM) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R version `r r_version` [@RCT2015].
 A GAM is a semi-parametric fitting method that allows the estimation of smoothed trends in the data, including confidence intervals.
 In all fitted GAMs, we used a cubic regression spline.
-For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed LAI dynamics.
+For the smoothed term in the model (days since start of experiment), we used up to `r .gambase` degrees of freedom, which resulted in biologically realistic smoothed *L* dynamics.
 Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
-Random effects were included to account for the repeated measures structure of the data, except when we analysed average LAI across all rings (Fig.
+Random effects were included to account for the repeated measures structure of the data, except when we analysed average *L* across all rings (Fig.
 6).
@@
 \
-To test for effects of C~a~ treatment on LAI and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
+To test for effects of C~a~ treatment on *L* and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C~a~ treatment and date as factor variables.
 The random effect accounts for the repeated measures nature of the data.
-For LAI, we further used pre-treatment basal area (BA) of each FACE ring as a covariate (see Fig.
+For *L*, we further used pre-treatment basal area (BA) of each FACE ring as a covariate (see Fig.
 S4).
@@
 \
-To compare the dynamics of LAI with litter production, we calculated the change in LAI for every time period when litter was collected.
-We did this by fitting a GAM to LAI for each ring separately (similar to that shown for $\tau_d$ in Fig.
-1), and calculating the change in LAI for each litter fall collection period ($\Delta$LAI).
-Next, we estimated gross leaf production (G~L~) as the sum of the net change in LAI ($\Delta$LAI) and P~L~, because mass balance requires that $\Delta$LAI = G~L~ - P~L~.
-We estimated mean leaf lifespan (LL, yr) as the ratio of mean LAI (m^2^ m^-2^) for the entire litter fall collection period to total P~L~ (m^2^ m^-2^ yr^-1^).
+To compare the dynamics of *L* with litter production, we calculated the change in *L* for every time period when litter was collected.
+We did this by fitting a GAM to *L* for each ring separately (similar to that shown for $\tau_d$ in Fig.
+1), and calculating the change in *L* for each litter fall collection period ($\Delta$*L*).
+Next, we estimated gross leaf production (G~L~) as the sum of the net change in *L* ($\Delta$*L*) and P~L~, because mass balance requires that $\Delta$*L* = G~L~ - P~L~.
+We estimated mean leaf lifespan (LL, yr) as the ratio of mean *L* (m^2^ m^-2^) for the entire litter fall collection period to total P~L~ (m^2^ m^-2^ yr^-1^).
 
@@
 We quantified the relationship between timing of leaf flushing and environmental drivers.
-During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of LAI were available.
+During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of *L* were available.
 These events occurred in spring and summer, with no clear seasonal cycle (October 2012, January 2013, December 2013 and January 2015).
-These leaf growth events were compared to the dynamics of LAI as estimated from a GAM fitted to the average daily LAI across the six FACE rings.
-From the fitted model, we estimated the numerical derivative of LAI (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014, see also @Curtis2014].
-
-\
-The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXX), published on figshare (doi:XXX).
-The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, using the 'rmarkdown' R package for conversion).
+These leaf growth events were compared to the dynamics of *L* as estimated from a GAM fitted to the average daily *L* across the six FACE rings.
+From the fitted model, we estimated the numerical derivative of *L* (dL/dt), as well as its confidence interval following Simpson [-@Simpson2014, see also @Curtis2014].
 
@@
 Andrew Gherlenda, Markus Riegler].
-This rapid decrease in LAI was evident in a quick increase in $\tau_d$ (Fig.
+This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
 1).
@@
 \
-We found no differences in LAI between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig.
+We found no differences in *L* between the C~a~ treatments, as shown by consistent overlap of the 95% confidence intervals of the generalized additive model (thus, p > 0.05 at any point in time, Fig.
 2).
 A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig.
-3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on LAI (p > 0.5) (Table 1).
+3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on *L* (p > 0.5) (Table 1).
 
 \
-Litter production (P~L~) correlated significantly with $\Delta$LAI when the latter was separated into periods where $\Delta$LAI > 0 (net leaf growth) and $\Delta$LAI <  0 (net leaf loss) (Fig.
+Litter production (P~L~) correlated significantly with $\Delta$*L* when the latter was separated into periods where $\Delta$*L* > 0 (net leaf growth) and $\Delta$*L* <  0 (net leaf loss) (Fig.
 3).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where LAI was decreasing (see also Fig.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where *L* was decreasing (see also Fig.
 6).
-During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta$LAI and litter production (Fig.
+During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta$*L* and litter production (Fig.
 4).
@@
 \
-Next, we related total P~L~ to average LAI (during the litter collection period) across the six FACE rings (Fig.
+Next, we related total P~L~ to average *L* (during the litter collection period) across the six FACE rings (Fig.
 5a).
-If leaf lifespan (LL) is constant, P~L~ would be linearly related to LAI.
-Instead, we found that in eC~a~, P~L~ was lower at a given LAI as compared to aC~a~ (Fig.
+If leaf lifespan (LL) is constant, P~L~ would be linearly related to *L*.
+Instead, we found that in eC~a~, P~L~ was lower at a given *L* as compared to aC~a~ (Fig.
 6a).
@@
 \
-The dynamics of LAI from the sensor record matched leaf growth events observed in the field quite well (Fig.
+The dynamics of *L* from the sensor record matched leaf growth events observed in the field quite well (Fig.
 6a).
@@
 > 0.2 m^3^ m^-3^.
-Two periods of sustained decrease in LAI (red lines in Fig.
+Two periods of sustained decrease in *L* (red lines in Fig.
 6b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.
@@
 
-## Response of LAI to elevated [CO~2~]
-We report on `r nummon` months of continuous leaf area index (LAI) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013.
-We predicted that LAI would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability.
-We also expected that this eucalypt woodland, with its inherently low LAI [@Whitehead2004], would be even more responsive to eC~a~ than higher-LAI and more closed-canopy stands [@Norby2011].
+## Response of *L* to elevated [CO~2~]
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013.
+We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability.
+We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011].
 Counter to our expectations but similar to Norby et al.
-[-@Norby2003], we found no significant effect of eC~a~ on LAI.
+[-@Norby2003], we found no significant effect of eC~a~ on *L*.
 Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig.
 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
-It could be argued that the change in LAI in response to eC~a~ was too small to be detected.
-Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
+It could be argued that the change in *L* in response to eC~a~ was too small to be detected.
+Natural variation in *L* among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments.
 To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate.
-Using this method, we estimated that we can detect a difference of only a 6.1% in LAI (see Fig.
+Using this method, we estimated that we can detect a difference of only a 6.1% in *L* (see Fig.
 S5).
@@
 This lack of a response to eC~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low LAI stands were more responsive to eC~a~ than stands with high LAI.
-They argued that high LAI canopies would not benefit from further increases in LAI, because it would lead to only very small increases in total absorbed radiation.
-The mean LAI at the EucFACE was ca.
-`r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest LAI [@Iio2013], and the lowest LAI of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
+First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*.
+They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
+The mean *L* at the EucFACE was ca.
+`r f(mean(ba$LAI.mean))` m^2^m^-2^, which is low compared to a global compendium of forest *L* [@Iio2013], and the lowest *L* of any forest FACE experiment [@Norby2011], but in line with a range of values for native *Eucalyptus* forests and woodlands reported by Whitehead and Beadle [-@Whitehead2004] and Iio et al.
 [-@Iio2013].
-Based on this LAI estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in LAI under eC~a~.
+Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in *L* under eC~a~.
 In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
-Hence, it may be too early to draw general conclusions across FACE sites in terms of LAI response.
+Hence, it may be too early to draw general conclusions across FACE sites in terms of *L* response.
 
@@
 5) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
-Because we found no difference in LAI with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
+Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
 To further understand the response of biomass allocation to eC~a~ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete.
@@
 \
-Despite a lack of response of canopy LAI, we did find a significantly longer LL in eC~a~.
-We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean LAI and total accumulated leaf litter (Fig.
+Despite a lack of response of canopy *L*, we did find a significantly longer LL in eC~a~.
+We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
 5).
@@
 McCarthy et al.
-[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of LAI by eC~a~.
+[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of *L* by eC~a~.
 Taylor et al.
@@
 Although this indicates the canopy stayed green longer in eC~a~, Gielen et al.
-[-@Gielen2001] reported faster turnover rate of leaves in eC~a~ at the PopFACE, leading to similar LAI in aC~a~ despite higher productivity.
+[-@Gielen2001] reported faster turnover rate of leaves in eC~a~ at the PopFACE, leading to similar *L* in aC~a~ despite higher productivity.
 Li et al.
@@
 
-## Litter production and dynamics of LAI
-We found a correlation between LAI dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing LAI (Fig.
+## Litter production and dynamics of *L*
+We found a correlation between *L* dynamics and litter production, when the data were analyzed separately for periods of decreasing and increasing *L* (Fig.
 3).
-As expected, during periods of decreasing LAI, the change in LAI was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of LAI against litter production.
-During periods of increasing LAI, litter fall increased sharply (Figs.
+As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production.
+During periods of increasing *L*, litter fall increased sharply (Figs.
 4 and 5), indicating replacement of old leaves by young leaves.
 Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of December 2013.
-This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net LAI increase, but with a lag between litter fall and production of about 2 months.
+This is in line with Pook [-@Pook1984], who showed in *Corymbia maculata* (a genus closely related to *Eucalyptus*) accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months.
 Similarly, Stone and Bacon [-@Stone1995] showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale.
@@
 
-## Continuous measurement of LAI
+## Continuous measurement of *L*
 Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
-Unlike in deciduous canopies, litter production cannot be used to estimate LAI, because LL is unknown (instead, we used LAI and litter production to estimate LL).
+Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because LL is unknown (instead, we used *L* and litter production to estimate LL).
 We could also not estimate leaf area from allometric relationships with stem diameter or height, since these equations are not available for mature *E.
-tereticornis*, and would not be valuable anyway since LAI is temporally variable (Fig.
+tereticornis*, and would not be valuable anyway since *L* is temporally variable (Fig.
 6).
-We found that indirect estimates of LAI based on canopy transmittance were very reliable because relatively small temporal changes in LAI could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
+We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
 1, 4, 6).
@@
 \
-Although the response of LAI to water availability is qualitatively well known, current-generation phenology models fall short in incorporating this response.
+Although the response of *L* to water availability is qualitatively well known, current-generation phenology models fall short in incorporating this response.
 In a comparison of six leading ecosystem and global vegetation models,  Powell et al.
-[-@Powell2013] showed that the steady decrease in LAI in an Amazon drought experiment could be matched by only one model.
+[-@Powell2013] showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model.
 Simulations at the EucFACE site using the set of ecosystem models in De Kauwe et al.
@@
 *in review*).
-In most models  that provide prognostic estimates of LAI [i.e., excluding models that take LAI as an input parameter; e.g.
-@Kala2013], temporal changes in LAI are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and then allocating some fraction to the leaf pool.
-Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of LAI in response to environmental drivers such as eC~a~.
-We argue that this algorithm is not well suited to explain the dynamic nature of LAI in response to water availability observed in evergreen broad-leaved forests.
+In most models  that provide prognostic estimates of *L* [i.e., excluding models that take *L* as an input parameter; e.g.
+@Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and then allocating some fraction to the leaf pool.
+Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
+We argue that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests.
 In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability, as we have demonstrated that both processes are highly dynamic.
@@
 Additional support from the Australian Research Council is acknowledged.
+The entire analysis presented in this paper will be made available as a *git* repository (http://www.github.com/remkoduursma/XXX), published on figshare (doi:XXX).
+The repository includes the raw data needed to run the analyses, the source code for the analysis and manuscript (which was written in markdown, using the 'rmarkdown' R package for conversion).
 
@@
 **Table 1**.
-Results of a mixed-effects model of LAI with C~a~, Date and pre-treatment basal area (BA) as predictors.
+Results of a mixed-effects model of *L* with C~a~, Date and pre-treatment basal area (BA) as predictors.
 The model accounts for repeated measures by using FACE ring as the random effect.
-Sample size of LAI per ring was `r table(df$Ring)[1]`.
+Sample size of *L* per ring was `r table(df$Ring)[1]`.
 
@@
 Note the steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy.
-The arrow labelled 'calibration' indicates a drought period when $\tau_d$ steadily increased, used to calibrate LAI estimates to litter production (see Methods).
+The arrow labelled 'calibration' indicates a drought period when $\tau_d$ steadily increased, used to calibrate *L* estimates to litter production (see Methods).
 
@@
 **Figure 2**.
-Time series of LAI estimated from canopy transmittance.
+Time series of *L* estimated from canopy transmittance.
 Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect.
@@
 **Figure 3**.
-Comparison of absolute changes in LAI ($\Delta$LAI) to litter production (P~L~), calculated for each litter collection period.
-Green symbols (circles) indicate periods during which $\Delta L$ was positive (net leaf production), orange symbols (triangles) when $\Delta$LAI was negative (net leaf loss).
+Comparison of absolute changes in *L* ($\Delta$*L*) to litter production (P~L~), calculated for each litter collection period.
+Green symbols (circles) indicate periods during which $\Delta L$ was positive (net leaf production), orange symbols (triangles) when $\Delta$*L* was negative (net leaf loss).
 Solid black lines are 1:1 lines, coloured lines regression lines with 95% confidence intervals.
@@
 Timeseries of litter production (P~L~) and estimated leaf production rate.
-Leaf production rate was estimated as the sum of the net change in LAI and litter production (converted to a per area basis using specific leaf area estimates, see Methods).
+Leaf production rate was estimated as the sum of the net change in *L* and litter production (converted to a per area basis using specific leaf area estimates, see Methods).
 Error bars are omitted for clarity.
@@
 (a) Total litter production (converted from data in Fig.
-5 to an annual estimate) for each of the FACE rings plotted against mean LAI during the entire litter collection period.
-The ratio of mean LAI to annual litter production gives the mean leaf life span.
+5 to an annual estimate) for each of the FACE rings plotted against mean *L* during the entire litter collection period.
+The ratio of mean *L* to annual litter production gives the mean leaf life span.
 Grey lines indicate leaf lifespan isolines (in years).
@@
 **Figure 6**.
-Dynamics of LAI in relation to environmental drivers.
-(a) Smoothed time series of LAI averaged across all six FACE rings.
-(b) The daily change in LAI (dL/dt), including the 95% confidence interval.
+Dynamics of *L* in relation to environmental drivers.
+(a) Smoothed time series of *L* averaged across all six FACE rings.
+(b) The daily change in *L* (dL/dt), including the 95% confidence interval.
 Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05).
@@
 
-**Figure S3**. Correlation between LAI estimated from diffuse canopy transmittance and from canopy photos.
+**Figure S3**. Correlation between *L* estimated from diffuse canopy transmittance and from canopy photos.
 
-**Figure S4**. Average LAI over the entire study period is correlated with pre-treatment basal area (sum of stem cross-sectional areas) of the FACE rings.
+**Figure S4**. Average *L* over the entire study period is correlated with pre-treatment basal area (sum of stem cross-sectional areas) of the FACE rings.
 
-**Figure SI5**. Detectable differences in LAI between C~a~ treatments after accounting for pre-treatment variation in basal area.
+**Figure SI5**. Detectable differences in *L* between C~a~ treatments after accounting for pre-treatment variation in basal area.
 

### 2015-08-20 manuscript_supportinginfo.Rmd
changed LAI back to L

@@
 (a) The change in  $\tau_d$ and its 95% confidence interval was calculated from the slope of linear regression of $\tau_d$ with time (number of days since 8 July 2013).
-(b) Total litter production during the drought expressed in terms of LAI for each ring.
+(b) Total litter production during the drought expressed in terms of *L* for each ring.
 The confidence interval was calculated based on the 8 litter traps in each ring.
@@
 **Figure S3**.
-Correlation between LAI estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
+Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
 Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
@@
 **Figure SI4**.
-Average $LAI_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, LAI = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
-This relationship was used to calculate the LAI anomaly in Fig.
-3, as the difference between LAI predicted from this regression and the actual $LAI_\tau$ at any given time.
+Average $LAI_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `r tidy(balm)$estimate[2]` BA).
+This relationship was used to calculate the *L* anomaly in Fig.
+3, as the difference between *L* predicted from this regression and the actual $LAI_\tau$ at any given time.
 Red symbols are elevated C~a~, blue symbols ambient C~a~.
@@
 **Figure SI5**.
-Detectable differences in LAI between C~a~ treatments after accounting for pre-treatment variation in basal area.
-We fit LAI with a fitted GAM using C~a~, time and pre-treatment basal area (BA) as the covariates.
+Detectable differences in *L* between C~a~ treatments after accounting for pre-treatment variation in basal area.
+We fit *L* with a fitted GAM using C~a~, time and pre-treatment basal area (BA) as the covariates.
 In the GAM, time was entered as a smooth term (estimated by a cubic spline), and BA as a linear term.
@@
 2 in main text).
-The CI here thus accounts for the pre-treatment differences between rings, and visualizes how using BA as a covariate in the mixed-effects model increases the ability to resolve small difference in LAI that may develop over time.
-The minimum detectable difference in LAI between C~a~ treatments can be estimated from the average half-width of the CI of fitted LAI at a common BA.
+The CI here thus accounts for the pre-treatment differences between rings, and visualizes how using BA as a covariate in the mixed-effects model increases the ability to resolve small difference in *L* that may develop over time.
+The minimum detectable difference in *L* between C~a~ treatments can be estimated from the average half-width of the CI of fitted *L* at a common BA.
 A significant effect would could be observed (provided the error variance remains the same) when the two CIs (for aC~a~ and eC~a~) would just not overlap (so that p < 0.05), or twice the half-width CI.
-This difference was estimated as `r round(res$halfci,2)` m^2^ m^-2^, which when expressed relative to the mean LAI of `r round(mean(ba$LAI.mean),2)` is a difference of only `r round(100*res$minobsdiff,1)`%.
+This difference was estimated as `r round(res$halfci,2)` m^2^ m^-2^, which when expressed relative to the mean *L* of `r round(mean(ba$LAI.mean),2)` is a difference of only `r round(100*res$minobsdiff,1)`%.
 
@@
 
-We took advantage of the detailed timeseries of LAI estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al.
+We took advantage of the detailed timeseries of *L* estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al.
 [-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests.
@@
 We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold.
-LAI from canopy photos (LAI~p~) was calculated with a simple Beer-Lambert law (Eq.
+*L* from canopy photos (*L*~p~) was calculated with a simple Beer-Lambert law (Eq.
 4).
@@
 [-@Macfarlane2007] recommendation for eucalypt forests.
-Without further calibration, the two estimates were reasonably well correlated, though on average LAI$_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig.
+Without further calibration, the two estimates were reasonably well correlated, though on average *L*$_\tau$ was `r f(100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1))`% higher than $L_p$ (R^2^ = `r f(glance(lmphoto)$r.squared)`, RMSE = `r f(glance(lmphoto)$sigma)`) (Fig.
 S3).
-We also estimated the extinction coefficient that minimized the difference between the two estimates of LAI (Fig.
+We also estimated the extinction coefficient that minimized the difference between the two estimates of *L* (Fig.
 S2), giving *k* = `r f(kopt)`.
-These comparisons are encouraging because the photos measure LAI near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
+These comparisons are encouraging because the photos measure *L* near the zenith (+/- 15^o^), while $L_\tau$ is integrated over the sky hemisphere (Eq.
 2), weighted towards the zenith.

### 2015-08-20 R/figures.R

@@ -22,6 +22,8 @@ converttoJPEGs <- function(){
 # Smoothed gap fraction raw data
 figure1 <- function(df, ramp){
   
+  palette(my_co2cols())
+  
   l <- layout(matrix(c(1,2), ncol=1), heights=c(0.8,2))
   xl <- c(min(ramp$Date), max(df$Date))
   
@@ -64,7 +66,7 @@ figure1 <- function(df, ramp){
 figure2 <- function(df,
                     xlim=NULL, ylim=NULL,
                     legend=TRUE,
-                    ylab=expression(LAI~~(m^2~m^-2)),
+                    ylab=expression(italic(L)~~(m^2~m^-2)),
                     cex.lab=1.1, cex.axis=0.8, cex.legend=0.7,
                     legendwhere="topleft",
                     setpar=TRUE,
@@ -113,7 +115,7 @@ figure3 <- function(df){
                xlab=expression(P[L]~~(m^2~m^-2~mon^-1)),
                xlim=c(0,0.6),
                ylim=c(-0.3,0.5),
-               ylab=expression(Delta*LAI~~(m^2~m^-2~mon^-1))))
+               ylab=expression(Delta*italic(L)~~(m^2~m^-2~mon^-1))))
   with(dfdown, points(X, Y, pch=c(17,24)[treatment],
                   col=Cols[LAIchange]))
   
@@ -124,8 +126,8 @@ figure3 <- function(df){
   predline(lm(Y ~ X, data=dfdown), col=Cols[1])
   box()
   
-  l <- legend("bottomright", c(expression(Delta*italic(LAI) < 0),
-                               expression(Delta*italic(LAI) > 0)), 
+  l <- legend("bottomright", c(expression(Delta*italic(L) < 0),
+                               expression(Delta*italic(L) > 0)), 
               pt.bg=Cols, bty='n', cex=0.8, pch=c(24,21))
   
   legend(l$rect$left - l$rect$w, l$rect$top, 
@@ -169,7 +171,7 @@ figure5 <-  function(df){
   
   with(df, plot(LAI.mean.litterperiod, LAIlitter_annual, pch=19, cex=1.2, col=my_ringcols(),
                 ylab=expression(Litter~production~~(m^2~m^-2~yr^-1)),
-                xlab=expression(bar(italic(LAI))~~(m^2~m^-2)),
[…]

### 2015-08-21 manuscript.Rmd
Edited according to co-authors comments on revision; suppinfo has own template

@@
 \
-In ecosystems where water availability limits productivity, we expect an especially large response of *L* to eC~a~ [@Donohue2009].
+In ecosystems where water availability limits productivity, we expect a larger response of *L* to eC~a~ [@Donohue2009].
 This is because eC~a~ leads to reduced stomatal conductance, and hence water use per unit leaf area [@Warren2010] leading to increased soil water availability.
 These 'water savings' can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008].
-In support of this prediction, Donohue et al.
+In line with this prediction, Donohue et al.
 [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising atmospheric [CO~2~].
-Based on the same water-savings principle, we predicted that *L* should increase under eC~a~ in water-limited forests.
+Based on the same water-savings principle, we predicted that *L* increases under eC~a~ in water-limited forests.
 
@@
 \
-We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment.
+We estimated *L* from canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to Free-Air CO~2~ Enrichment (FACE), the EucFACE experiment.
 The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L* and its components (litter fall and production) in response to water availability.
@@
 In addition, we expect that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allow a positive response of *L* to eC~a~ either via an increase in the rate of leaf production or leaf retention (leaf lifespan).
-Our second objective is to quantify the temporal dynamics of *L* and its components (litter fall and leaf production), in particular in relation to fluctuating water availability.
+Our second hypothesis is that the temporal dynamics of *L* and its components (litter fall and leaf production), are related to to fluctuating water availability.
 
@@
 The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.).
-The vegetation consists of a native woodland undisturbed (unmanaged) for ca.
-70 years and maintained in this location since the 1880's.
 The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (1881 - 2014, Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au).
-An estimate of potential evapotranspiration (PET) at the site (1350 mm) was from the Global-PET database [1950 - 2000 climatology, http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
+Estimated potential evapotranspiration (PET) is 1350 mm [1950 - 2000 climatology, http://www.cgiar-csi.org, @Zomer2008], giving a ratio of P/PET of ca.
 0.59.
@@
 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay.
-The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015].
+The soil is nutrient poor, and a strongly P-limited [@Crous2015].
 Groundwater depth is 12.6 - 12.9 m below the soil surface.
@@
 amplifolia* Naudin.
-The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst.
-& G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg.
-and *Bursaria spinosa* Cav.
+The understory stratum is dominated by grasses, with lower densities of forbs and occasional shrubs.
 
@@
 Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high.
-Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009].
+Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation [@Lewin2009].
 The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received aC~a~, and are used as control plots.
-The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above aC~a~ in full operation mode.
+The target daytime C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above aC~a~ in full operation mode.
 However, C~a~ was increased gradually for six months at a rate of ca.
 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig.
-1a).
+1a, [@gimeno2015]).
 
@@
 This avoids the need to switch between different terminologies when comparing temporal changes in *L* to litter production, and absolute values between different methods.
-The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and woody stem area is very nearly constant over the two-year study period since no tree mortality has occurred.
+The temporal changes in *L* are due to leaf area dynamics only, because infrastructure area is constant, and variation in woody stem area can be assumed negligible since no tree mortality has occurred over the two-year study period.
 
 ## Leaf litter production
-In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at random locations (**can I say this?**) at 0.5m above the forest floor.
-To minimize contamination from trees outside the rings, the baskets were installed at least 2m from the plot edge.
+In each ring, eight litter traps were installed at randomly selected locations.
+The litter traps were 0.1979 m^2^ each with a depth of 0.4 m, and sat at 0.5m above the forest floor.
+Litter traps were custom-made from plastic baskets coated with a fine nylon mesh with a 200 $/mu$m mesh size.
+To minimize contamination with litter from trees outside the rings, the baskets were installed at least 2 m from the plot edge.
 Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc.
 Litter production was converted to units of *L* by multiplying it with an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14).
-We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.
+We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) on any of the collection dates.
 
@@
 In each of the six FACE rings, three photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height.
-Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 October 2012, and automatically uploaded every day to a data management system.
-At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height.
+In each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height.
+Data were logged on data loggers (CR3000, Campbell Scientific) at one minutely intervals beginning on 15 October 2012, and automatically uploaded every day to a data management system.
 All PAR data were averaged at 30-min time steps.
 At a nearby flux tower site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30-min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending ca.
-5 m above the canopy.
+5 m above the forest canopy.
 For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98.
@@
 We then estimated leaf area index (*L*) from $\tau_d$ following Campbell and Norman [-@Campbell2000].
-We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space.
+We assumed that leaves had a spherical leaf angle distribution, and that leaves were randomly distributed in space.
 We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below.
@@
 
-where L is *L*, $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
+where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [@Campbell2000].
 Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006].
@@
 1).
-During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
+During this period (14 July 2013 to 12 Nov 2013), no leaf growth occurred and *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig.
 1).
-During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm).
+During this period total precipitation was 42mm, the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm).
 A total of n = 11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period.
-Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance.
+Cumulative *L* leaf litter production (expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance.
 We calibrated Eq.
@@
 
-where *c* is a calibration parameter, $L_{\tau,t=0}$ is *L* at the start of the drought period.
-We used linear regression without intercept to estimate *c* in Eq.
+where *c* is a calibration parameter, $L_{\tau,t=0}$ and $L_{\tau,t=1}$ are the *L* at the start and the end of the drought period, respectively.
+We used linear regression without an intercept to estimate *c* in Eq.
 4 as `r f(calib)` (SE = `r f(tidy(caliblm)$std.error)`).
@@
 Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data.
-Random effects were included to account for the repeated measures structure of the data, except when we analysed average *L* across all rings (Fig.
-6).
+Random effects were included to account for the repeated measures structure of the data, except when we analysed average *L* across all rings.
 
@@
 2).
-A linear mixed-effects model with C~a~, time and pre-treatment basal area (Fig.
-3) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on *L* (p > 0.5) (Table 1).
+A linear mixed-effects model with C~a~, time and pre-treatment basal area (Table 1) as fixed effects, and FACE ring as a random effect also showed no significant effect of C~a~ on *L* (p > 0.5) (Table 1).
 
@@
 3).
-During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where *L* was decreasing (see also Fig.
+During periods of net leaf loss, we found a significant correlation (with a slope of `r f(tidy(lmlitdown)$estimate[2])`, SE = `r f(tidy(lmlitdown)$std.error[2])`, n = `r glance(lmlitdown)$df.residual+2`, p < 0.001), close to the 1:1 line, as expected since no leaf growth was observed in the field during any period where *L* was decreasing (see also Fig.
 6).
-During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta$*L* and litter production (Fig.
-4).
+During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`, p < 0.001) between $\Delta$*L* and litter production (Fig.
+4), indicating that most shedding of old foliage occurred during periods of active leaf growth.
 
@@
 \
-Next, we related total P~L~ to average *L* (during the litter collection period) across the six FACE rings (Fig.
+Next, we related total P~L~ to average *L* across the six FACE rings (Fig.
 5a).
@@
 Instead, we found that in eC~a~, P~L~ was lower at a given *L* as compared to aC~a~ (Fig.
-6a).
+5a).
 We found a significant (p `r pval(LLtest$p.value)`) increase in LL with eC~a~ (Fig.
@@
 It is also obvious that every leaf flush event occurred shortly after a significant rainfall event (Fig.
-3 and 5), when the soil water content in the top 80 cm increased to ca.
+6b,c), when the soil water content in the top 80 cm increased to ca.
 > 0.2 m^3^ m^-3^.
@@
 ## Response of *L* to elevated [CO~2~]
-We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013.
+We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~).
 We predicted that *L* would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability.
@@
 [-@Norby2003], we found no significant effect of eC~a~ on *L*.
-Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig.
-5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
+Because estimated leaf lifespan (LL) averaged `r round(mean(ba$LL),1)` yrs (Fig.
+5), the evergreen canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
 It could be argued that the change in *L* in response to eC~a~ was too small to be detected.
@@
 To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate.
-Using this method, we estimated that we can detect a difference of only a 6.1% in *L* (see Fig.
+Using this method, we estimated that we can detect a difference as little as 6.1% in *L* (see Fig.
 S5).
@@
 This lack of a response to eC~a~ counters our expectations and those in previous publications [@Norby2003; @McCarthy2007] for a number of reasons.
-First, in a review of forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*.
+First, in a review of all forest FACE sites, Norby and Zak [-@Norby2011] showed that low *L* stands were more responsive to eC~a~ than stands with high *L*.
 They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation.
@@
 [-@Iio2013].
-Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expect about a 30% increase in *L* under eC~a~.
+Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak [-@Norby2011], we expected about a 30% increase in *L* under eC~a~.
 In contrast to other forest FACE sites, the EucFACE is strongly limited by nutrient availability, in particular phosphorus [@Crous2015; @Nielsen2015].
@@
 \
-Despite a lack of response of canopy *L*, we did find a significantly longer LL in eC~a~.
+Despite a lack of response of canopy *L* to eC~a~, we did find a significantly longer LL in eC~a~.
 We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
@@
 4), suggesting the change in LL reflects a small but significant shift in leaf retention, not related to decreased litter production during drought.
-There is a variety of reports from other FACE sites on the relationship between LL and eC~a~.
+Previous results from other forest FACE studies are not conclusive in this regard.
 Norby et al.
@@
 At the same site, Warren et al.
-[-@Warren2011] found increased senescence in the eC~a~ treatment in response to a combined drought and heat wave.
+[-@Warren2011] reported increased senescence in the eC~a~ treatment in response to a combined drought and heat wave.
 McCarthy et al.
-[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* or *Liquidambar* canopies at the DukeFACE, despite enhancement of *L* by eC~a~.
+[-@McCarthy2007] found no effects of eC~a~ on phenology in either *Pinus* sp.
+or *Liquidambar* sp.
+canopies at the DukeFACE, despite enhancement of *L* by eC~a~.
 Taylor et al.
-[-@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE).
+[-@Taylor2008] reported delayed autumn senescence in *Populus* sp.
+in two FACE experiments (PopFACE and AspenFACE).
 Although this indicates the canopy stayed green longer in eC~a~, Gielen et al.
 [-@Gielen2001] reported faster turnover rate of leaves in eC~a~ at the PopFACE, leading to similar *L* in aC~a~ despite higher productivity.
-Li et al.
-[-@Li2000] reported delayed senescence of the oldest leaf cohort in evergreen *Quercus myrtifolia* exposed to eC~a~ in an open-top chamber experiment.
 Clearly, no consistent picture emerges from these comparisons.
@@
 As expected, during periods of decreasing *L*, the change in *L* was correlated with litter production, and we were able to use a four-month period in 2013 without significant rainfall to calibrate estimates of *L* against litter production.
-During periods of increasing *L*, litter fall increased sharply (Figs.
-4 and 5), indicating replacement of old leaves by young leaves.
+During periods of increasing *L*, litter fall increased sharply (Fig.
+3), indicating replacement of old leaves by young leaves.
 Indeed we observed in the field that very few old leaves were present in the canopy after the significant flushing event of December 2013.
@@
 4).
-The consequence is that leaf turnover is highly variable in time.
+Given the high inter-annual variation in rainfall at the EucFACE site, leaf turnover is highly variable in time.
 
 ## Continuous measurement of *L*
-Studying canopy leaf area index in an evergreen natural woodland poses unique challenges.
+Studying canopy leaf area index in an evergreen native woodland poses unique challenges.
 Unlike in deciduous canopies, litter production cannot be used to estimate *L*, because LL is unknown (instead, we used *L* and litter production to estimate LL).
@@
 6).
-We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected, indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
+We found that indirect estimates of *L* based on canopy transmittance were very reliable because relatively small temporal changes in *L* could be detected (see Fig.
+S5), indicating leaf flushing events after rainfall, increased litter production during drought, and an insect attack (Figs.
 1, 4, 6).
@@
 Specht and Brouwer [-@Specht1975] demonstrated for four *Eucalyptus* species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth.
-Similarly, Pook  [-@Pook1984a] demonstrated occurrence of leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods.
-Thus a minimum temperature is also required for leaf growth to occur.
-We also have not observed leaf flushing in winter time during the study period, but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig.
+Similarly, Pook  [-@Pook1984a] demonstrated occurrence of leaf production during most months in *Corymbia maculata*, except during winter and during dry periods.
+Thus there seems to be a minimum temperature threshold for *Eucalyptus* leaf growth.
+Similarly, we also have not observed leaf flushing in winter during the study period, but at our site low temperatures were confounded with water availability as both winters during the study period were relatively dry (Fig.
 6).
@@
 @Kala2013], temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and then allocating some fraction to the leaf pool.
-Combined with - usually constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
+Combined with - usually assumed constant - leaf turnover and specific leaf area, this yields predictions of *L* in response to environmental drivers such as eC~a~.
 We argue that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests.
 In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability, as we have demonstrated that both processes are highly dynamic.
-Water limitation is known to directly influence leaf turnover [@Pook1997], possibly to maintain sufficient rates of water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and rates of leaf expansion [e.g.
+Water limitation is known to directly increase leaf turnover [@Pook1997], possibly to maintain sufficient water availability to the remaining leaves [@Vilagrosa2003; @Misson2011], and decrease rates of leaf expansion [e.g.
 @Muller2011].
@@
 2005), but this process has been poorly quantified.
+Our study demonstrates that both responses of leaf turnover and leaf growth to fluctuating water availability are important to improve predictions of seasonal dynamics of *L* in ecosystem models.
 
 # Acknowledgements {.unnumbered}
-We thank Steven Wohl, Vinod Kumar, Craig Barton and Craig McNamara for maintaining the FACE site and its wide array of sensors, and Andrew Gherlenda, Cassie Mosdal, Tamara Leetham and Catherine Beattie for significant efforts in separating and weighing the litter samples.
+We thank Steven Wohl, Vinod Kumar, Craig Barton and Craig McNamara for maintaining the FACE site and its wide array of sensors, and Andrew Gherlenda, Cassie Mosdal, Tamara Leetham and Catherine Beattie are thanked for significant efforts in separating and weighing the litter samples.
 We also thank Gerry Devine for support in data management.
@@
 (a) Target [CO~2~] (C~a~) relative to the aC~a~ treatment at the EucFACE.
-The eC~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in February 2013.
+The eC~a~ treatment proceeded in steps of 30 ppm, reaching ambient +150 ppm in February 2013.
 (b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions (F~diff~ > 0.98) only.
@@
 Time series of *L* estimated from canopy transmittance.
-Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect.
-Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05.
+Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect of the three replicate rings.
+Note that the 95% confidence always overlap, denoting no significant eC~a~ treatment effect at $\alpha$ = 0.05.
 Variation between rings was substantial, and largely due to pre-treatment differences in stand basal area (see Fig.
-S3 and S5).
+S4).
 
@@
 Error bars are omitted for clarity.
-We did not find any differences in litter production or leaf production between C~a~ treatments (see Results).
+We did not find any differences in P~L~ or leaf production between C~a~ treatments (see Results).
 
@@
 **Figure 5**.
-(a) Total litter production (converted from data in Fig.
-5 to an annual estimate) for each of the FACE rings plotted against mean *L* during the entire litter collection period.
+(a) Total P~L~ (converted from data in Fig.
+4 to an annual estimate) for each of the FACE rings plotted against mean *L* during the entire litter collection period.
 The ratio of mean *L* to annual litter production gives the mean leaf life span.
@@
 
-**Figure S1**. Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately.
+**Figure S1**. Increase in gapfraction ($\tau_d$) and total litter production (P~L~) during the 4-month drought period in 2013, calculated for each ring separately.
 
-**Figure S2**. Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
+**Figure S2**. Correlation between *L* estimated from diffuse canopy transmittance and from canopy photos.
 
-**Figure S3**. Correlation between *L* estimated from diffuse canopy transmittance and from canopy photos.
+**Figure S3**. Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
 
@@
 
-**Figure SI5**. Detectable differences in *L* between C~a~ treatments after accounting for pre-treatment variation in basal area.
+**Figure S5**. Detectable differences in *L* between C~a~ treatments after accounting for pre-treatment variation in basal area.
 

### 2015-08-21 manuscript_supportinginfo.Rmd
Edited according to co-authors comments on revision; suppinfo has own template

@@
   word_document:
-    reference_docx: manuscriptstyle.docx
+    reference_docx: manuscriptsuppinfostyle.docx
   pdf_document: default
@@
 
+For the article "Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability" by Remko A.
+Duursma, Teresa E.
+Gimeno, Matthias M.
+Boer, Kristine Y.
+Crous, Mark G.
+Tjoelker, David S.
+Ellsworth.
+
 ## Supporting Figures
@@
 **Figure S1**.
-Increase in gapfraction ($\tau_d$) and total litter production during the 4-month drought period in 2013, calculated for each ring separately.
+Increase in gapfraction ($\tau_d$) and total litter production (P~L~) during the 4-month drought period in 2013, calculated for each ring separately.
 Blue colours are ambient (nrs 2,3,6), red colours elevated C~a~ (nrs 1,4,5).
@@
 \
+```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=TRUE}
+figureSI2(flatcan_byring)
+```
+
+**Figure S2**.
+Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
+Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
+(a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
+[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
+
+\
 ```{r fig.width=8, fig.height=5, echo=FALSE, message=FALSE, eval=TRUE}
@@
 
-**Figure S2.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
+**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering.
 Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals).
@@
 This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.
-
-\
-```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=TRUE}
-figureSI2(flatcan_byring)
-```
-
-**Figure S3**.
-Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~).
-Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates.
-(a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al.
-[-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).
 

### 2015-08-21 R/figures.R

@@ -338,7 +338,7 @@ figureSI1 <- function(litring, df){
   box()
   mtext(side=2, at=0.75, text=expression(Delta*tau[d]), line=3, outer=TRUE,
         cex=1.2)
-  mtext(side=2, at=0.25, text=expression(Litter~production~(m^2~m^-2)),
+  mtext(side=2, at=0.25, text=expression(P[L]~~(m^2~m^-2)),
         line=3, outer=TRUE, cex=1.2)
   mtext(side=1, at=0.5, text="Ring", outer=TRUE, line=3, cex=1.2)
   

### 2015-08-21 review1.Rmd
edits to response letter

@@
 
-In this manuscript, Duursma et al.
+*Reviewer comment 1* : In this manuscript, Duursma et al.
 present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland.
@@
 
-**We now use pre-treatment BA as a covariate in a mixed-effects model.
+**Our response : We now use pre-treatment BA as a covariate in a mixed-effects model.
 We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate.
-There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect.
+There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to detect a CO2 effect.
 On this point, we have also added a new figure in the supporting information (Fig.
-S5) that addresses the minimum detectable difference with a new method.
+S5) that addresses the minimum detectable difference with a new method, which is 0.05 m2 m-2 or 6%.
 Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**
 
-The only statistically significant response of LAI that would be expected under the current statistical design would be from a CO2 x time interaction.
+*Reviewer comment 2* : The only statistically significant response of LAI that would be expected under the current statistical design would be from a CO2 x time interaction.
 While no statistics are reported, an interaction may have been detected but attributed exclusively to an insect outbreak, whereas it also has the potential to due to a real response.
@@
 
-**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect.
-Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly in the previous manuscript, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect.
-For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect), since it does not confound the overall findings.**
+**Our response : As noted above, we have redone the analysis and find no CO2 or CO2 x time effect over the entire 28-month sampling period (Fig.
+2) that includes the 3 to 4 month period towards the end of the study when insect-mediated leaf loss was observed.
+Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly in the previous manuscript, though not when LAI is considered, see new Fig.2), this comparatively brief period did not result in a significant overall CO2 x time effect.
+For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect), since it does not confound the overall findings.
+In any case, removal of ring 6 (one of the ambient rings) during this period has no effect on CO2 or CO2 x time effects.**
 
-The techniques used to collect litter are also concerning.
+*Reviewer comment 3 and 4* : The techniques used to collect litter are also concerning.
 The authors deployed eight 0.2 m2 litter traps in each ~123 m2 plot and used the litter collected from these traps to estimate litter production.
@@
 
-**The size, number, and relative coverage of the litterfall traps at EucFACE is consistent with the design of other forest FACE studies (see Table below), with in fact the largest trap size compared to other sites.
+**Our response : The size, number, and relative coverage of the litterfall traps at EucFACE is consistent with the design of other forest FACE studies (see Table below), with in fact the largest trap size compared to other sites.
 The size and coverage of the traps is therefore fully justified.
@@
 
-Another issue lies in the inclusion of an add-on analysis comparing two LAI estimation techniques.
+**Additionally, we provide a more detailed description of our custom-made litter traps.**
+
+*Reviewer comment* 5 : Another issue lies in the inclusion of an add-on analysis comparing two LAI estimation techniques.
 While useful, this comparison should be its own methods paper and does not align well with the scope of the rest of the manuscript.
@@
 
-**We have moved this analysis, including the methods, in its entirety to the Supporting Information.
+**Our response : We have moved this analysis, including the methods, in its entirety to the Supporting Information.
 We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only.
@@
 
-Finally, the manuscript requires extensive editing for structure and flow.
+*Reviewer comment* 6 :Finally, the manuscript requires extensive editing for structure and flow.
 
@@
 
-The introduction does not sufficiently establish the expectation of litter responses.
+*Reviewer comment* 7 :The introduction does not sufficiently establish the expectation of litter responses.
 
@@
 
-The methods are incomplete and unclear. Specifically, there are many portions of the Results section that should be included in Materials and Methods.
+*Reviewer comment* 8 :The methods are incomplete and unclear. Specifically, there are many portions of the Results section that should be included in Materials and Methods.
 
@@
 
-Additionally, it is unclear if the authors’ used repeated measures to analyze their data, which would be necessary considering the experimental design.
+*Reviewer comment* 9 :Additionally, it is unclear if the authors’ used repeated measures to analyze their data, which would be necessary considering the experimental design.
 
@@
 
-Beyond these issues, the manuscript also should be revised for clarity and brevity.
+*Reviewer comment* 10 :Beyond these issues, the manuscript also should be revised for clarity and brevity.
 
@@
 
-Specific Comments (line number given):
+**Specific Comments** (line number given):
   Abstract
@@
 
-**Changed to LAI throughout the manuscript.**
+**We keep the abbreviation L because it is more correct to use single letter abbreviations in equations, and it is confusing to use both L (in equations) and LAI (in the text).
+Although LAI is standard, many recent papers use L instead.**
 
@@
 
-**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is hopefully obvious from the statement that the woodland is Eucalyptus-dominated.**
+**We have added another definition of FACE in the Introduction.**
 
@@
 
-**Added '(unmanaged)'.**
+**We have removed this sentence, as it would take too much space to give a detailed account of the site history.**
 
@@
 
-**I HAVE NOT DONE THIS - CAN SOMEONE TELL ME WHAT IS DIFFICULT ABOUT IT???**
+**We have edited this section for clarity.**
 
@@
 
-**We have responded to the data analysis concerns above.**
+**We have responded to the data analysis concerns in comment 2.**
 
@@
 Although we did not set out to examine the outbreak, it occurred nonetheless and we think it is informative as the outbreak reflects natural dynamics of Eucalyptus-dominated woodlands.
-Also, we do not discuss the outbreak in any sort of detail, so removing it from the figure does not shorten the manuscript.**
+Also, we do not discuss the outbreak in any sort of detail, so removing it from the figure does not shorten the manuscript.
+Furthermore, removal of ring 6 does not alter treatment effects.**
 
@@
 
-**We have responded to the concerns about data analysis above.**
+**We have responded to the concerns about data analysis in comment 2**
 
@@
 We disagree that this method 'assumes litter is grown and lost in the same time period'.
-Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shedded leaf area, as measured by litter production.
+Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shed leaf area, as measured by litter production.
 And to  "this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production", this is only true if the reduction in LAI is less than the litter produced.
@@
 
-**The previous analysis was essentially equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis.**
+**The previous analysis was essentially equivalent to using BA as a covariate, however this is now irrelevant since we have redone the analysis using BA as a covariate (see commen 2).**
 
@@
 **It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant to compare the results of this study to the other FACE sites.
-This is informative because the results from the EucFACE do not fit well with a generalization across other FACE sites as presented by Norby and Zak (2011).**
+This is informative because the results from the EucFACE do not fit well with a generalization across other forest FACE sites as presented by Norby and Zak (2011).**
 

### 2015-08-21 manuscript.Rmd
fixed figure references

@@
 The two methods compared quite well (R^2^ = `r f(glance(lmcanphot)$r.squared)`) (Notes S1, Fig.
-S3), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
+S2), suggesting both diffuse transmittance and the photography-based method can be reliably used in *Eucalyptus*-dominated woodlands to estimate *L*.
 
@@
 The random effect accounts for the repeated measures nature of the data.
-For *L*, we further used pre-treatment basal area (BA) of each FACE ring as a covariate (see Fig.
-S4).
+For *L*, we further used pre-treatment basal area (BA) of each FACE ring as a covariate (see Figs.
+S4, S5).
 Because the EucFACE woodland is well established, the differences in BA are not due to C~a~ treatment effects, but instead represent natural variation within the woodland.
@@
 comm.
-Andrew Gherlenda, Markus Riegler].
+M.
+Riegler, A.
+Gherlenda and A.
+Hall].
 This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig.
@@
 During periods of net leaf growth, we found a significant correlation (with a slope of `r f(tidy(lmlitup)$estimate[2])`, SE = `r f(tidy(lmlitup)$std.error[2])`, n = `r glance(lmlitup)$df.residual+2`, p < 0.001) between $\Delta$*L* and litter production (Fig.
-4), indicating that most shedding of old foliage occurred during periods of active leaf growth.
+3), indicating that most shedding of old foliage occurred during periods of active leaf growth.
 
@@
 We found a significant (p `r pval(LLtest$p.value)`) increase in LL with eC~a~ (Fig.
-5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~.
+5b), with LL = `r f(LLtest$estimate1)`yr in aC~a~ and `r f(LLtest$estimate2)`yr in eC~a~ (Fig.
+5b).
 
@@
 Because estimated leaf lifespan (LL) averaged `r round(mean(ba$LL),1)` yrs (Fig.
-5), the evergreen canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
+5b), the evergreen canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur.
 It could be argued that the change in *L* in response to eC~a~ was too small to be detected.
@@
 We found a significant effect of eC~a~ on LL, with leaves living approximately `r round(abs(LLtest$estimate*12),1)` months longer in eC~a~ (Fig.
-5) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
+5b) (an increase of `r round(100*with(LLtest,estimate2/estimate1 - 1),0)`%).
 Because we found no difference in *L* with C~a~ treatment, this suggests a compensating difference in the fractional allocation of net primary production to leaf growth.
@@
 We could not determine whether there was an interaction between LL and water availability, since we estimated leaf turnover from mean *L* and total accumulated leaf litter (Fig.
-5).
+5b).
 Although LL was increased in eC~a~, this was not reflected in a difference in leaf litter production at any time point (Fig.
@@
 6; @Pook1984a], net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing (Fig.
-4).
+3).
 Given the high inter-annual variation in rainfall at the EucFACE site, leaf turnover is highly variable in time.

### 2015-08-21 R/figures.R
Fixed flaw in colors of figure2

@@ -87,7 +87,8 @@ figure2 <- function(df,
   smoothplot(Date, LAI, g=treatment, R="Ring", ylim=ylim, xlim=xlim, 
              ylab=ylab, xlab="",
              data=df, kgam=18, axes=FALSE,
-             pointcolor=rev(my_co2cols()),
+             pointcols=rev(my_co2cols()),
+             linecols=rev(my_co2cols()),
              polycolor=rev(c(alpha("royalblue",0.7),alpha("pink",0.7))))
   
   l <- legend("topleft", c("Ambient","Elevated"), title=expression(italic(C)[a]~treatment), 
